% Chapter3/chapter3.tex -- System Architecture and Design
% -----------------------------------------------------------------------------
% This chapter describes WHAT you built and WHY you designed it that way.
% The Implementation chapter (Chapter 4) then explains HOW you built it in code.

\chapter{System Architecture and Design}
\label{ch:architecture}

\section{Top-Level System Overview}
\label{sec:system_overview}

The inference system is designed around the heterogeneous architecture of the
Xilinx ZCU102, which integrates a quad-core ARM Cortex-A53 processing system
(PS) and a Zynq UltraScale+ programmable logic (PL) fabric on a single SoC.
The two subsystems communicate through a high-bandwidth AXI interconnect and
share a common DDR4 memory pool. This heterogeneous arrangement is well-suited
to transformer inference because the workload naturally decomposes into two
qualitatively different classes of computation: a small amount of sequential
control logic that manages the autoregressive loop, and a large volume of
repetitive arithmetic that is amenable to pipelined hardware acceleration.

The PS is responsible for all control-flow operations. These include loading
the quantised model weights from storage into DDR at startup, running the
tokeniser to convert an input prompt into a sequence of integer token indices,
orchestrating the layer-by-layer forward pass by issuing kernel invocations
over the XRT API, performing post-attention sampling (temperature scaling and
argmax over the vocabulary logit vector), and appending each generated token
back to the context for the next decode step. The PL accelerator kernel
performs the computationally intensive operations within each transformer
layer: the RMSNorm normalisation, the quantise-and-dequantise operations
around each matrix multiplication, the grouped-query attention mechanism
including the scaled dot-product attention and weighted value accumulation,
and the residual stream addition.

The overall execution model for a single decode step proceeds as follows.
The host application supplies the current token index and the current
sequence position to the PL kernel. The kernel loads the token embedding
from DDR, then iterates over all transformer layers, performing the attention
sub-block for each layer and writing the updated key and value vectors into
the KV cache in DDR. After the final layer, the resulting activation vector
is written back to a DDR output buffer. The host then reads this buffer,
applies the final RMSNorm and the language model head projection entirely in
software on the ARM cores, and samples the next token. The process repeats
until the desired output length is reached.

This decomposition was chosen at the project outset with the intention of
incrementally offloading more of the forward pass to the PL fabric as the
hardware design matured. The FFN block, the output attention projection, and
the LM head are designed and specified in this chapter; the scope of the
synthesised hardware kernel and the rationale for the current partitioning
boundary are discussed in \cref{sec:hw_sw_partition}.

\section{Hardware/Software Partitioning}
\label{sec:hw_sw_partition}

The partitioning of operations between the PS and PL is governed by two
criteria applied jointly: arithmetic intensity and implementation complexity.
Arithmetic intensity is defined as the ratio of floating-point operations to
bytes of memory transferred, and it determines whether a computation is
bottlenecked by compute throughput or by memory bandwidth. Operations with
high arithmetic intensity benefit from pipelined hardware acceleration because
the FPGA fabric can sustain many multiply-accumulate operations per clock
cycle once the operands have been loaded. Operations with low arithmetic
intensity are memory-bandwidth-bound and gain little from hardware parallelism
relative to the overhead of transferring data to and from the PL kernel.

The matrix--vector multiplications that dominate each transformer layer have
intermediate arithmetic intensity: each weight element is loaded once and used
once, giving an intensity of approximately two operations per byte at INT8
precision. This is above the threshold where FPGA acceleration begins to
outperform a general-purpose ARM core, particularly given that the ZCU102's
PL fabric supports deeply pipelined DSP48E2 blocks that can sustain one
multiply-accumulate per clock cycle at 200 MHz. The attention score
computation and the weighted value accumulation have similar structure and
are likewise accelerated in the PL.

Tokenisation and sampling are retained on the ARM PS.
Tokenisation is a byte-level string operation with negligible arithmetic
content, and moving it to the PL would introduce significant control-flow
complexity for no performance benefit. Sampling involves an argmax over the
vocabulary logit vector of length 32{,}000 followed by optional temperature
scaling: the memory access pattern is irregular and the arithmetic is trivial,
making it a poor candidate for hardware acceleration. The residual stream
additions, by contrast, are kept inside the PL kernel and executed immediately
after each sub-block's output, avoiding a round-trip of the activation vector
through the AXI interface back to the ARM core on every layer of every decode step.

\Cref{tab:partition} summarises the partitioning decisions for all major
operations in the TinyLlama forward pass.

\begin{table}[H]
  \centering
  \caption{Hardware/software partitioning of the TinyLlama inference pipeline.
           Operations marked PL are executed in the programmable logic
           accelerator kernel; operations marked PS run on the ARM Cortex-A53.}
  \label{tab:partition}
  \begin{tabular}{llll}
    \toprule
    Operation & Dominant cost & Assigned to & Status \\
    \midrule
    Token embedding lookup       & Memory (1 row read) & PS & Implemented \\
    RMSNorm                      & Compute + memory    & PL & Implemented \\
    QKV matrix--vector multiply  & Compute             & PL & Implemented \\
    RoPE positional encoding     & Compute             & PL & Designed \\
    KV cache read/write          & Memory              & PL & Implemented \\
    Scaled dot-product attention & Compute             & PL & Implemented \\
    Output projection ($W_O$)    & Compute             & PL & Designed \\
    Residual addition (attn + FFN) & Memory            & PL & Implemented \\
    FFN RMSNorm                  & Compute + memory    & PL & Designed \\
    SwiGLU FFN                   & Compute             & PL & Designed \\
    Final RMSNorm                & Compute + memory    & PS & Implemented \\
    LM head projection           & Compute             & PS & Implemented \\
    Sampling (argmax/temperature)& Control             & PS & Implemented \\
    \bottomrule
  \end{tabular}
\end{table}

The \emph{Designed} status in \Cref{tab:partition} indicates that the
component was fully specified at the architecture level and the corresponding
HLS kernel code was written and functionally verified in C simulation, but
was excluded from the synthesised bitstream due to resource constraints at the
current simulation scale. The rationale for this boundary is discussed further
in \cref{ch:implementation}.


\section{Memory Architecture}
\label{sec:memory_arch}

The memory architecture is organised around a two-tier hierarchy that reflects
the fundamental trade-off between capacity and access speed available on the
ZCU102. The off-chip DDR4 memory offers large capacity but is shared between
the PS and PL and incurs significant latency for each access. The on-chip
BRAM and flip-flop registers inside the PL fabric are fast and directly
accessible to the compute pipeline, but are scarce relative to the volume of
data involved in transformer inference. The guiding principle behind the
memory organisation is therefore straightforward: data that is too large to
fit on-chip or that must persist across multiple decode steps is placed in
DDR, while data that is small, short-lived, and accessed repeatedly during
a single layer computation is held on-chip.

\noindent\textbf{Off-chip DDR memory.} Three categories of data reside in
DDR: the model weights, the KV cache, and the output buffer. The model weights
are by far the largest component. At full TinyLlama scale, after INT8 quantisation (described in
\cref{sec:quant_scheme}), the attention weight matrices across all 22 layers
total approximately 221\,MB, the FFN weight matrices total approximately
809\,MB, and the token embedding table requires a further 70\,MB. Together
these far exceed the on-chip memory available on any FPGA device, so the
weights must remain in DDR and be streamed into the compute pipeline one
layer at a time. The access pattern
is sequential and read-only during inference, which is well-matched to the
burst transfer capability of the DDR memory controller. A key physical constraint on all DDR buffers shared between the PS and PL is that they must occupy a contiguous region of physical memory, which is enforced through the Linux Contiguous Memory Allocator (CMA) managed transparently by XRT on the ZCU102.

The KV cache must also reside in DDR because its size grows linearly with
sequence length and it must persist across every decode step. Unlike the
weights, the KV cache is both written when a new key or value vector is
computed for the current position, and read when computing attention scores
against all previous positions. At the maximum context length of 2{,}048 tokens, the combined key and
value cache occupies approximately 24.5\,MB for the full TinyLlama model
after INT8 quantisation of the cached vectors, which fits comfortably
within the ZCU102's 4\,GB DDR alongside the weights. A small output buffer, holding a single
activation vector of $d_{\text{model}}$ values, is also allocated in DDR
for the PL kernel to write its result back to the PS at the end of each
decode step.

\noindent\textbf{On-chip memory.} All intermediate activation state
generated during a single layer's computation is held in on-chip BRAM
and registers. This includes the working activation vectors for the current
token, the query, key, and value vectors, the attention score accumulator,
and the intermediate representations produced by the FFN sub-block. The
critical property of all this data is that none of it needs to persist
between layers: once the residual addition has been applied at the end of
a layer, the intermediate values are discarded and the same on-chip buffers
are reused in the next iteration of the layer loop. The on-chip memory
footprint is therefore determined solely by the dimensions of a single layer,
not by the depth of the model. At full TinyLlama scale the on-chip activation
state amounts to tens of kilobytes, well within the ZCU102's available BRAM
capacity. The two-tier memory organisation is summarised in
\cref{tab:memory_hierarchy}.

\begin{table}[H]
  \centering
  \caption{Memory hierarchy of the TinyLlama inference accelerator on the
           ZCU102. Weight and KV cache sizes are after INT8 quantisation.
           On-chip buffers are reused across all 22 layers.}
  \label{tab:memory_hierarchy}
  \begin{tabular}{llll}
    \toprule
    Data & Location & Approx.\ size (full model) & Access pattern \\
    \midrule
    Attention weights ($W_Q, W_K, W_V, W_O$) & DDR     & $\sim$221\,MB  & Streamed per layer \\
    FFN weights ($W_1, W_2, W_3$)             & DDR     & $\sim$809\,MB  & Streamed per layer \\
    Token embedding table                     & DDR     & $\sim$70\,MB   & One lookup per step \\
    KV cache (keys + values)                  & DDR     & $\sim$24.5\,MB & Read/write per step \\
    Output buffer                             & DDR     & $<$1\,MB       & One write per step \\
    Intermediate activation buffers           & On-chip & $\sim$96\,KB   & Reused per layer \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Quantisation Scheme}
\label{sec:quant_scheme}

All weight tensors are quantised offline to INT8 using symmetric per-group
post-training quantisation (PTQ). The decision to use INT8 rather than FP32
was motivated by two considerations. First, INT8 arithmetic is
supported natively by the ZCU102's DSP48E2 blocks as an 8$\times$8 multiply
accumulating into a 27-bit result, which avoids the resource overhead of
implementing floating-point multipliers in LUT fabric. Second, the decode
phase of transformer inference is memory-bandwidth-bound at batch size one,
as established in the GPU profiling of \cref{sec:motivations}: reducing the
weight precision from FP32 to INT8 gives a 4$\times$ reduction in the bytes
transferred per decode step, which directly translates to a 4$\times$
reduction in the DRAM bandwidth required to sustain a given throughput.

For a weight vector $\mathbf{w} \in \mathbb{R}^n$ the symmetric per-group
quantisation scheme divides $\mathbf{w}$ into consecutive non-overlapping
groups of $\text{GS}$ elements and computes a separate scale factor for each
group:

\begin{equation}
  s_g = \frac{\max_{i \in \mathcal{G}_g} |w_i|}{127}
  \label{eq:scale_factor}
\end{equation}

\begin{equation}
  \hat{w}_i = \text{clip}\!\left(\text{round}\!\left(\frac{w_i}{s_g}\right),\,-128,\,127\right)
  \label{eq:quantise}
\end{equation}

\noindent where $\mathcal{G}_g = \{g \cdot \text{GS}, \ldots, (g+1) \cdot
\text{GS} - 1\}$ is the index set of group $g$ and $s_g$ maps the maximum
absolute value in the group to the INT8 boundary of 127. The quantised
weights $\hat{w}_i \in [-128, 127]$ are stored as \texttt{int8\_t} values
and the scale factors $s_g$ are stored alongside as \texttt{float32}, one
per group, in the \texttt{QuantizedTensor} data structure.

The group size $\text{GS}$ controls the granularity of the approximation.
A group size of 1 reduces to per-element quantisation, which minimises
quantisation error at the cost of storing one scale factor per weight. A
group size equal to the full vector length is equivalent to per-tensor
quantisation, which uses minimal storage for scale factors but suffers
higher error when the weight distribution has outliers. The design uses a
group size of $\text{GS} = 64$ for the full TinyLlama model, following the
convention established in the llama2.c reference implementation
\cite{PLACEHOLDER}, which was found to provide an acceptable balance between
quantisation error and the memory overhead of scale factor storage. For the
scaled-down HLS simulation configuration ($d_{\text{model}} = 8$), a group
size of $\text{GS} = 4$ is used so that the constraint $\text{GS} \leq
d_{\text{model}}$ is satisfied and at least one complete quantisation group
is present per weight vector.

At runtime, activations are quantised dynamically before each matrix-vector
multiplication. The quantisation is performed using the same symmetric
per-group scheme, with scale factors computed from the current activation
values rather than precomputed offline. After the integer multiply-accumulate
inside the matrix multiplication, the result is dequantised back to float32
by multiplying each partial sum by the product of the corresponding weight
and activation scale factors. This approach keeps all inner product arithmetic
in the INT8 domain while accumulating into 32-bit integers to prevent
overflow, and defers the floating-point conversion to the final output of
each matrix-vector row. The interaction between the quantisation scheme and
the matrix-vector kernel implementation is described in detail in
\cref{ch:implementation}.
