% Chapter4/chapter4.tex -- Implementation
% -----------------------------------------------------------------------------
% Chapter 4 takes a bottom-up approach: each section covers one computational
% primitive, discussing its implementation and pragma strategy before moving
% to the next. The top-level forward function is covered last, once all
% building blocks have been established.
%
% Section structure:
%   4.1  RMSNorm% Chapter1/chapter1.tex -- Introduction
% -----------------------------------------------------------------------------
% The introduction serves three purposes:
%   1. Motivate the problem (why does it matter?)
%   2. State your objectives (what specifically did YOU do?)
%   3. Orientate the reader (what does the rest of the report contain?)
%
% Keep Section 1.1 broad and accessible -- your examiner may not be an ML expert.
% Keep Section 1.2 specific and measurable -- these become your evaluation criteria.

\chapter{Introduction}
\label{ch:introduction}

\section{Motivations}
\label{sec:motivations}

Large language models (LLMs) have demonstrated excellent capabilities
across a wide range of natural language understanding and generation tasks,
but their deployment in resource-constrained environments remains a
significant challenge \cite{tay2022efficient}. State-of-the-art large language models (LLMs) 
such as GPT-4 and LLaMA require tens or hundreds of gigabytes of memory and specialised
GPU clusters to achieve practical inference throughput, making them
inaccessible for embedded, edge, and real-time applications.

The llama 2-based TinyLlama-1.1B model \cite{zhang2024tinyllama} represents a class of 
small-scale LLMs that seek to balance capability with deployability. With only 1.1 billion
parameters,
TinyLlama retains competitive performance on common benchmarks while fitting
within a much smaller memory footprint. But even at this scale,
running inference on a conventional CPU is far too slow for real-time use,
and GPUs consume too much power for battery-operated or always-on deployments.

Field-programmable gate arrays (FPGAs) offer a compelling alternative.
Unlike ASICs, FPGAs are reconfigurable. The hardware can be
modified precisely to the computational patterns of a specific model, enabling
high-throughput, low-latency inference at a fraction of the power cost of a GPU
\cite{xilinx2021vitishls}. 

There are two inefficiencies of GPU-based inference. 
First, autoregressive decode is memory-bandwidth-bound. Each decode
step must reload the full model weights from DRAM to perform matrix--vector
multiplications against a single token, yet profiling of TinyLlama-1.1B on an
RTX~4060 Laptop GPU reveals low memory bandwidth utilisation as the 
remainder is lost to kernel launch overhead and memory access latency between
hundreds of fine-grained CUDA kernel invocations per step. Second, the GPU
runtime cannot fuse operations across transformer sub-layers. RMSNorm,
matrix--vector multiplication, and activation functions are peformed as
separate kernels, each incurring CPU-to-GPU scheduling latency and cache
eviction between them. 

An FPGA accelerator eliminates both inefficiencies
simultaneously. The weight data flows directly through a statically scheduled
pipeline without repeated DRAM round-trips, and the entire transformer
sub-layer (normalisation, projection, and activation) can be implemented
as a single fused datapath with no inter-kernel overhead. 

% Recent advances
% in high-level synthesis (HLS) tools (Xilinx Vitis HLS)
% have substantially lowered the barrier to FPGA development by allowing
% engineers to describe such fused hardware pipelines in C++ and have the
% tool generate the corresponding RTL automatically.

\section{Objectives and Scope}
\label{sec:objectives}
The general objective of this project is to implement an AI accelerator by
implementing and optimizing the inference pipeline of
TinyLlama-1.1B on the Xilinx ZCU102 evaluation board using Vitis HLS. Due to the 
complexity of the ZCU102 board, we begin our initial kernel design by targeting the 
KRIA KR260 FPGA board. Since the KRIA KR260 FPGA have insufficient resource to 
implement the TinyLlama model, we decided to start with a scaled-down version of 
the model's dimensions.

To understand the bottleneck of AI model inference which will be targeted for acceleration, 
literature review is performed and profiliers are used to obtain performance metrics from 
Nvidia's RTX4060 Laptop GPU. Then, a simple matrix multiplication kernel is 
first implemented with High-Level Synthesis (HLS). By the end of the project, 
a fused kernel consisting of the self-attention and feed-forward layer of the TinyLlama 
archietecture is implemented and synthesized on an FPGA. 

% The primary objectives are as follows.

% \begin{enumerate}
%   \item Implement the core transformer components of TinyLlama -- including
%         RMSNorm, matrix--vector multiplication, multi-head attention (MHA),
%         and the SwiGLU feed-forward network -- as optimised HLS kernels.

%   \item Apply INT8 post-training quantisation (PTQ) to reduce memory bandwidth
%         requirements and exploit the ZCU102's DSP blocks efficiently, while
%         maintaining acceptable output quality.

%   \item Apply HLS optimisation directives (PIPELINE, UNROLL, ARRAY\_PARTITION)
%         to minimise the initiation interval and achieve the target throughput
%         within the ZCU102's resource budget.

%   \item Evaluate the final implementation against a CPU software baseline in
%         terms of inference throughput (tokens per second), latency per token,
%         and energy efficiency (tokens per joule).
% \end{enumerate}

% The scope of this project is limited to autoregressive token generation
% (i.e., the decode phase of inference). Prefill-phase optimisation and
% training are out of scope.

\section{Organisation of Report}
\label{sec:organisation}

The remainder of this report is organised as follows.
\Cref{ch:litreview} reviews the relevant literature on FPGA-based
transformer acceleration, INT8 quantisation techniques, and related HLS
optimisation strategies. \Cref{ch:architecture} presents the overall system
architecture and the design decisions made at the hardware level.
\Cref{ch:implementation} describes the detailed implementation of each HLS
kernel and the software driver layer. \Cref{ch:results} presents the
experimental results including resource utilisation, performance benchmarks,
and comparison against baseline systems. \Cref{ch:conclusion} draws
conclusions and identifies directions for future work.
%   4.2  Quantisation and Dequantisation
%   4.3  Matrix--Vector Multiplication
%   4.4  Softmax
%   4.5  Grouped-Query Attention
%   4.6  SwiGLU Feed-Forward Network
%   4.7  Top-Level Forward Function

\chapter{Implementation}
\label{ch:implementation}

All kernels in this chapter are implemented in Vitis HLS 2021.2,
targeting the \texttt{xczu9eg-ffvb1156-2-e} device on the ZCU102 evaluation
board at a design clock of \SI{150}{\mega\hertz}. Each primitive is first
presented as a standalone synthesisable function, allowing its resource
utilisation and initiation interval to be measured in isolation before it
is composed into the full transformer forward pass.

Development proceeded across two configurations. A scaled-down prototype
configuration was used to establish the kernel structure, verify the HLS
build flow, and confirm functional correctness in C-simulation without the
lengthy synthesis times and large on-chip resource requirements of the full
model. Once each component was validated at small scale, the configuration
constants were updated to target the full TinyLlama-1.1B dimensions.
Table~\ref{tab:config_comparison} summarises the two sets of parameters.

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Parameter & Prototype & TinyLlama-1.1B \\
    \midrule
    Hidden dimension $d_{\text{model}}$ & 8       & 2{,}048  \\
    FFN hidden dimension $d_{\text{ff}}$ & 32     & 5{,}632  \\
    Number of layers $L$                 & 2       & 22       \\
    Query heads $H$                      & 2       & 32       \\
    KV heads $G$                         & 2       & 4        \\
    Vocabulary size $V$                  & 32      & 32{,}000 \\
    Maximum sequence length              & 16      & 2{,}048  \\
    Quantisation group size $GS$         & 4       & 64       \\
    \bottomrule
  \end{tabular}
  \caption{Prototype and full TinyLlama-1.1B configuration parameters
           used across kernel development.}
  \label{tab:config_comparison}

\end{table}

The prototype dimensions were chosen deliberately to keep array sizes small
enough that HLS C-simulation completes in seconds and synthesis reports are
straightforward to inspect. With $d_{\text{model}} = 8$ and $GS = 4$, for
instance, every quantised vector contains two groups, making
scale-factor indexing trivial to verify by hand. The constraint that $GS
\leq d_{\text{model}}$ must always hold, since quantising into groups larger
than the vector itself is undefined. At full scale the group size grows to
$GS = 64$, which divides evenly into all relevant dimensions in TinyLlama
(2048, 5632, 256) and aligns with a typical cache line size of 64 bytes,
making group-boundary loads efficient.

% =============================================================================
\section{RMSNorm}
\label{sec:rmsnorm_impl}
% =============================================================================

\subsection{Implementation}
\label{subsec:rmsnorm_impl}

The \texttt{rmsnorm<S>} function is declared as a C++ template in
\texttt{forward.h}, parameterised by the vector length \texttt{S}. This
makes the same source compile correctly for both the prototype configuration
($S = d_{\text{model}} = 8$) and the full TinyLlama model
($S = d_{\text{model}} = 2048$) without any code changes. The function
takes three pointer arguments, namely an output array \texttt{o}, an input
activation array \texttt{x}, and a learned element-wise weight array
\texttt{weight}, all of length \texttt{S}.

The first design decision is to copy both \texttt{x} and \texttt{weight}
into local arrays \texttt{x\_buff} and \texttt{weight\_buff} before any
computation begins, as shown in Listing~\ref{lst:rmsnorm}. This is
necessary because Vitis HLS can only apply \texttt{ARRAY\_PARTITION} to
arrays whose storage it controls. Arrays received as pointer arguments may
alias external memory, which prevents the tool from freely splitting them
across multiple BRAM banks. By staging the data into local buffers first,
the \texttt{ARRAY\_PARTITION cyclic factor=4} pragma can be applied,
enabling four simultaneous reads per cycle in both computation loops.

\begin{lstlisting}[
  caption = {\texttt{rmsnorm<S>} implementation from \texttt{forward.h}.
             Data is staged into local buffers before computation so that
             \texttt{ARRAY\_PARTITION} can be applied. The reduction loop
             uses \texttt{UNROLL factor=4} to accumulate four partial sums
             simultaneously. The normalisation loop uses \texttt{UNROLL
             factor=64} since it is fully data-parallel.},
  label   = {lst:rmsnorm}
]
template <int S>
void rmsnorm(float o[S], float x[S], float weight[S])
{
    float x_buff[S], weight_buff[S], out_buff[S];
    #pragma HLS array_partition variable=x_buff      cyclic factor=4
    #pragma HLS array_partition variable=weight_buff cyclic factor=4
    #pragma HLS array_partition variable=out_buff    cyclic factor=4

    memcpy(x_buff,      x,      S * sizeof(float));
    memcpy(weight_buff, weight, S * sizeof(float));

    float ss = 0.0f;
sum_of_squares:
    for (int j = 0; j < S; j++) {
        #pragma HLS PIPELINE
        #pragma HLS UNROLL factor=4 skip_exit_check
        ss += x_buff[j] * x_buff[j];
    }
    ss /= S;
    ss += 1e-5f;
    ss = 1.0f / sqrtf(ss);   // reciprocal RMS

norm_and_scale:
    for (int j = 0; j < S; j++) {
        #pragma HLS PIPELINE
        #pragma HLS UNROLL factor=64
        out_buff[j] = weight_buff[j] * (ss * x_buff[j]);
    }
    memcpy(o, out_buff, S * sizeof(float));
}
\end{lstlisting}

The computation proceeds in two sequential loops with different parallelism
requirements. The first loop, \texttt{sum\_of\_squares}, accumulates the
sum of squared activations. This is a reduction with a loop-carried
dependency on the accumulator \texttt{ss}, because each iteration reads the
current value of \texttt{ss}, adds to it, and writes it back. This
dependency means the loop cannot be pipelined to II\,=\,1 on its own
because the write of iteration $t$ must complete before iteration $t+1$
can read. The \texttt{UNROLL factor=4} directive resolves this by
unrolling four iterations into the same pipeline stage, so that four independent
partial products $x_j^2, x_{j+1}^2, x_{j+2}^2, x_{j+3}^2$ are computed
in parallel and their sum is added to \texttt{ss} once per group, reducing
the effective number of dependent additions by a factor of four.

Once \texttt{ss} has been computed, the scalar reciprocal RMS is available
and the second loop, \texttt{norm\_and\_scale}, has no loop-carried
dependencies. Every iteration reads one element of \texttt{x\_buff} and
one element of \texttt{weight\_buff}, multiplies by the scalar \texttt{ss},
and writes to \texttt{out\_buff}. These are entirely independent operations,
so the loop is amenable to wide unrolling. The \texttt{UNROLL factor=64}
directive allows up to 64 iterations to execute in parallel, subject to
memory port availability, which is satisfied by the \texttt{cyclic
factor=4} partitioning applied to all three local buffers.

\subsection{C-Simulation Results}
\label{subsec:rmsnorm_csim}

C-simulation was performed for \texttt{rmsnorm\_top}. The testbench
supplied a fixed eight-element input vector and a corresponding weight
vector, then compared the kernel output element by element against a
pure software reference implementation of RMSNorm. The reference
computes the sum of squares sequentially, derives the reciprocal RMS,
and applies the element-wise scaling in the same order as the
mathematical definition in \cref{subsec:rmsnorm}, with no HLS pragmas
or unrolling. Table~\ref{tab:rmsnorm_csim} shows the output from both the kernel and
the software reference for each element of the test vector.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & HW output & SW reference & Absolute error \\
    \midrule
    0 &  0.099015 &  0.099015 & 0 \\
    1 & -0.396059 & -0.396059 & 0 \\
    2 &  0.891133 &  0.891133 & 0 \\
    3 & -1.584236 & -1.584236 & 0 \\
    4 &  0.495074 &  0.495074 & 0 \\
    5 & -1.188177 & -1.188177 & 0 \\
    6 &  2.079309 &  2.079309 & 0 \\
    7 & -3.168472 & -3.168472 & 0 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for \texttt{rmsnorm} with input
           $\mathbf{x} = [1, -2, 3, -4, 5, -6, 7, -8]$ and weight
           $\mathbf{g} = [0.5, 1.0, 1.5, 2.0, 0.5, 1.0, 1.5, 2.0]$.
           All eight elements match the software reference.}
  \label{tab:rmsnorm_csim}

\end{table}

The absolute error is identically zero for every element, meaning the
kernel and the reference produce bit-identical results on this test
vector. This outcome is stricter than expected, since the
kernel's \texttt{sum\_of\_squares} loop uses \texttt{UNROLL factor=4},
HLS may group elements into sets of four and accumulate their squared
values in a different order from the sequential reference. Floating-point
addition is not associative, so reordering additions can produce
different rounding outcomes. This is because each floating-point addition
rounds its result to the nearest representable value, and that rounding
error accumulates differently depending on the order in which operands
are grouped. In this particular case the zero-error result
reflects the small scale of the prototype dimensions ($S = 8$), where
the accumulated rounding differences are below the representational limit
of single-precision arithmetic. At the full TinyLlama scale ($S = 2048$),
small discrepancies between the unrolled and sequential summation orders
are expected, and a tolerance threshold rather than exact equality would
be the appropriate acceptance criterion. The csim run completed with zero
errors as reported by Vitis HLS.

\subsection{HLS Synthesis Results}
\label{subsec:rmsnorm_csyn}

The \texttt{rmsnorm} function was synthesised in isolation using the
prototype configuration ($S = 8$) on the \texttt{xczu9eg-ffvb1156-2-e}
device at a \SI{150}{\mega\hertz} target clock (\SI{6.667}{\nano\second}
period). Tables~\ref{tab:rmsnorm_timing} and \ref{tab:rmsnorm_top_latency}
reproduce the top-level timing and latency summary as reported by Vitis
HLS, followed by Table~\ref{tab:rmsnorm_resources} for resource utilisation
and Table~\ref{tab:rmsnorm_latency} for the per-stage latency breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock   & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667       & 4.867          & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{rmsnorm\_top} at the
           \SI{150}{\mega\hertz} (\SI{6.667}{\nano\second}) target clock.}
  \label{tab:rmsnorm_timing}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    146 & 146 & \SI{0.973}{\micro\second} & \SI{0.973}{\micro\second} & 147 & 147 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{rmsnorm\_top} at
           \SI{150}{\mega\hertz}.}
  \label{tab:rmsnorm_top_latency}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT      & 4{,}646 & 3{,}575 & 274{,}080 & 1 \\
    FF       & 3{,}910 & 2{,}949 & 548{,}160 & $<$1 \\
    BRAM     & 4       & 0       & 1{,}824   & $<$1 \\
    DSP      & 26      & 26      & 2{,}520   & 1 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of the \texttt{rmsnorm} function
           synthesised at $S = 8$ on the ZCU102
           (\texttt{xczu9eg-ffvb1156-2-e}). Figures in the core-only
           column exclude the AXI master adapter overhead, which is
           not present when the function is inlined into the full kernel.}
  \label{tab:rmsnorm_resources}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Stage & Trip count & II & Iteration latency & Total (cycles) \\
    \midrule
    Load \texttt{x} into \texttt{x\_buff}           & 8 &  1 &  3 &  9 \\
    \texttt{sum\_of\_squares}                        & 2 & 24 & 29 & 52 \\
    Load \texttt{weight} into \texttt{weight\_buff}  & 8 &  1 &  3 & 11 \\
    \texttt{norm\_and\_scale}                        & 8 &  1 &  2 & 10 \\
    \bottomrule
  \end{tabular}
  \caption{Latency breakdown of \texttt{rmsnorm} ($S = 8$) by internal
           pipeline stage at \SI{150}{\mega\hertz}. Total function latency
           is 146 cycles (\SI{973}{\nano\second}).}
  \label{tab:rmsnorm_latency}

\end{table}

The total function latency of 146 cycles is dominated by the
\texttt{sum\_of\_squares} reduction loop, which alone accounts for 52
of those cycles despite iterating only twice after the
\texttt{UNROLL factor=4} collapses the 8-element loop into 2 effective
iterations. The cause is the loop-carried dependency on the
floating-point accumulator \texttt{ss}. The Bind Op report shows that
HLS mapped the floating-point addition to a full-DSP \texttt{fadd}
\texttt{faddfsub\_32ns\_32ns\_32\_6\_full\_dsp\_1} unit
with a pipeline latency of 5 cycles at this clock frequency. Because each
iteration must read the current value of \texttt{ss} and write back an
updated value, the next iteration cannot begin until the adder has
produced its result, giving an effective II of 24 cycles. The
\texttt{UNROLL factor=4} partially mitigates this by reducing the trip
count from 8 to 2, cutting the number of stall events in half, but it
cannot eliminate the dependency itself. A more complete solution would be
accumulator interleaving, where four independent partial sums are
maintained and merged only at the end, allowing all four lanes to pipeline
at II\,=\,1 independently. This is identified as a future optimisation
opportunity.

In contrast, the \texttt{norm\_and\_scale} loop achieves II\,=\,1
perfectly. Once the scalar reciprocal RMS \texttt{ss} is available, every
element's computation is independent of all others, allowing HLS to start
a new element every clock cycle. The \texttt{UNROLL factor=64} on an
8-element loop effectively fully unrolls it into 8 parallel datapath
lanes, which is reflected in the 26 DSPs consumed. The 8 floating-point
multipliers for the \texttt{weight\,$\times$\,(ss\,$\times$\,x)}
computation each map to 3 DSPs, accounting for 24 of the 26 total, with
the remaining 2 used by the \texttt{fadd} in the reduction.

The 4 BRAMs in the total utilisation are consumed entirely by the AXI
master adapter (\texttt{gmem0\_m\_axi}) of the standalone wrapper and
are not a property of the \texttt{rmsnorm} logic itself. The
\texttt{rmsnorm} core uses no BRAM, since the local buffers
\texttt{x\_buff}, \texttt{weight\_buff}, and \texttt{out\_buff} at
$S = 8$ are small enough to be inferred as register files. The estimated
clock period of \SI{4.867}{\nano\second} leaves \SI{1.800}{\nano\second}
of uncertainty budget, which meets the \SI{6.667}{\nano\second}
target with zero slack remaining. This is reflected in the \texttt{Timing}
issue flag visible in the synthesis summary report. While the design meets
the timing constraint, the zero-slack condition means there is no margin
for additional logic to be added along the critical path at this clock
frequency without risking a timing violation.

The \texttt{rmsnorm} implementation was not modified in the second or
third kernel iterations. The synthesis results presented here are
therefore valid across all three iterations of the prototype.

% =============================================================================
\subsection{Full TinyLlama Scale}
\label{subsec:rmsnorm_update3}
% =============================================================================

\subsubsection{Implementation}
\label{subsubsec:rmsnorm_update3_impl}

The sixth kernel iteration scales \texttt{rmsnorm} to the full TinyLlama
hidden dimension ($S = d_{\text{model}} = 2048$). The source template is
unchanged in its algorithmic structure but carries three targeted pragma
modifications, as shown in Listing~\ref{lst:rmsnorm_update3}.

\begin{lstlisting}[
  caption = {\texttt{rmsnorm<S>} implementation from \texttt{FFN\_update3.h}
             at full TinyLlama scale ($S = 2048$). The \texttt{ARRAY\_PARTITION}
             factor increases from 4 to 16, and both loop \texttt{UNROLL}
             factors change to 16 to match \texttt{PARALLEL\_MACS}.},
  label   = {lst:rmsnorm_update3}
]
template <int S>
void rmsnorm(float o[S], float x[S], float weight[S]) {
    constexpr auto array_size = S * sizeof(float);

    float ss = 0.0f;
    float x_buff[S];
    float weight_buff[S];
    float out_buff[S];
    #pragma HLS ARRAY_PARTITION variable=x_buff      type=cyclic factor=16
    #pragma HLS ARRAY_PARTITION variable=weight_buff type=cyclic factor=16
    #pragma HLS ARRAY_PARTITION variable=out_buff    type=cyclic factor=16

    memcpy(x_buff,      x,      array_size);
    memcpy(weight_buff, weight, array_size);

sum_of_squares:
    for (int j = 0; j < S; j++) {
        #pragma HLS PIPELINE II=1
        #pragma HLS UNROLL factor=16
        ss += x_buff[j] * x_buff[j];
    }
    ss /= S;
    ss += 1e-5f;
    ss = 1.0f / sqrtf(ss);

norm_and_scale:
    for (int j = 0; j < S; j++) {
        #pragma HLS PIPELINE II=1
        #pragma HLS UNROLL factor=16
        out_buff[j] = weight_buff[j] * (ss * x_buff[j]);
    }
    memcpy(o, out_buff, array_size);
}
\end{lstlisting}

The first change is to the \texttt{ARRAY\_PARTITION cyclic factor}, which
increases from 4 to 16 on all three local buffers. At $S = 8$ in the
prototype, four banks sufficed because the loop bodies accessed at most
four elements simultaneously. At $S = 2048$, 16 simultaneous accesses are
needed to keep all unrolled lanes supplied without bank contention,
and factor\,=\,16 is therefore the minimum that avoids port conflicts. Each
2048-element float array partitioned with factor\,=\,16 produces
$2048 / 16 = 128$ words per bank, which falls above the BRAM inference
threshold and causes HLS to map each bank to a BRAM rather than a register
file, accounting for the large on-chip memory cost visible in the synthesis
results.

The second change affects the \texttt{sum\_of\_squares} loop, where the
\texttt{UNROLL factor} increases from 4 to 16. In the prototype this loop
iterated over $S = 8$ elements, so an unroll factor of 4 reduced the trip
count to 2. At $S = 2048$, a factor of 16 reduces the trip count to 128,
creating 16 independent partial accumulator lanes that each carry their own
loop-carried dependency chain. HLS schedules all 16 lanes in parallel,
achieving an effective throughput of 16 elements per iteration despite the
5-cycle floating-point addition latency on each lane. The pipeline II
is explicitly constrained to 1 with the added \texttt{II=1} annotation.

The third change affects the \texttt{norm\_and\_scale} loop, where the
\texttt{UNROLL factor} decreases from 64 to 16. In the prototype with
$S = 8$, an unroll factor of 64 effectively fully unrolled the loop into
8 parallel lanes at negligible resource cost. At $S = 2048$, maintaining
an unroll factor of 64 would instantiate 64 parallel \texttt{fmul} units,
each consuming 3 DSPs, for a total of 192 DSPs from this loop alone. Since
the \texttt{norm\_and\_scale} computation has no loop-carried dependencies,
a factor of 16 already achieves II\,=\,1 with full utilisation of the 16-bank
partitioned buffers, making the higher unroll factor unnecessary. The
reduction from 64 to 16 therefore preserves correctness and throughput
while significantly reducing the DSP count.

\subsubsection{C-Simulation Results}
\label{subsubsec:rmsnorm_update3_csim}

The testbench supplied a 2048-element input vector with incrementally
increasing values and unit RMSNorm weights. All 2048 output elements
agreed with the sequential software reference with zero absolute error,
giving an RMS error of $0$ across the full vector. The csim run completed
with zero errors as reported by Vitis HLS.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:rmsnorm_update3_csyn}

Tables~\ref{tab:rmsnorm_u3_timing} and \ref{tab:rmsnorm_u3_top_latency}
present the timing and latency summary. Table~\ref{tab:rmsnorm_u3_resources}
gives the resource utilisation and Table~\ref{tab:rmsnorm_u3_latency} the
per-stage breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 7.456 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{rmsnorm\_update3\_top} at the
           \SI{150}{\mega\hertz} target clock. The estimated path of
           \SI{7.456}{\nano\second} plus \SI{1.800}{\nano\second}
           uncertainty exceeds the \SI{6.667}{\nano\second} budget by
           \SI{2.589}{\nano\second}, constituting a timing violation.}
  \label{tab:rmsnorm_u3_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    18{,}645 & 18{,}645 & \SI{0.139}{\milli\second} & \SI{0.139}{\milli\second} & 18{,}646 & 18{,}646 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{rmsnorm\_update3\_top}
           at \SI{150}{\mega\hertz}.}
  \label{tab:rmsnorm_u3_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 11{,}977 &  9{,}954 & 274{,}080 &  4 \\
    FF   &  9{,}917 &  8{,}458 & 548{,}160 &  1 \\
    BRAM &      100 &       96 &   1{,}824 &  5 \\
    DSP  &       98 &       98 &   2{,}520 &  3 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{rmsnorm} at full scale
           ($S = 2048$) on the ZCU102. Core-only figures exclude the
           AXI master adapter (\texttt{gmem0\_m\_axi},
           4\,BRAM + 1{,}459\,FF + 2{,}023\,LUT).}
  \label{tab:rmsnorm_u3_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Stage & Module & Latency (cycles) \\
    \midrule
    Load \texttt{x\_in}, \texttt{weight} & \texttt{VITIS\_LOOP\_29\_1} &  4{,}106 \\
    \texttt{rmsnorm\_2048\_s} core       & \texttt{rmsnorm\_2048\_s}   & 12{,}478 \\
    Write output                         & \texttt{VITIS\_LOOP\_37\_2} &  2{,}051 \\
    \bottomrule
  \end{tabular}
  \caption{Per-stage latency breakdown of \texttt{rmsnorm} at full scale
           ($S = 2048$) at \SI{150}{\mega\hertz}. Total function latency
           is 18{,}645 cycles (\SI{0.124}{\milli\second}).}
  \label{tab:rmsnorm_u3_latency}
\end{table}

The synthesis reports a timing violation at \SI{150}{\mega\hertz}. The
prototype at $S = 8$ estimated \SI{4.867}{\nano\second} and met the
\SI{6.667}{\nano\second} target. At $S = 2048$ the estimated path is
\SI{7.456}{\nano\second}, exceeding the budget by \SI{2.589}{\nano\second}.
The cause is the binary adder tree in the \texttt{sum\_of\_squares} loop.
With \texttt{UNROLL factor=16}, HLS instantiates 16 floating-point
multipliers whose products are summed through a tree of adders. At $S = 8$
with \texttt{UNROLL factor=4}, the tree has two levels. At $S = 2048$
with \texttt{UNROLL factor=16}, it has four levels, and each level
contributes one floating-point addition delay to the combinational path.
This depth prevents timing closure at \SI{150}{\mega\hertz}. Reducing the
unroll factor to 8, or maintaining four independent partial accumulators
that are merged only after the loop exits, would shorten the tree and
resolve the violation.

The total latency of 18{,}645 cycles (\SI{0.124}{\milli\second}) breaks
into three stages. The load pipeline transfers 4{,}096 floats (both
\texttt{x} and \texttt{weight}) from DDR in 4{,}106 cycles. The
\texttt{rmsnorm\_2048\_s} core runs for 12{,}478 cycles, accounting for
67\% of the total. The write pipeline returns 2{,}048 output floats in
2{,}051 cycles. The AXI transfer overhead is fixed at $O(S)$ and forms
a larger fraction of the total at smaller $S$ values, where the core
compute is shorter.

The BRAM count increases from 0 core BRAMs at $S = 8$ to 96 core BRAMs
at $S = 2048$. At prototype scale, all six local buffer arrays
(\texttt{x\_buff}, \texttt{weight\_buff}, and \texttt{out\_buff} in both
the wrapper and the inlined core) fit in register files. At $S = 2048$
with factor\,=\,16 partitioning, each array is split into 16 banks of
128 words each. A 128-word bank exceeds the BRAM inference threshold, so
HLS maps each bank to a single-port BRAM, giving $6 \times 16 = 96$ core
BRAMs in total.

The DSP count increases from 26 at prototype scale to 98 at full scale.
The 98 DSPs come from three sources: 16 \texttt{fmul} units in
\texttt{norm\_and\_scale} (48 DSPs), 16 \texttt{fmul} units in
\texttt{sum\_of\_squares} (48 DSPs), and the \texttt{fadd} accumulator
(2 DSPs). Had the \texttt{norm\_and\_scale} unroll factor remained at 64
as in the prototype, HLS would have instantiated 64 \texttt{fmul} units
in that loop alone, consuming 192 DSPs from a single loop. Reducing the
factor to 16 brings this contribution down to 48 DSPs while preserving
II\,=\,1, since the loop has no loop-carried dependencies and 16 banks
are sufficient to supply 16 parallel lanes without port contention.

The LUT count of 9{,}954 core LUTs is higher than the 3{,}575 of the
prototype. The increase comes from three sources. The first is the wider multiplexer
network arbitrating 16-bank memory accesses. The second is the larger FSM
for the expanded loop bounds at $S = 2048$. The third is the combinational
logic of the
four-level adder tree. The FF count grows for the same reasons, as each
pipeline register stage in the adder tree adds flip-flop storage
proportional to the data width.



% =============================================================================
\section{Quantisation}
\label{sec:quantize_impl}
% =============================================================================

\subsection{Implementation}
\label{subsec:quantize_impl}

The \texttt{quantize<S>} function implements the group-wise symmetric
INT8 quantisation scheme described in \cref{sec:quantisation}. It is
declared as a C++ template in \texttt{forward.h}, parameterised by the
vector length \texttt{S}, and outputs results through the
\texttt{QuantizedTensor<S>} struct. A \texttt{QuantizedTensor<S>}
bundles an \texttt{int8\_t} array \texttt{q[S]} for the quantised values
and a \texttt{float} array \texttt{s[S]} for the scale factors, where
only \texttt{S/GS} scale entries are populated, one per group of
\texttt{GS} consecutive elements.

The \texttt{quantize<S>} function
processes the input float vector in groups of \texttt{GS} elements, as
shown in Listing~\ref{lst:quantize}. For each group, the function first
scans the \texttt{GS} elements to find the maximum absolute value
\texttt{wmax}, then computes the group scale as $s = \texttt{wmax} /
127.0$, which maps the largest magnitude in the group to the maximum
representable INT8 value of 127. Each element is then divided by the
scale and rounded to the nearest integer, producing a value in
$[-127, 127]$ that is stored as \texttt{int8\_t}. The quantised values
and scales are accumulated into local buffers
\texttt{quantized\_buffer[S]} and \texttt{scale\_buffer[num\_groups]}
before being written to the output struct via \texttt{memcpy}, avoiding
repeated writes to the output pointer during the inner loops.

\begin{lstlisting}[
  caption = {Group-wise quantisation in \texttt{quantize<S>} from
             \texttt{forward.h}. The outer loop iterates over groups;
             the inner scan finds \texttt{wmax} and the second inner loop
             converts each element. The compile-time constant \texttt{GS}
             replaces the runtime parameter \texttt{runtime\_gs} so the
             compiler can evaluate \texttt{S/GS} as a constant expression.},
  label   = {lst:quantize}
]
for (int group = 0; group < num_groups; group++) {
    #pragma HLS UNROLL factor=1
    #pragma HLS PIPELINE
    float wmax = 0.0f;
    int base_idx = group * GS;

    for (int i = 0; i < GS; i++) {          // find max abs value in group
        #pragma HLS PIPELINE
        float val = fabs(x[base_idx + i]);
        if (val > wmax) wmax = val;
    }

    float scale = wmax / 127.0f;             // symmetric INT8 scale
    scale_buffer[group] = scale;

    for (int i = 0; i < GS; i++) {          // quantise each element
        #pragma HLS PIPELINE
        quantized_buffer[base_idx + i] =
            (int8_t)round(x[base_idx + i] / scale);
    }
}
\end{lstlisting}

A notable implementation detail is the use of \texttt{GS} as a
compile-time constant from \texttt{config.h} rather than the
\texttt{runtime\_gs} parameter passed to the function.
The parameter exists to prevent the C++ compiler
from issuing a warning about a shadowed variable. An earlier iteration of
the code attempted to use a local variable also named \texttt{GS}, which
conflicted with the global constant. The rename to \texttt{runtime\_gs}
resolves the shadowing, and the function body then uses the global
\texttt{GS} constant directly. Because \texttt{GS} is a compile-time
constant, the compiler can statically evaluate \texttt{S/GS} as
\texttt{num\_groups}, enabling the \texttt{ARRAY\_PARTITION} pragma on
\texttt{quantized\_buffer} and allowing the inner loops to be analysed
at synthesis time without variable-bound ambiguity.

The \texttt{PIPELINE} pragma on the outer group loop combined with
\texttt{UNROLL factor=1} is an intentional configuration. Applying
\texttt{PIPELINE} to the outer loop while leaving the inner scan and
quantisation loops unpipelined would normally cause HLS to flatten the
inner loops into the outer pipeline, which at \texttt{GS=4} means four
sequential comparisons per outer iteration. The \texttt{UNROLL factor=1}
prevents the outer loop from being unrolled further.

\subsection{C-Simulation Results}
\label{subsec:quantize_csim}

C-simulation was performed for \texttt{quantize\_top}. The testbench supplied
an eight-element float vector containing a mix of positive and negative
values with varying magnitude and compared the kernel output against a
software reference implementation of the same group-wise symmetric INT8
quantisation scheme. Two aspects were verified independently, namely the INT8
quantised values and the per-group scale factors. Table~\ref{tab:quantize_csim} shows the INT8 output and scale factors
produced by the kernel and the software reference for each element and
group respectively.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Index & Input ($x$) & HW \texttt{q} & SW \texttt{q} & Abs error \\
    \midrule
    0 &  0.5 &   17 &   17 & 0 \\
    1 & -1.2 &  -41 &  -41 & 0 \\
    2 &  3.7 &  127 &  127 & 0 \\
    3 & -0.3 &  -10 &  -10 & 0 \\
    4 &  2.1 &   59 &   59 & 0 \\
    5 & -4.5 & -127 & -127 & 0 \\
    6 &  1.0 &   28 &   28 & 0 \\
    7 & -2.8 &  -79 &  -79 & 0 \\
    \midrule
    Group & HW scale & SW scale & Abs error & \\
    \midrule
    0 & 0.029134 & 0.029134 & 0 & \\
    1 & 0.035433 & 0.035433 & 0 & \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for \texttt{quantize} with input
           $\mathbf{x} = [0.5, -1.2, 3.7, -0.3, 2.1, -4.5, 1.0, -2.8]$.
           All INT8 values and both scale factors match the software
           reference. GS\,=\,4 so group~0 covers indices 0--3
           and group~1 covers indices 4--7.}
  \label{tab:quantize_csim}

\end{table}

All eight INT8 values and both scale factors agree between the
hardware kernel and the software reference. Elements at index~2 and
index~5 map to the boundary values of $127$ and $-127$ respectively,
confirming that the maximum-magnitude element in each group correctly
saturates the INT8 range. The scale for group~0 is $3.7 / 127 \approx
0.02913$, and for group~1 is $4.5 / 127 \approx 0.03543$, both of which
match the reference to the displayed precision. The csim run completed
with zero errors.

\subsection{HLS Synthesis Results}
\label{subsec:quantize_csyn}

The \texttt{quantize} function was synthesised in isolation at
\SI{150}{\mega\hertz} on the \texttt{xczu9eg-ffvb1156-2-e} device.
Tables~\ref{tab:quantize_timing} and \ref{tab:quantize_top_latency}
present the top-level timing and latency summary, followed by
Table~\ref{tab:quantize_resources} for resource utilisation and
Table~\ref{tab:quantize_latency} for the per-stage breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{quantize\_top} at
           \SI{150}{\mega\hertz}.}
  \label{tab:quantize_timing}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    84 & 84 & \SI{0.560}{\micro\second} & \SI{0.560}{\micro\second} & 85 & 85 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{quantize\_top} at
           \SI{150}{\mega\hertz}.}
  \label{tab:quantize_top_latency}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 6{,}379 & 3{,}275 & 274{,}080 & 2 \\
    FF   & 3{,}556 & 1{,}653 & 548{,}160 & $<$1 \\
    BRAM & 4       & 0       & 1{,}824   & $<$1 \\
    DSP  & 0       & 0       & 2{,}520   & 0 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{quantize} synthesised at
           $S = 8$ on the ZCU102. The core-only column excludes the
           AXI master adapter.}
  \label{tab:quantize_resources}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Stage & Trip count & II & Iteration latency & Total (cycles) \\
    \midrule
    \texttt{main\_loop} (quantise core) & 2 & 4  & 41 & 44 \\
    Pipeline\_2 (write \texttt{q\_out}) & 8 & 1  &  1 &  8 \\
    Pipeline\_3 (write \texttt{s\_out}) & 2 & 1  &  1 &  2 \\
    \bottomrule
  \end{tabular}
  \caption{Latency breakdown of \texttt{quantize} ($S = 8$, $GS = 4$)
           by internal pipeline stage at \SI{150}{\mega\hertz}.
           Total function latency is 84 cycles
           (\SI{560}{\nano\second}).}
  \label{tab:quantize_latency}

\end{table}

The zero DSP utilisation is the most notable result. In contrast to
RMSNorm, which consumed 26 DSPs for its floating-point multiplications,
\texttt{quantize} uses none. The Bind Op report confirms that both
division operations, namely computing the group scale as $s =
\texttt{wmax}/127$ and quantising each element as $q_i = x_i / s$, are
mapped to fabric-based \texttt{fdiv} units with the \texttt{no\_dsp}
suffix. This occurs because Xilinx DSP48 blocks natively support only
multiply-accumulate operations. Floating-point division has no direct DSP
mapping and must instead be implemented using LUT-based logic. The
consequence is higher LUT usage (3{,}275 core LUTs versus 3{,}575 for
RMSNorm) but zero DSP consumption, which is favourable since DSPs are the
most scarce resource on the ZCU102 relative to LUTs.

The \texttt{main\_loop} achieves II\,=\,4 across its two group
iterations, meaning a new group begins every 4 clock cycles. This is
determined by the inner loop over $GS = 4$ elements within each group.
Since the inner quantisation loop is pipelined at II\,=\,1, processing
four elements takes four cycles and the outer group loop can start its
next iteration once those four cycles have elapsed. The iteration latency
of 41 cycles is dominated by the two \texttt{fdiv} units
\texttt{fdiv\_32ns\_32ns\_32\_12\_no\_dsp\_1}, each with an
11-cycle pipeline depth at \SI{150}{\mega\hertz}. Two \texttt{fdiv}
instances are synthesised, one for the scale computation and one for the
per-element quantisation divisions, allowing the two operations to
overlap in the pipeline. The intermediate \texttt{QuantizedTensor} struct
fields \texttt{qt.q} and \texttt{qt.s} are inferred as small register
files (\texttt{qt\_q\_U} and \texttt{qt\_s\_U}) rather than BRAMs, since
their sizes of 8 bytes and 32 bytes respectively fall well below the
BRAM threshold. The timing closure situation is identical to RMSNorm:
estimated \SI{4.867}{\nano\second} plus \SI{1.80}{\nano\second}
uncertainty exhausts the \SI{6.667}{\nano\second} budget,
leaving zero slack.

The \texttt{quantize} implementation was not modified in the second or
third kernel iterations. The synthesis results presented here are
therefore valid across all three iterations of the prototype.

% =============================================================================
\subsection{Full TinyLlama Scale}
\label{subsec:quantize_update3}
% =============================================================================

\subsubsection{Implementation}
\label{subsubsec:quantize_update3_impl}

The sixth iteration applies four targeted changes to \texttt{quantize},
shown in Listing~\ref{lst:quantize_update3}. The
\texttt{ARRAY\_PARTITION cyclic factor} on \texttt{quantized\_buffer}
increases from 4 to 16, matching \texttt{PARALLEL\_MACS\,=\,16} and
providing 16 independent write ports for the quantisation loop.
The \texttt{runtime\_gs} parameter is removed, and the function uses the
compile-time constant \texttt{GS\,=\,64} directly. With
$\text{GS} = 64 = 2^6$, the group index computation \texttt{i/GS}
synthesises as a 6-bit right shift at no hardware cost. The
\texttt{round()} call in the element quantisation step is replaced by
manual arithmetic, where the sign of the value determines which
rounding offset is added before truncation. In the prototype,
\texttt{round()} caused HLS to infer a
\texttt{generic\_round\_double\_s} rounding IP consuming 4 BRAMs per
call. The manual rounding eliminates this IP. A reciprocal trick is
applied to the group scale, computing \texttt{inv\_scale =
1.0f/scale} once per group and multiplying per element, removing the
per-element \texttt{fdiv} that otherwise appears in each
\texttt{quantize\_elems} iteration.

\begin{lstlisting}[
  caption = {\texttt{quantize<S>} implementation from \texttt{FFN\_update3.h}
             at full TinyLlama scale ($S = 2048$, $\text{GS} = 64$).
             The reciprocal trick and manual rounding replace the
             prototype's per-element \texttt{fdiv} and \texttt{round()}
             calls. The \texttt{quantized\_buffer} partition factor
             increases from 4 to 16.},
  label   = {lst:quantize_update3}
]
template <int S>
void quantize(QuantizedTensor<S> *qx, float x[S]) {
    constexpr int   num_groups = S / GS;
    constexpr float Q_MAX      = 127.0f;

    float  scale_buffer[num_groups];
    int8_t quantized_buffer[S];
    #pragma HLS ARRAY_PARTITION variable=quantized_buffer cyclic factor=16
    #pragma HLS ARRAY_PARTITION variable=scale_buffer     cyclic factor=4

quant_groups:
    for (int group = 0; group < num_groups; group++) {
        int base_idx = group * GS;
        float wmax = 0.0f;
find_max:
        for (int i = 0; i < GS; i++) {
            #pragma HLS PIPELINE II=1
            float val = fabs(x[base_idx + i]);
            if (val > wmax) wmax = val;
        }
        float scale     = wmax / Q_MAX;
        float inv_scale = (scale != 0.0f) ? (1.0f / scale) : 0.0f;
        scale_buffer[group] = scale;
quantize_elems:
        for (int i = 0; i < GS; i++) {
            #pragma HLS PIPELINE II=1
            float quant_value = x[base_idx + i] * inv_scale;
            int32_t rounded   = (int32_t)(quant_value +
                                (quant_value >= 0.0f ? 0.5f : -0.5f));
            if (rounded >  127) rounded =  127;
            if (rounded < -127) rounded = -127;
            quantized_buffer[base_idx + i] = (int8_t)rounded;
        }
    }
    memcpy(qx->q, quantized_buffer, S          * sizeof(int8_t));
    memcpy(qx->s, scale_buffer,     num_groups * sizeof(float));
}
\end{lstlisting}

\subsubsection{C-Simulation Results}
\label{subsubsec:quantize_update3_csim}

The testbench supplied a 2048-element input vector across 32 groups of
$\text{GS} = 64$ elements each. All INT8 values and all 32 group scales
agreed with the software reference with zero error, and the csim run
completed with zero errors as reported by Vitis HLS.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:quantize_update3_csyn}

Tables~\ref{tab:quantize_u3_timing} and \ref{tab:quantize_u3_top_latency}
present the timing and latency summary. Table~\ref{tab:quantize_u3_resources}
gives the resource utilisation and Table~\ref{tab:quantize_u3_latency} the
per-stage breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{quantize\_update3\_top} at the
           \SI{150}{\mega\hertz} target clock.}
  \label{tab:quantize_u3_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    10{,}362 & 10{,}362 & \SI{69.083}{\micro\second} & \SI{69.083}{\micro\second} & 10{,}363 & 10{,}363 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{quantize\_update3\_top}
           at \SI{150}{\mega\hertz}.}
  \label{tab:quantize_u3_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  &  4{,}919 &  3{,}874 & 274{,}080 &  1 \\
    FF   &  2{,}874 &  2{,}362 & 548{,}160 & $<$1 \\
    BRAM &       22 &       18 &   1{,}824 &  1 \\
    DSP  &        5 &        5 &   2{,}520 & $<$1 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{quantize} at full scale
           ($S = 2048$, $\text{GS} = 64$, 32 groups) on the ZCU102.
           Core-only figures exclude the AXI master adapter.}
  \label{tab:quantize_u3_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Stage & Module & Latency (cycles) \\
    \midrule
    Load \texttt{x\_in}          & \texttt{VITIS\_LOOP\_28\_1}  & 2{,}056 \\
    \texttt{quantize\_2048\_s} core & \texttt{quantize\_2048\_s} & 8{,}260 \\
    Write \texttt{q\_out}, \texttt{s\_out} & \texttt{VITIS\_LOOP\_32\_2} & 35 \\
    \bottomrule
  \end{tabular}
  \caption{Per-stage latency breakdown of \texttt{quantize} at full scale
           ($S = 2048$) at \SI{150}{\mega\hertz}. Total function latency
           is 10{,}362 cycles (\SI{69.083}{\micro\second}).}
  \label{tab:quantize_u3_latency}
\end{table}

Timing meets the \SI{6.667}{\nano\second} constraint with the same
estimated path of \SI{4.867}{\nano\second} as the prototype, confirming
that the implementation changes in iteration~6 did not lengthen the
critical path.

The total latency of 10{,}362 cycles (\SI{69.083}{\micro\second}) is
dominated by the \texttt{quantize\_2048\_s} core at 8{,}260 cycles,
which accounts for 80\% of the total. Loading the 2{,}048-element input
float vector takes 2{,}056 cycles, and writing the quantised output
(\texttt{q\_out} and \texttt{s\_out}) takes only 35 cycles due to the
small output size: 2{,}048 bytes of INT8 values and 32 floats of scale
factors. The asymmetry between the 2{,}056-cycle load and the 35-cycle
write reflects the difference in transfer volume, with the input being
$2048 \times 4 = 8{,}192$ bytes versus the output's 2{,}176 bytes.

The core latency of 8{,}260 cycles for 32 groups warrants examination.
Each group executes a \texttt{find\_max} scan of 64 elements pipelined
at II\,=\,1 (64 cycles) followed by a \texttt{quantize\_elems} loop
of 64 elements also at II\,=\,1 (64 cycles), plus the overhead of one
\texttt{fdiv} to compute \texttt{scale} and one \texttt{fdiv} to compute
\texttt{inv\_scale}. At \SI{150}{\mega\hertz}, each \texttt{fdiv} unit
has a pipeline latency of approximately 11 cycles. The total per-group
cost is therefore approximately $64 + 64 + 2 \times 11 = 150$ cycles,
and across 32 groups this gives $32 \times 150 + \text{overhead}
\approx 4{,}800$ cycles for the pure arithmetic, with the remainder
from AXI burst sequencing inside the core module.

The BRAM count of 18 core BRAMs is higher than the prototype's
zero-BRAM result. At prototype scale ($S = 8$, 2 groups), all internal
buffers fit in register files. At $S = 2048$ with 32 groups, the
\texttt{quantized\_buffer[2048]} array partitioned cyclic factor\,=\,16
creates 16 banks of 128 int8 words each, which HLS maps to 16 BRAMs.
The \texttt{scale\_buffer[32]} float array with cyclic factor\,=\,4
is small enough (8 words per bank) to remain as registers, contributing
no BRAM. The remaining 2 BRAMs come from the \texttt{qt.q} and
\texttt{qt.s} fields of the output \texttt{QuantizedTensor} struct, which
at $S = 2048$ exceed the register threshold. No
\texttt{generic\_round\_double\_s} instances appear anywhere in the Memory
or Instance sections of the synthesis report, confirming that the manual
rounding arithmetic eliminates the rounding IP entirely.

The DSP count increases from 0 in the prototype to 5 in this version.
The prototype used \texttt{round()} and per-element \texttt{fdiv}, both
of which map to LUT fabric with no DSP involvement. The reciprocal trick
in iteration~6 introduces one \texttt{fmul} per element, which maps to
a 3-DSP \texttt{fmul} unit, and an \texttt{fadd} that accounts for the
remaining 2 DSPs. The \texttt{fdiv} units for \texttt{scale} and
\texttt{inv\_scale} continue to use LUT fabric, consistent with the
prototype behaviour. The net effect is a small but acceptable DSP increase
in exchange for eliminating the rounding IP's BRAM cost and the
pipeline stalls from per-element division.



% =============================================================================
\section{Dequantisation}
\label{sec:dequantize_impl}
% =============================================================================

\subsection{Implementation}
\label{subsec:dequantize_impl}

The \texttt{dequantize<S>} function is the inverse of
\texttt{quantize<S>}. For each element at index $i$, it multiplies the
INT8 value \texttt{qx->q[i]} by the corresponding group scale
\texttt{qx->s[i/GS]}, recovering an approximation of the original float.
The division \texttt{i/GS} computes the group index, and because
\texttt{GS} is a power of two, the compiler replaces this with a
bitwise right shift at no hardware cost. No pragma directives are
applied to \texttt{dequantize} in the prototype, since it operates on
a single vector element at a time with no loop-carried dependencies and
no memory partitioning decisions to make.

The maximum reconstruction error per element is bounded by
$\frac{1}{2} \cdot s = \frac{\texttt{wmax}}{254}$, which for typical
activation magnitudes of order 1.0 gives an absolute error of
approximately 0.004. This error is acceptable for the attention
projection inputs but motivates the design decision in the full
TinyLlama kernel to store KV cache values with per-group scales rather
than a single scale shared across the full \texttt{kv\_dim}-dimensional
vector. Listing~\ref{lst:dequantize} shows the complete function body.

\begin{lstlisting}[
  caption = {\texttt{dequantize<S>} implementation from \texttt{forward.h}.
             Each element is recovered by multiplying its INT8 value by
             the corresponding group scale. No pragma directives are
             applied because the loop has no carried dependencies and no
             memory port conflicts. The division \texttt{i/GS} is
             replaced by a bitwise right shift by the compiler since
             \texttt{GS} is a compile-time power of two.},
  label   = {lst:dequantize}
]
template <int S>
void dequantize(QuantizedTensor<S> *qx, float x[S], int runtime_gs)
{
    for (int i = 0; i < S; i++)
    {
        x[i] = qx->q[i] * qx->s[i / GS]; // GS is a compile-time constant
    }
}
\end{lstlisting}

\subsection{C-Simulation Results}
\label{subsec:dequantize_csim}

C-simulation was performed for \texttt{dequantize\_top}. The testbench
supplied eight known INT8 values along with two explicit group scale
factors, one per \texttt{GS\,=\,4} element group, then compared the
kernel output element by element against a software reference computing
$x_{\text{ref}}[i] = q[i] \times s_{\lfloor i/GS \rfloor}$ directly.
The same group scale factors from the \texttt{quantize} csim run
($s_0 = 0.029134$, $s_1 = 0.035433$) were reused, making the two
testbenches complementary. Table~\ref{tab:dequantize_csim} shows the
kernel and reference outputs for each element.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & HW output & SW reference & Absolute error \\
    \midrule
    0 &  1.835442 &  1.835442 & 0 \\
    1 & -2.942534 & -2.942534 & 0 \\
    2 &  3.700018 &  3.700018 & 0 \\
    3 & -1.107092 & -1.107092 & 0 \\
    4 &  2.976372 &  2.976372 & 0 \\
    5 & -4.499991 & -4.499991 & 0 \\
    6 &  1.807083 &  1.807083 & 0 \\
    7 & -4.039362 & -4.039362 & 0 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for \texttt{dequantize} with group scales
           $s_0 = 0.029134$ (indices 0--3) and $s_1 = 0.035433$
           (indices 4--7). All eight dequantised float values match the
           software reference with zero absolute error.}
  \label{tab:dequantize_csim}

\end{table}

All eight dequantised float values agree with the software
reference, with zero absolute error across every element. Elements at
indices 2 and 5, corresponding to the INT8 boundary values $127$ and
$-127$, correctly reconstruct to $3.700018$ and $-4.499991$
respectively, confirming that the scale multiplication path handles
boundary saturation without loss. The zero-error result is expected for
dequantisation. Unlike quantisation, which introduces an irreducible
rounding error at the point of integer conversion, dequantisation is a
pure floating-point multiplication with no rounding decision, and the
kernel output is therefore bit-identical to the sequential software
reference on every element. The csim run completed with zero errors as
reported by Vitis HLS.

\subsection{HLS Synthesis Results}
\label{subsec:dequantize_csyn}

The \texttt{dequantize} function was synthesised in isolation under
identical conditions to \texttt{quantize}.
Tables~\ref{tab:dequantize_timing} and \ref{tab:dequantize_top_latency}
present the top-level timing and latency summary, followed by
Table~\ref{tab:dequantize_resources} for resource utilisation and
Table~\ref{tab:dequantize_latency} for the per-stage breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{dequantize\_top} at
           \SI{150}{\mega\hertz}.}
  \label{tab:dequantize_timing}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    45 & 45 & \SI{0.300}{\micro\second} & \SI{0.300}{\micro\second} & 46 & 46 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{dequantize\_top} at
           \SI{150}{\mega\hertz}.}
  \label{tab:dequantize_top_latency}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 3{,}128 &   273 & 274{,}080 & 1 \\
    FF   & 2{,}244 &   372 & 548{,}160 & $<$1 \\
    BRAM & 4       &     0 & 1{,}824   & $<$1 \\
    DSP  & 3       &     3 & 2{,}520   & $<$1 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{dequantize} synthesised at
           $S = 8$ on the ZCU102. The core-only column excludes the
           AXI master adapter.}
  \label{tab:dequantize_resources}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Stage & Trip count & II & Iteration latency & Total (cycles) \\
    \midrule
    \texttt{VITIS\_LOOP\_55\_1} (dequantise loop) & 8 & 1 & 12 & 18 \\
    \bottomrule
  \end{tabular}
  \caption{Latency breakdown of \texttt{dequantize} ($S = 8$, $GS = 4$)
           at \SI{150}{\mega\hertz}. Total function latency is 45 cycles
           (\SI{300}{\nano\second}).}
  \label{tab:dequantize_latency}

\end{table}

There is an asymmetry between the two
inverse operations. \texttt{Dequantize} is nearly twice as fast (45
cycles versus 84 cycles), uses only a twelfth of the core LUTs
(273 versus 3{,}275), and achieves a perfect II\,=\,1 on its inner
loop. The root cause of this asymmetry is the difference between the
arithmetic operations each function requires.

\texttt{Dequantize} performs a single floating-point multiplication per
element, $x_i = q_i \times s_{\lfloor i/GS \rfloor}$, which maps to one
\texttt{fmul} \texttt{fmul\_32ns\_32ns\_32\_4\_max\_dsp\_1} unit using 3 DSP blocks with a 3-cycle pipeline latency.
Since each element's computation depends only on its own INT8 value and its
group scale with no dependency on any previous element, there are no loop-carried
dependencies, and II\,=\,1 is achievable. The 12-cycle iteration latency
is dominated by the AXI read latency for loading \texttt{q\_in} and
\texttt{s\_in} from DDR, with the 3-cycle \texttt{fmul} pipeline depth
making up the remainder.

For the \texttt{quantize} function, two floating-point divisions are required per
group, one to compute $s = \texttt{wmax}/127$ and one per element for
$\hat{x}_i = x_i/s$. Unlike multiplication, floating-point division has
no mapping to Xilinx DSP48 blocks and is implemented entirely in LUT
fabric, as confirmed by the \texttt{no\_dsp} suffix on both
\texttt{fdiv} instances in the Bind Op report. This explains why
\texttt{quantize} uses zero DSPs while using significantly more LUTs.
The 11-cycle \texttt{fdiv} latency at \SI{150}{\mega\hertz}, combined
with the sequential dependency between scale computation and element
quantisation within each group, is what drives the 41-cycle iteration
latency in \texttt{main\_loop}.

The \texttt{quantize} Bind Storage
report also shows two small ROM lookup tables,
\texttt{mask\_table} and \texttt{one\_half\_table}, which HLS infers to
implement the IEEE~754 round-to-nearest operation used by the
\texttt{round()} call. These tables are inferred in register fabric
rather than BRAM at this prototype scale and contribute to the higher
LUT count. The AXI data width report for \texttt{dequantize} also shows
a difference in that the \texttt{q\_in} port operates at an 8-bit
native width widened to a 32-bit AXI bus width (8\,$\rightarrow$\,32),
reflecting the fact that INT8 data must be packed into the 32-bit AXI
data bus in groups of four bytes per transfer.

The \texttt{dequantize} implementation was not modified in the second or
third kernel iterations. The synthesis results presented here are
therefore valid across all three iterations of the prototype.

% =============================================================================
\subsection{Full TinyLlama Scale}
\label{subsec:dequantize_update3}
% =============================================================================

\subsubsection{Implementation}
\label{subsubsec:dequantize_update3_impl}

The sixth iteration applies two changes to \texttt{dequantize}, shown in
Listing~\ref{lst:dequantize_update3}. The \texttt{runtime\_gs} parameter
is removed and the function uses the compile-time constant
\texttt{GS\,=\,64} directly. A \texttt{\#pragma HLS PIPELINE II\,=\,1}
directive is added to the \texttt{dequant\_loop}.

\begin{lstlisting}[
  caption = {\texttt{dequantize<S>} from \texttt{FFN\_update3.h} at full
             TinyLlama scale ($S = 2048$, $\text{GS} = 64$). The only
             structural change from the prototype is the explicit
             \texttt{PIPELINE II=1} annotation and removal of the
             \texttt{runtime\_gs} parameter.},
  label   = {lst:dequantize_update3}
]
template <int S>
void dequantize(QuantizedTensor<S> *qx, float x[S]) {
dequant_loop:
    for (int i = 0; i < S; i++) {
        #pragma HLS PIPELINE II=1
        x[i] = qx->q[i] * qx->s[i / GS];
    }
}
\end{lstlisting}

The algorithmic content is unchanged. For each element at index $i$, the
INT8 value \texttt{qx->q[i]} is multiplied by the group scale
\texttt{qx->s[i/GS]}, where the division \texttt{i/GS} synthesises as
a 6-bit right shift since $\text{GS} = 64 = 2^6$. The added
\texttt{PIPELINE II\,=\,1} annotation does not alter the datapath and
simply makes the throughput constraint visible in the schedule report.

\subsubsection{C-Simulation Results}
\label{subsubsec:dequantize_update3_csim}

The testbench supplied 2048 INT8 values with a uniform scale of $0.03$.
All 2048 output elements agreed with the software reference with zero
absolute error, and the csim run completed with zero errors as reported
by Vitis HLS.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:dequantize_update3_csyn}

Tables~\ref{tab:dequantize_u3_timing} and \ref{tab:dequantize_u3_top_latency}
present the timing and latency summary. Table~\ref{tab:dequantize_u3_resources}
gives the resource utilisation and Table~\ref{tab:dequantize_u3_latency} the
per-stage breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{dequantize\_update3\_top} at the
           \SI{150}{\mega\hertz} target clock.}
  \label{tab:dequantize_u3_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    6{,}223 & 6{,}223 & \SI{41.489}{\micro\second} & \SI{41.489}{\micro\second} & 6{,}224 & 6{,}224 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{dequantize\_update3\_top}
           at \SI{150}{\mega\hertz}.}
  \label{tab:dequantize_u3_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 3{,}003 & 2{,}423 & 274{,}080 & 1 \\
    FF   & 1{,}855 & 1{,}343 & 548{,}160 & $<$1 \\
    BRAM &      22 &      18 &   1{,}824 & 1 \\
    DSP  &       3 &       3 &   2{,}520 & $<$1 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{dequantize} at full scale
           ($S = 2048$, $\text{GS} = 64$) on the ZCU102. Core-only
           figures exclude the AXI master adapter (\texttt{gmem0\_m\_axi}).}
  \label{tab:dequantize_u3_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Stage & Module & Latency (cycles) \\
    \midrule
    Load \texttt{q\_in}    & \texttt{VITIS\_LOOP\_25\_1}       & 2{,}058 \\
    Load \texttt{s\_in}    & \texttt{VITIS\_LOOP\_29\_2}       &      35 \\
    \texttt{dequant\_loop} & \texttt{Pipeline\_dequant\_loop}  & 2{,}060 \\
    Write \texttt{x\_out}  & \texttt{VITIS\_LOOP\_39\_3}       & 2{,}051 \\
    \bottomrule
  \end{tabular}
  \caption{Per-stage latency breakdown of \texttt{dequantize} at full scale
           ($S = 2048$) at \SI{150}{\mega\hertz}. Total function latency
           is 6{,}223 cycles (\SI{41.489}{\micro\second}).}
  \label{tab:dequantize_u3_latency}
\end{table}

Timing meets the \SI{6.667}{\nano\second} constraint with the same
\SI{4.867}{\nano\second} estimated path as the prototype. The added
\texttt{PIPELINE II\,=\,1} annotation does not affect the critical path
since \texttt{dequantize} already achieved II\,=\,1 in the prototype.

The total latency of 6{,}223 cycles is distributed across four sequential
stages. Loading the 2{,}048 INT8 input values from DDR takes 2{,}058
cycles. Loading the 32 float scale values takes 35 cycles, since the
transfer is only $32 \times 4 = 128$ bytes. The \texttt{dequant\_loop}
executes for 2{,}060 cycles at II\,=\,1. Writing the 2{,}048 float
outputs takes 2{,}051 cycles. The total is therefore dominated by AXI
transfer time rather than arithmetic, since the three
2{,}048-element transfer stages each cost approximately 2{,}055 cycles
while the compute itself takes only one \texttt{fmul} per element pipelined
at II\,=\,1.

The 18 core BRAMs arise from the scale-up from $S = 8$ to $S = 2048$.
At prototype scale all buffers fitted in register files. At full scale
the output buffer \texttt{x\_buf[2048]}, partitioned cyclic
factor\,=\,16, creates 16 banks of 128 floats each. At 128 words per
bank, each bank exceeds the register inference threshold and maps to a
BRAM, giving 16 BRAMs. The \texttt{qt.q[2048]} field of the input
\texttt{QuantizedTensor} struct adds 1 BRAM for the 2{,}048-byte INT8
array, and \texttt{qt.s[32]} adds 1 further BRAM for the 32 float
scales, totalling 18 core BRAMs. The 3 DSPs come from the single
\texttt{fmul} unit in the \texttt{dequant\_loop} pipeline, unchanged
from the prototype.



% =============================================================================
\section{Matrix--Vector Multiplication}
\label{sec:matmul_impl}
% =============================================================================

\subsection{Implementation}
\label{subsec:matmul_impl}

The \texttt{matmul<N,D>} function computes the quantised matrix--vector
product $\mathbf{W}\mathbf{x} \to \mathbf{y}$, where $\mathbf{W} \in
\mathbb{Z}^{D \times N}$ is an INT8 weight matrix stored in row-major
order and $\mathbf{x} \in \mathbb{Z}^N$ is an INT8 input vector. Both
operands are accompanied by their group-wise scale factors, and the output
$\mathbf{y} \in \mathbb{R}^D$ is returned as a dequantised float vector.
The function is declared as a C++ template in \texttt{forward.h},
parameterised by \texttt{N} (input dimension) and \texttt{D} (output
dimension), and is instantiated at the call sites in \texttt{forward.cpp}
for the Q, K, V, and output projections. Listing~\ref{lst:matmul} shows
the core loop structure.

\begin{lstlisting}[
  caption = {Core loop structure of \texttt{matmul<N,D>} from
             \texttt{forward.h}. The outer row loop is pipelined. The
             inner loading loops (\texttt{matmul1}, \texttt{matmul2})
             and the grouped dot product loops (\texttt{matmul3},
             \texttt{matmul4}) execute sequentially within each pipeline
             iteration. The \texttt{UNROLL} pragmas on \texttt{matmul1}
             and \texttt{matmul2} are commented out to prevent AXI
             AXI bus contention at full scale.},
  label = {lst:matmul}
]
for (i = 0; i < D; i++) {
  #pragma HLS PIPELINE
  float val = 0.0f;
  int8_t  w_buffer[N];   float ws_buffer[N/GS];
  #pragma HLS ARRAY_PARTITION variable=w_buffer  cyclic factor=32
  #pragma HLS ARRAY_PARTITION variable=ws_buffer cyclic factor=32

matmul1: for (int j = 0; j < N; j++)   // no UNROLL: deliberate
    w_buffer[j]  = wq[j + i*N];
matmul2: for (int j = 0; j < N/GS; j++) // no UNROLL: deliberate
    ws_buffer[j] = ws[i*(N/GS) + j];

matmul3: for (int j = 0; j <= N-GS; j += GS) {
    int32_t ival = 0;
matmul4:  for (int k = 0; k < GS; k++) // UNROLL commented out
        ival += (int32_t)x_buffer[j+k] * (int32_t)w_buffer[j+k];
    val += (float)ival * ws_buffer[j/GS] * xs_buffer[j/GS];
  }
  xout[i] = val;
}
\end{lstlisting}

The implementation is organised into three phases. The first phase
pre-loads the input vector and its associated scale factors into two
static local buffers before any row computation begins. The input buffer
\texttt{x\_buffer[N]} receives the INT8 input values via the
\texttt{x\_buff} loop with \texttt{UNROLL factor=16}, and the scale buffer
\texttt{xs\_buffer[N/GS]} receives the per-group float scales via the
\texttt{xs\_buff} loop with \texttt{UNROLL factor=4}. Both buffers are declared \texttt{static} so that HLS allocates them
as standalone addressable memory structures rather than dissolving them
into pipeline wires within the datapath. This is necessary because
\texttt{ARRAY\_PARTITION} can only be applied to a dedicated memory
structure with a fixed address space. Without \texttt{static}, HLS may
embed the array directly into the pipeline datapath as a set of
registers without creating a separately partitionable memory bank.
Both are partitioned with
\texttt{ARRAY\_PARTITION cyclic} to provide sufficient read ports for the
downstream compute loops. Pre-loading into local buffers serves the same
purpose as in \texttt{rmsnorm}, separating the AXI burst reads for the
input from the per-row AXI burst reads for the weights, preventing
contention on the shared \texttt{gmem0} port.

The second and third phases execute inside the outer row loop over
\texttt{D} output elements, which has the \texttt{\#pragma HLS
PIPELINE} directive. For each row \texttt{i}, HLS infers two
non-static local buffers, \texttt{w\_buffer[N]} and
\texttt{ws\_buffer[N/GS]}, each partitioned with \texttt{ARRAY\_PARTITION
cyclic factor=32}. The \texttt{matmul1} loop loads the \texttt{N} INT8
weight values for row \texttt{i} from DDR into \texttt{w\_buffer}, and
\texttt{matmul2} loads the corresponding \texttt{N/GS} scale factors into
\texttt{ws\_buffer}.

The third phase is the grouped integer dot product. The \texttt{matmul3}
loop iterates over groups of \texttt{GS} consecutive elements, and within
each group the \texttt{matmul4} loop accumulates the products
into a 32-bit integer accumulator \texttt{ival}. Using \texttt{int32\_t}
for the accumulator is needed because the product of two INT8 values can reach
$127 \times 127 = 16{,}129$, and summing \texttt{GS = 4} such products
gives a maximum of $4 \times 16{,}129 = 64{,}516$, which fits within
\texttt{int16\_t} but would overflow \texttt{int8\_t}. Accumulating into
\texttt{int32\_t} prevents any possibility of overflow. After all
\texttt{GS} products have been accumulated, the group result is
dequantised inline. The integer sum is cast to float and multiplied by the
product of the weight group scale \texttt{ws\_buffer[j/GS]} and the input
group scale \texttt{xs\_buffer[j/GS]}, recovering the floating-point
partial sum for that group. This per-group dequantisation is added
directly to the running float accumulator \texttt{val} before moving to
the next group.

\subsection{C-Simulation Results}
\label{subsec:matmul_csim}

C-simulation was performed for \texttt{matmul\_top}. The testbench
constructs an $8 \times 8$ weight matrix whose diagonal entries are all
127 and whose off-diagonal entries are all zero, with all weight scales
set to $1/127$. With this configuration the computation simplifies to
$\texttt{xout}[i] = \texttt{xq}[i] \times \texttt{xs}[\lfloor i/GS
\rfloor]$, which is the dequantised input vector. The testbench
reuses the quantised values and scale factors from the \texttt{quantize}
csim as input, making the three testbenches a matched chain in which
quantise, identity matrix--vector multiply, and dequantise can be
cross-validated against one another. Table~\ref{tab:matmul_csim} shows the output from both the kernel and the
software reference for each of the eight output elements.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & HW output & SW reference & Absolute error \\
    \midrule
    0 &  0.495278 &  0.495278 & 0 \\
    1 & -1.194494 & -1.194494 & 0 \\
    2 &  3.700018 &  3.700018 & 0 \\
    3 & -0.291340 & -0.291340 & 0 \\
    4 &  2.090547 &  2.090547 & 0 \\
    5 & -4.499991 & -4.499991 & 0 \\
    6 &  0.992124 &  0.992124 & 0 \\
    7 & -2.799207 & -2.799207 & 0 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for \texttt{matmul} with quantised input
           $\hat{\mathbf{x}} = [17, -41, 127, -10, 59, -127, 28, -79]$
           (group scales $s_0 = 0.029134$, $s_1 = 0.035433$) and an
           $8 \times 8$ scaled identity weight matrix. All eight output
           elements match the software reference within tolerance.}
  \label{tab:matmul_csim}

\end{table}

All eight output elements agree between the hardware kernel and
the software reference. The expected round-trip relationship holds, with
output values being the dequantised equivalents of the original float inputs
$[0.5, -1.2, 3.7, -0.3, 2.1, -4.5, 1.0, -2.8]$ to within the precision
limits of symmetric INT8 quantisation. The csim run completed with zero errors as reported by
Vitis HLS.

\subsection{HLS Synthesis Results}
\label{subsec:matmul_csyn}

The \texttt{matmul} function was synthesised in isolation using the
prototype configuration ($N = D = 8$, $GS = 4$) on the
\texttt{xczu9eg-ffvb1156-2-e} device at a \SI{150}{\mega\hertz} target
clock (\SI{6.667}{\nano\second} period).
Tables~\ref{tab:matmul_timing} and \ref{tab:matmul_top_latency} present
the top-level timing and latency summary, followed by
Table~\ref{tab:matmul_resources} for resource utilisation and
Table~\ref{tab:matmul_latency} for the per-stage latency breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock   & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667       & 4.867          & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{matmul\_top} at the
           \SI{150}{\mega\hertz} (\SI{6.667}{\nano\second}) target clock.}
  \label{tab:matmul_timing}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    146 & 146 & \SI{0.973}{\micro\second} & \SI{0.973}{\micro\second} & 147 & 147 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{matmul\_top} at
           \SI{150}{\mega\hertz}.}
  \label{tab:matmul_top_latency}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 5{,}577 & 2{,}506 & 274{,}080 & 2 \\
    FF   & 4{,}163 & 1{,}944 & 548{,}160 & $<$1 \\
    BRAM & 4       & 0       & 1{,}824   & $<$1 \\
    DSP  & 9       & 9       & 2{,}520   & $<$1 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{matmul} synthesised at $N = D
           = 8$ on the ZCU102 (\texttt{xczu9eg-ffvb1156-2-e}). The
           core-only column excludes the AXI master adapter and S-AXILite
           control interface, which are not present when the function is
           inlined into the full kernel.}
  \label{tab:matmul_resources}

\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Stage & Trip count & II & Iteration latency & Total (cycles) \\
    \midrule
    Pre-loop buffer loads (\texttt{x\_buff}, \texttt{xs\_buff}) & — & — & — & 24 \\
    \texttt{VITIS\_LOOP\_143\_1} (row loop)                     & 8 & 10 & 42 & 120 \\
    \bottomrule
  \end{tabular}
  \caption{Latency breakdown of \texttt{matmul} ($N = D = 8$, $GS = 4$)
           at \SI{150}{\mega\hertz}. Total function latency is 146 cycles
           (\SI{973}{\nano\second}).}
  \label{tab:matmul_latency}

\end{table}

Before any
arithmetic can begin on row~\texttt{i}, the pipeline must first issue AXI
reads for the \texttt{N = 8} INT8 weight values from \texttt{matmul1} and
the \texttt{N/GS = 2} float scale values from \texttt{matmul2}. Because
neither loading loop has an \texttt{UNROLL} pragma, HLS issues these
reads sequentially, with eight weight reads followed by two scale reads, giving
$8 + 2 = 10$ sequential AXI transactions before the first arithmetic
operation of the iteration can complete. This sequential load count
directly determines the achieved II of 10.

The iteration latency of 42 cycles reflects the time required to complete
the full body of one row iteration. The sequential weight and scale loads
are followed by the grouped integer dot product across two groups of four
elements each.
The total row loop latency is therefore $\text{II} \times (D - 1) +
\text{iteration latency} = 10 \times 7 + 42 = 112$ cycles for the
arithmetic, with the remaining 8 cycles accounting for pipeline drain and
sub-module handshaking overhead. The 24-cycle pre-loop overhead covers the
AXI address setup and the two buffer-loading loops for the input vector
and its scales, which execute once before any row computation begins.

The DSP utilisation of 9 breaks down as follows. The Bind Op report
identifies four \texttt{mac\_muladd\_8s\_8s\_16s\_17} units (one per INT8
element of the \texttt{matmul4} inner loop within one group, using 1
DSP each), one \texttt{fmul\_32ns} unit for the floating-point
dequantisation product $\texttt{ival} \times \texttt{ws} \times
\texttt{xs}$ (3 DSPs), and one \texttt{fadd\_32ns} unit for the
float accumulator \texttt{val} (2 DSPs), totalling $4 + 3 + 2 = 9$. An
additional four \texttt{mul\_8s\_8s\_16} instances appear in the report
with zero DSP assignment. These correspond to the second group's INT8
multiplications, which HLS chose to implement in LUT fabric rather than
DSP48 blocks. The total DSP count of 9
is higher than the zero DSPs consumed by \texttt{quantize} and the
3 DSPs consumed by \texttt{dequantize}. The timing situation is
identical to all preceding sections, with an estimated \SI{4.867}{\nano\second}
plus \SI{1.800}{\nano\second} uncertainty that exhausts the
\SI{6.667}{\nano\second} budget, leaving zero slack.

The \texttt{matmul} implementation was not modified in the second or
third kernel iterations. The synthesis results presented here are
therefore valid across all three iterations of the prototype.

% =============================================================================
\subsection{Full TinyLlama Scale: Tiled Implementation}
\label{subsec:matmul_update3}
% =============================================================================

\subsubsection{Implementation}
\label{subsubsec:matmul_update3_impl}

The sixth iteration replaces the prototype matmul with a tiled
architecture, shown in Listing~\ref{lst:matmul_update3}. The prototype
used \texttt{\#pragma HLS PIPELINE} on the outer row loop, which was safe
at $N = 8$ because only $8 + 2 = 10$ AXI reads were issued per outer
iteration, staying within the AXI adapter's request queue. At full scale
$N = 2048$, the same approach would issue $2048 + 32 = 2{,}080$ simultaneous
reads per iteration, saturating the queue and stalling the pipeline. The
tiled design resolves this by separating weight loading from computation
into explicit phases.

\begin{lstlisting}[
  caption = {Core tile loop structure of the tiled \texttt{matmul<N,D>}
             from \texttt{FFN\_update3.h}. Each tile loads
             \texttt{TILE\_D\,=\,4} weight rows via \texttt{memcpy} bursts
             (Phase 1), computes \texttt{TILE\_D} dot products from BRAM
             (Phase 2), and writes results (Phase 3). The
             \texttt{dot\_inner} unroll factor matches
             \texttt{PARALLEL\_MACS\,=\,16}.},
  label   = {lst:matmul_update3}
]
tile_loop:
for (int tile = 0; tile < num_full_tiles + (remainder>0?1:0); tile++) {
    #pragma HLS loop_tripcount min=1 max=1408

    const int d_start = tile * TILE_D;
    const int d_count = (tile < num_full_tiles) ? TILE_D : remainder;

    // Phase 1: burst-load TILE_D weight rows from DDR
load_tile:
    for (int d = 0; d < d_count; d++) {
        memcpy(w_tile[d],  &wq[d_start+d)*N],  N        * sizeof(int8_t));
        memcpy(ws_tile[d], &ws[(d_start+d)*(N/GS)], (N/GS) * sizeof(float));
    }

    // Phase 2: compute TILE_D dot products from BRAM
compute_tile:
    for (int d = 0; d < d_count; d++) {
        float val = 0.0f;
dot_groups:
        for (int g = 0; g < N/GS; g++) {
            #pragma HLS PIPELINE II=1
            int32_t ival = 0;
dot_inner:
            for (int k = 0; k < GS; k++) {
                #pragma HLS UNROLL factor=PARALLEL_MACS
                ival += (int32_t)x_buf[g*GS+k] * (int32_t)w_tile[d][g*GS+k];
            }
            val += (float)ival * ws_tile[d][g] * xs_buf[g];
        }
        out_tile[d] = val;
    }

    // Phase 3: write tile output
write_tile:
    for (int d = 0; d < d_count; d++) {
        #pragma HLS PIPELINE II=1
        xout[d_start + d] = out_tile[d];
    }
}
\end{lstlisting}

The template function \texttt{matmul<N,D>} processes \texttt{TILE\_D\,=\,4}
output rows per tile. Phase 1 uses \texttt{memcpy} to burst-read each of
the \texttt{TILE\_D} weight rows contiguously from DDR into the local BRAM
tile \texttt{w\_tile[TILE\_D][N]}. Using \texttt{memcpy} is deliberate, as it
causes HLS to infer a single burst transaction per row rather than
individual scattered element reads, which would each consume one
transaction slot in the AXI adapter queue. Phase 2 computes \texttt{TILE\_D}
dot products entirely from the BRAM tile with no DDR access. The innermost
\texttt{dot\_inner} loop carries the \texttt{UNROLL factor=PARALLEL\_MACS}
pragma, giving 16 parallel INT8 multiply-accumulate operations per group of
\texttt{GS\,=\,64} elements and allowing each group's 64 multiply-accumulate
operations to be completed in $64/16 = 4$ pipeline cycles. Phase 3 writes
the \texttt{TILE\_D} float results to the output buffer, pipelined at
II\,=\,1.

Two compile-time constants govern the area-performance trade-off:
\texttt{PARALLEL\_MACS\,=\,16} sets the inner MAC parallelism, and
\texttt{TILE\_D\,=\,4} sets the number of rows buffered per tile. The
weight tile occupies $4 \times 2048 = 8{,}192$ bytes of INT8 plus
$4 \times 32 = 128$ floats, approximately 8.7\,KB per matmul call.
Load and compute phases run sequentially within each tile. A
double-buffering scheme overlapping tile $k+1$'s load with tile $k$'s
compute is identified in the source comments as a future optimisation
that would reduce the latency by approximately half.

\subsubsection{C-Simulation Results}
\label{subsubsec:matmul_update3_csim}

The testbench computed \texttt{matmul<2048,\,5632>} using fixed-pattern
weights and compared the first 8 of 5632 output elements against a software
reference. All outputs agreed with zero absolute error, the testbench
confirmed 1{,}408 full tiles and 0 remainder tiles ($5632/4 = 1408$
exactly), and the csim run completed with zero errors as reported by
Vitis HLS.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:matmul_update3_csyn}

Tables~\ref{tab:matmul_u3_timing} and \ref{tab:matmul_u3_top_latency}
present the timing and latency summary. Table~\ref{tab:matmul_u3_resources}
gives the resource utilisation and Table~\ref{tab:matmul_u3_latency} the
per-stage breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{matmul\_update3\_top} at the
           \SI{150}{\mega\hertz} target clock.}
  \label{tab:matmul_u3_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    12{,}752{,}954 & 12{,}752{,}954 & \SI{85.024}{\milli\second} & \SI{85.024}{\milli\second} & 12{,}752{,}955 & 12{,}752{,}955 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{matmul\_update3\_top}
           at \SI{150}{\mega\hertz}.}
  \label{tab:matmul_u3_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 12{,}438 & 11{,}111 & 274{,}080 &  4 \\
    FF   &  5{,}631 &  4{,}402 & 548{,}160 &  1 \\
    BRAM &       84 &       80 &   1{,}824 &  4 \\
    DSP  &       35 &       35 &   2{,}520 &  1 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of the tiled \texttt{matmul} at full scale
           ($N = 2048$, $D = 5632$) on the ZCU102. Core-only figures
           exclude the AXI master adapter.}
  \label{tab:matmul_u3_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Stage & Module & Latency (cycles) \\
    \midrule
    \texttt{matmul\_2048\_5632\_s} core & \texttt{matmul\_2048\_5632\_s} & 12{,}752{,}952 \\
    AXI wrapper overhead & & 2 \\
    \bottomrule
  \end{tabular}
  \caption{Latency breakdown of the tiled \texttt{matmul}
           ($N = 2048$, $D = 5632$, \texttt{TILE\_D} = 4) at
           \SI{150}{\mega\hertz}. Total function latency is
           12{,}752{,}954 cycles (\SI{85.024}{\milli\second}).}
  \label{tab:matmul_u3_latency}
\end{table}

Timing meets the \SI{6.667}{\nano\second} constraint with the same
\SI{4.867}{\nano\second} estimated path as the prototype. The tiled
design does not alter the innermost arithmetic, which remains an INT8
multiply-accumulate accumulated into an INT32 partial sum followed by
a float dequantisation step, so the critical path is unchanged.

The latency of 12{,}752{,}954 cycles (\SI{85.024}{\milli\second}) is the
highest of any scalar primitive in this project. With 1{,}408 tiles
($D/\texttt{TILE\_D} = 5632/4$), each tile costs approximately
$12{,}752{,}952 / 1{,}408 \approx 9{,}058$ cycles. Within each tile,
Phase 1 loads 4 weight rows from DDR, each a contiguous 2{,}048-byte
burst. At \SI{150}{\mega\hertz} and 32-bit AXI bus width, transferring
2{,}048 bytes requires $2048/4 = 512$ bus cycles, so loading 4 rows
takes $4 \times 512 = 2{,}048$ cycles. Phase 2 then computes 4 dot
products over $N/\texttt{GS} = 32$ groups of 64 elements each.
With \texttt{PARALLEL\_MACS\,=\,16}, each group of 64 elements is
processed in $64/16 = 4$ cycles, so Phase 2 costs $4 \times 32 \times 4 =
512$ cycles. Phase 3 writes 4 float results in 4 cycles. The tile cost
is therefore approximately $2{,}048 + 512 + 4 = 2{,}564$ cycles per
4-row tile, but the measured $9{,}058$ cycles per tile indicates
additional overhead from the inter-phase control and the
\texttt{loop\_tripcount}-guided scheduling decisions. Load and compute
phases are sequential, so the load cost dominates. A double-buffering
scheme that issues tile $k+1$'s DDR read while tile $k$'s Phase 2 is
executing would approximately halve the total latency to
\SI{44}{\milli\second}, as identified in the source comments.

The 80 core BRAMs hold the on-chip weight tile buffers. The
\texttt{w\_tile[TILE\_D][N]} array is partitioned with
\texttt{ARRAY\_PARTITION complete dim=1} on the row dimension, making all
4 rows addressable simultaneously, and \texttt{cyclic
factor=PARALLEL\_MACS dim=2} on the column dimension, giving
$4 \times 16 = 64$ independent BRAM banks. The
\texttt{ws\_tile[TILE\_D][N/GS]} array contributes a further
$4 \times 4 = 16$ banks, for 80 core BRAMs in total. The input buffers
\texttt{x\_buf[N]} and \texttt{xs\_buf[N/GS]} are loaded once before
the tile loop and contribute no additional BRAMs beyond what is counted
in the prototype matmul.

The 35 DSPs come from the floating-point accumulation in
\texttt{compute\_tile}: the expression \texttt{val += (float)ival *
ws\_tile[d][g] * xs\_buf[g]} requires two \texttt{fmul} units (6 DSPs)
and one \texttt{fadd} unit (2 DSPs). However, the synthesis report shows
35 DSPs, not 8, because HLS infers multiple instances of these units to
serve the 4-row \texttt{compute\_tile} loop. With \texttt{TILE\_D\,=\,4},
HLS may schedule the four rows' accumulators in parallel where ports
allow, instantiating additional \texttt{fmul}/\texttt{fadd} units for
each concurrent row, bringing the total to 35. The INT8 multipliers
inside \texttt{dot\_inner} map to LUT fabric rather than DSPs, consistent
with the prototype behaviour where \texttt{mul\_8s\_8s} instances consume
no DSPs.



% =============================================================================
\section{Softmax}
\label{sec:softmax_impl}
% =============================================================================

\subsection{Implementation}
\label{subsec:softmax_impl}

The \texttt{softmax<MAXSIZE>} function converts a vector of raw attention
scores into a probability distribution over sequence positions. It is
declared as a C++ template in \texttt{forward.h}, parameterised by
\texttt{MAXSIZE}, which sets the size of the local working buffer
allocated on-chip. The function accepts a runtime integer \texttt{size}
that specifies how many of the \texttt{MAXSIZE} elements are active for
the current call, reflecting the fact that the number of attended
positions grows with each decode step from 1 up to the maximum context
length. In the baseline kernel, the template is instantiated as
\texttt{softmax<16>}, matching the prototype's \texttt{seq\_len = 16}.
The second iteration of the kernel widens this to
\texttt{softmax<257>}, accommodating larger context windows without
any code change to the function body, at the cost of a larger static
on-chip buffer allocation.

The implementation uses the numerically stable max-subtract form,
decomposed into four sequential named loops. Each loop addresses a
distinct phase of the computation with its own pragma strategy, since
the dependencies between phases prevent them from being merged into a
single pipelined stage.

The first loop, \texttt{max}, scans the active \texttt{size} elements
to find the maximum score value. It has a \texttt{\#pragma HLS
PIPELINE} directive and a \texttt{loop\_tripcount} hint, but no
\texttt{UNROLL}. The reason is that finding the maximum is a
sequential reduction, as each comparison must read the current running
maximum and potentially update it before the next iteration can
proceed. This loop-carried dependency on \texttt{max\_val} prevents
parallelism in the same way as the \texttt{sum\_of\_squares} reduction
in \texttt{rmsnorm}.

The second loop, \texttt{exp}, subtracts \texttt{max\_val} from each
element and applies the exponential function, writing results into the
local \texttt{buffer[MAXSIZE]} array. This loop carries both
\texttt{\#pragma HLS PIPELINE} and \texttt{\#pragma HLS UNROLL
factor=16}. Because each element's computation is entirely
independent, the \texttt{UNROLL factor=16} on a
\texttt{MAXSIZE = 16}-element loop amounts to complete unrolling. The exponential function
\texttt{expf} is an operation with no direct arithmetic circuit equivalent that Vitis HLS maps to
the Xilinx \texttt{fexp} floating-point IP core, which requires 7 DSP
blocks and an 8-cycle pipeline depth. The cost of this operation
dominates the resource budget of the entire function, as discussed in
the synthesis subsection.

The third loop, \texttt{sum}, accumulates the sum of all
\texttt{buffer} entries. It carries only a \texttt{loop\_tripcount}
hint and no \texttt{PIPELINE} or \texttt{UNROLL} pragma, because the
loop-carried dependency on \texttt{sum} means it cannot achieve
II\,=\,1 regardless.

The fourth loop, \texttt{norm}, divides each \texttt{buffer} entry by
the total sum to produce the final probability values. Rather than
performing a floating-point division per element, the implementation
precomputes the reciprocal \texttt{inv\_sum = 1.0f / sum} as a single
scalar before the loop, then multiplies each element by
\texttt{inv\_sum}. This replaces \texttt{MAXSIZE} divisions with one
division and \texttt{MAXSIZE} multiplications, which is a significant
saving since \texttt{fdiv} \texttt{fdiv\_32ns\_32ns\_32\_12\_no\_dsp\_1}
requires 11 cycles and cannot map to
DSP48 blocks, whereas \texttt{fmul} \texttt{fmul\_32ns\_32ns\_32\_4\_max\_dsp\_1}
requires 3 cycles and maps
efficiently to DSPs. Like the \texttt{exp} loop, \texttt{norm} carries
\texttt{PIPELINE} and \texttt{UNROLL factor=16}, fully unrolling the
16-element loop into 16 parallel multiply lanes.
Listing~\ref{lst:softmax} shows the four-loop structure and its pragma
assignments.

\begin{lstlisting}[
  caption = {\texttt{softmax<MAXSIZE>} implementation from
             \texttt{forward.h} (baseline iteration). The \texttt{max}
             and \texttt{sum} loops carry no \texttt{UNROLL} because of
             their loop-carried dependencies on \texttt{max\_val} and
             \texttt{sum} respectively. The \texttt{exp} and \texttt{norm}
             loops are fully unrolled into 16 parallel datapath lanes.
             The \texttt{inv\_sum} reciprocal hoists the single division
             out of the \texttt{norm} loop body.},
  label = {lst:softmax}
]
template <int MAXSIZE>
void softmax(float *x, int size) {
  float buffer[MAXSIZE];
  float max_val = x[0];
max:
  for (int i = 1; i < size; i++) {
    #pragma HLS loop_tripcount min=1 max=16 avg=8
    #pragma HLS PIPELINE
    if (x[i] > max_val) max_val = x[i];   // carried dep: no UNROLL
  }
exp:
  for (int i = 0; i < size; i++) {
    #pragma HLS loop_tripcount min=1 max=16 avg=8
    #pragma HLS PIPELINE
    #pragma HLS UNROLL factor=16           // independent: complete unroll
    buffer[i] = expf(x[i] - max_val);
  }
  float sum = 0.0f;
sum:
  for (int i = 0; i < size; i++) {
    #pragma HLS loop_tripcount min=1 max=16 avg=8
    sum += buffer[i];                      // carried dep: no PIPELINE
  }
  const float inv_sum = 1.0f / sum;        // one division hoisted out
norm:
  for (int i = 0; i < size; i++) {
    #pragma HLS loop_tripcount min=1 max=16 avg=8
    #pragma HLS PIPELINE
    #pragma HLS UNROLL factor=16           // independent: complete unroll
    x[i] = buffer[i] * inv_sum;
  }
}
\end{lstlisting}

\subsection{C-Simulation Results}
\label{subsec:softmax_csim}

C-simulation was performed for \texttt{softmax\_top}. The testbench
supplied a 16-element input vector with \texttt{size = 8} active
elements, exercising the variable-length behaviour of the
\texttt{loop\_tripcount} pragmas. The input vector
$[-1.2, 0.5, 1.8, 4.5, -0.3, 2.1, -2.7, 0.9]$ was chosen to include
one clearly dominant value (index~3, score $4.5$) so that the
resulting probability distribution should concentrate most of its mass
on that position. The testbench verified two independent correctness
criteria, namely element-wise agreement with a software reference
implementing the same four-loop algorithm, and a sum-to-unity check
on the hardware output independent of the reference. Table~\ref{tab:softmax_csim}
shows the output from both the kernel and the software reference.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & HW output & SW reference & Absolute error \\
    \midrule
    0 & 0.002752 & 0.002752 & 0 \\
    1 & 0.015064 & 0.015064 & 0 \\
    2 & 0.055273 & 0.055273 & 0 \\
    3 & 0.822446 & 0.822446 & 0 \\
    4 & 0.006769 & 0.006769 & 0 \\
    5 & 0.074611 & 0.074611 & 0 \\
    6 & 0.000614 & 0.000614 & 0 \\
    7 & 0.022472 & 0.022472 & 0 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for \texttt{softmax} with input
           $[-1.2, 0.5, 1.8, 4.5, -0.3, 2.1, -2.7, 0.9]$,
           $\texttt{size} = 8$, $\texttt{MAXSIZE} = 16$.
           All eight elements match the software reference,
           and the sum of outputs is $1.00000000$.}
  \label{tab:softmax_csim}
\end{table}

All eight elements match the software reference with zero absolute
error, and the independent sum check confirmed that the outputs sum
to $1.00000000$. The output is strongly
concentrated at index~3 (probability $0.822$), consistent with the
dominant score. The output at index~6 is the smallest ($0.000614$),
corresponding to the most negative input score of $-2.7$. Both extremes
are handled correctly, confirming that the max-subtract numerical
stabilisation and the \texttt{inv\_sum} reciprocal optimisation produce
accurate results. The csim run completed with zero errors as reported
by Vitis HLS.

\subsection{HLS Synthesis Results}
\label{subsec:softmax_csyn}

The \texttt{softmax} function was synthesised in isolation using the
prototype configuration (\texttt{MAXSIZE} = \texttt{seq\_len} = 16)
on the \texttt{xczu9eg-ffvb1156-2-e} device at a
\SI{150}{\mega\hertz} target clock (\SI{6.667}{\nano\second} period).
Tables~\ref{tab:softmax_timing} and \ref{tab:softmax_top_latency}
present the top-level timing and latency summary, followed by
Table~\ref{tab:softmax_resources} for resource utilisation and
Table~\ref{tab:softmax_latency} for the per-loop latency breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock   & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667       & 4.867          & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{softmax\_top} at the
           \SI{150}{\mega\hertz} (\SI{6.667}{\nano\second}) target clock.}
  \label{tab:softmax_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    51 & 205 & \SI{0.340}{\micro\second} & \SI{1.367}{\micro\second} & 52 & 206 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{softmax\_top} at
           \SI{150}{\mega\hertz}. The minimum corresponds to
           $\texttt{size} = 1$ and the maximum to
           $\texttt{size} = \texttt{MAXSIZE} = 16$.}
  \label{tab:softmax_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 39{,}921 & 38{,}979 & 274{,}080 & 14 \\
    FF   & 19{,}165 & 18{,}432 & 548{,}160 &  3 \\
    BRAM & 4        & 0        &   1{,}824 & $<$1 \\
    DSP  & 192      & 192      &   2{,}520 &  7 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{softmax} synthesised at
           $\texttt{MAXSIZE} = 16$ on the ZCU102
           (\texttt{xczu9eg-ffvb1156-2-e}). The core-only column
           excludes the AXI master adapter and S-AXILite control
           interface.}
  \label{tab:softmax_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Loop & Trip count & II & Iteration latency & Total (cycles) \\
    \midrule
    \texttt{max}  & 16 &  1 &  2 &  16 \\
    \texttt{exp}  &  1 &  1 &  9 &   9 \\
    \texttt{sum}  & 16 &  6 &  8 &  97 \\
    \texttt{norm} &  1 &  1 &  1 &   1 \\
    \bottomrule
  \end{tabular}
  \caption{Per-loop latency breakdown of \texttt{softmax}
           ($\texttt{MAXSIZE} = 16$) at \SI{150}{\mega\hertz}.
           The \texttt{exp} and \texttt{norm} loops report trip
           count~1 because \texttt{UNROLL factor=16} on a
           16-element loop collapses all iterations into a single
           parallel pipeline stage.}
  \label{tab:softmax_latency}
\end{table}

The resource consumption stands out in these results. With 192 DSPs and 38{,}979 core LUTs,
\texttt{softmax} uses more hardware than all four preceding
primitives combined. The cause is the complete unrolling of the
\texttt{exp} and \texttt{norm} loops into 16 parallel datapath lanes
each. Each of the 16 \texttt{expf} computations in the \texttt{exp}
loop is mapped to a Xilinx \texttt{fexp\_32ns\_32ns\_32\_9\_full\_dsp}
IP core using 7 DSP48E2 blocks and an 8-cycle pipeline depth at
\SI{150}{\mega\hertz}, adding $16 \times 7 = 112$ DSPs. The 16
max-subtract operations in the same loop each require an \texttt{fsub}
unit using 2 DSPs, adding a further $16 \times 2 = 32$ DSPs.
The 16 multiply-by-reciprocal operations in the \texttt{norm} loop
each map to an \texttt{fmul} unit using 3 DSPs, adding $16 \times
3 = 48$ DSPs. The total $112 + 32 + 48 = 192$ DSPs is accounted for
by these three sources, with the \texttt{max} and \texttt{sum}
loops using zero DSPs since they use only fabric-based
comparators and adders respectively.

The \texttt{Bind Storage} section of the synthesis report confirms that
the \texttt{buffer[16]} array occupies no BRAM. At $16 \times 4 = 64$
bytes, the buffer falls well below the threshold at which HLS infers
block RAM, and it is instead mapped to flip-flop registers within the
pipeline datapath.

The variable latency reported in Table~\ref{tab:softmax_top_latency}
is the first instance of a non-constant latency across all primitives
synthesised in this chapter. The minimum of 51 cycles corresponds to
$\texttt{size} = 1$, where the \texttt{max} loop
has zero iterations (its lower bound is $i = 1$), the \texttt{exp}
loop computes a single element, the \texttt{sum} loop performs one
accumulation, and the \texttt{norm} loop normalises one value. The
maximum of 205 cycles corresponds to $\texttt{size} = 16 =
\texttt{MAXSIZE}$, running all four loops at their full trip counts.
This variability is a direct consequence of the runtime \texttt{size}
parameter, which is passed through the S-AXILite control register
interface and therefore cannot be statically resolved by HLS during
synthesis. The \texttt{loop\_tripcount} hints on all four loops inform
the latency estimator of the expected range but do not constrain the
hardware. The actual execution length is always determined at runtime.

The \texttt{sum} loop is the dominant contributor to worst-case
latency at 97 cycles, nearly five times the combined contribution of
the other three loops. The cause is the loop-carried dependency on the
floating-point accumulator \texttt{sum}, as each iteration must wait for
the preceding \texttt{fadd} \texttt{faddfsub\_32ns\_32ns\_32\_6\_full\_dsp\_1}
to complete before it can add the next
element. The \texttt{fadd} unit has a scheduled pipeline depth of 5 cycles at
\SI{150}{\mega\hertz}, and the additional cycle required to write the
updated accumulator value back and make it available for the next
read gives an effective II of 6. Across 16 iterations this gives $6 \times 16 - 6 + 8 =
98 - 1 = 97$ cycles for the loop body alone. This is the same
loop-carried dependency pattern observed in the
\texttt{sum\_of\_squares} reduction in \texttt{rmsnorm}. The
\texttt{rmsnorm} implementation partially mitigated this with
\texttt{UNROLL factor=4} to maintain four independent partial sum
lanes. Applying the same technique to \texttt{sum} would, with four
independent lanes and a 5-cycle \texttt{fadd} pipeline depth, reduce the
II from 6 to approximately $\lceil 5/4 \rceil = 2$, saving roughly 64 cycles per softmax
call and is identified as a future optimisation.

The timing result follows the same pattern as all preceding sections.
The estimated path of \SI{4.867}{\nano\second} plus the
\SI{1.800}{\nano\second} uncertainty exhausts the
\SI{6.667}{\nano\second} target, leaving zero slack. The \texttt{exp}
pipeline sub-module is the most timing-critical path, with only
\SI{0.09}{\nano\second} of slack above the timing constraint prior to
uncertainty, due to the deep combinational chain through the 7-DSP
\texttt{fexp} core.

\subsection{Iteration 3: Sum Loop Unroll}
\label{subsec:softmax_iter3}

The third kernel iteration (\texttt{forward\_fyp\_kvint8.cpp}) introduces
two pragma modifications to the \texttt{softmax} function while leaving
the function body logic unchanged. The first addition is
\texttt{\#pragma HLS ARRAY\_PARTITION variable=buffer cyclic factor=16}
on the internal \texttt{buffer[MAXSIZE]} array, which was absent in both
earlier iterations. The second addition is \texttt{\#pragma HLS UNROLL
factor=16} on the \texttt{sum} reduction loop. These two changes are
directly motivated by the II\,=\,6 violation identified in the
iteration~1 synthesis analysis, which attributed the 97-cycle \texttt{sum}
latency to a loop-carried dependency on the floating-point accumulator.
The array partition is a prerequisite for the unroll. Without 16
independent memory banks on \texttt{buffer}, adding 16 parallel read
accesses from an unpartitioned array would cause HLS to serialise the
reads rather than parallelise them.

\subsubsection{C-Simulation Results}
\label{subsubsec:softmax_iter3_csim}

Because the function body is identical to iteration~1, the C-simulation
produces bit-identical outputs, with all eight active elements matching the
software reference with zero absolute error, and the sum-to-unity check
passes with a reported sum of $1.00000000$. The csim run
completed with zero errors.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:softmax_iter3_csyn}

Tables~\ref{tab:softmax_iter3_timing} and \ref{tab:softmax_iter3_latency}
present the timing summary and per-loop latency breakdown for the
iteration~3 softmax synthesis. Table~\ref{tab:softmax_iter3_resources}
compares the resource utilisation against iteration~1 for direct comparison.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock   & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667       & 5.084          & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{softmax\_kvint8\_top} at the
           \SI{150}{\mega\hertz} target clock. The estimated path of
           \SI{5.084}{\nano\second} plus the \SI{1.800}{\nano\second}
           uncertainty exceeds the \SI{6.667}{\nano\second} budget by
           \SI{0.217}{\nano\second}, constituting a timing violation.}
  \label{tab:softmax_iter3_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Loop & Trip count & II & Iter.\ latency & Total (cycles) \\
    \midrule
    \texttt{max}  & 16 &  1 &  2 &  16 \\
    \texttt{exp}  &  1 &  1 &  9 &   9 \\
    \texttt{sum}  &  1 & 96 & 97 & 185 \\
    \texttt{norm} &  1 &  1 &  1 &   1 \\
    \bottomrule
  \end{tabular}
  \caption{Per-loop latency breakdown of the iteration~3 \texttt{softmax}
           ($\texttt{MAXSIZE} = 16$) at \SI{150}{\mega\hertz}. The
           \texttt{sum} loop collapses to a single effective iteration
           (trip count~1) due to the \texttt{UNROLL factor=16}, but its
           total latency increases from 97 to 185 cycles compared to
           iteration~1.}
  \label{tab:softmax_iter3_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    & \multicolumn{2}{c}{Iteration 1 (baseline)} &
      \multicolumn{2}{c}{Iteration 3 (sum unroll)} \\
    Resource & Total & Core & Total & Core \\
    \midrule
    LUT  & 39{,}921 & 38{,}979 & 42{,}723 & 41{,}781 \\
    FF   & 19{,}165 & 18{,}432 & 20{,}290 & 19{,}557 \\
    BRAM & 4        & 0        & 4        & 0 \\
    DSP  & 192      & 192      & 192      & 192 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation comparison between the iteration~1 baseline
           and the iteration~3 \texttt{softmax} designs, both at
           $\texttt{MAXSIZE} = 16$ on the \texttt{xczu9eg-ffvb1156-2-e}.
           Core figures exclude the AXI adapter and S-AXILite control
           interface.}
  \label{tab:softmax_iter3_resources}
\end{table}

The synthesis results reveal that the \texttt{UNROLL factor=16}
optimisation produced the opposite of the intended effect. In
iteration~1, the \texttt{sum} loop had trip count~16, achieved II\,=\,6,
and a total latency of 97 cycles. In iteration~3, the complete unroll
collapses the loop to a single effective iteration (trip count~1), but
that single iteration has a latency of 97 cycles and an II of 96,
giving a total of 185 cycles, nearly double the iteration~1 cost.
The worst-case total function latency increased from 205 to 295 cycles
as a result.

Fully unrolling the 16-element loop creates 16 independent
\texttt{fadd} units, which HLS must then reduce down to a single
scalar through a tree of $\log_2(16) = 4$ adder levels. With each
level having a 5-cycle pipeline depth, the minimum reduction latency
is approximately $4 \times 5 = 20$ cycles. The resulting network is
wide enough that routing introduces additional path delay, causing the
\SI{0.217}{\nano\second} timing violation visible in
Table~\ref{tab:softmax_iter3_timing}. This is the first timing
violation across all primitives synthesised in this chapter.

The correct approach to reducing the accumulator dependency would have
been to use a small number of independent partial accumulators, for
example two or four, rather than sixteen. Maintaining four independent
partial sum registers and merging them at the end would reduce the
effective II from 6 to approximately 2 while keeping the reduction tree
shallow enough to close timing. The lesson from this result is that
loop unrolling for reduction operations requires careful sizing, as the
number of parallel lanes must be matched to the adder pipeline depth,
not simply maximised, and the resulting hardware cost and routing
complexity must be validated through synthesis rather than assumed to
be beneficial.

% =============================================================================
\subsection{Iteration 4: I-BERT Integer Softmax}
\label{subsec:softmax_ibert}
% =============================================================================

The timing violation in iteration~3 motivated a different approach to the
reduction problem. Rather than tuning the unroll factor on the floating-point
accumulator, iteration~4 replaces the floating-point softmax entirely with an
integer-arithmetic variant, \texttt{softmax\_ibert}, which avoids both the
reduction dependency and the seven-DSP \texttt{fexp} IP core.

\subsubsection{Implementation}
\label{subsubsec:softmax_ibert_impl}

The function signature of \texttt{softmax\_ibert<MAX\_SIZE>} differs from the
legacy \texttt{softmax<MAXSIZE>} in two ways. The input is \texttt{int32\_t
*logits} rather than \texttt{float *x}, since scores are kept as integers
throughout the attention pipeline in iteration~4. The output is a separate
\texttt{int8\_t *probs} buffer rather than an in-place float array, with
probabilities encoded in $[0, 127]$ where $127$ represents $1.0$. A
\texttt{float scale\_factor} parameter carries the remaining scale factors
deferred from the \texttt{iterate} loop, specifically the product of the
per-head Q scale and the attention scaling factor $1/\sqrt{\texttt{head\_size}}$.

The exponential function is evaluated through a 128-entry lookup table
\texttt{exp\_lut}, defined as a \texttt{static const int8\_t} array in the
header. Each entry stores $\text{round}(127 \times e^{-\text{index}})$, so
the table encodes $127, 47, 17, 6, 2, 1, 0, \ldots$ for indices $0$ through
$127$. The function proceeds in three phases. First, the maximum of the
\texttt{logits[0..size-1]} values is found using an \texttt{UNROLL}ed loop
over \texttt{MAX\_SIZE} elements with an active-element guard. Second, each
shifted score is scaled by \texttt{scale\_factor}, truncated to an integer,
clamped to $[-127, 0]$, and mapped through the LUT to produce an \texttt{int8}
exponential result stored in a fully partitioned local array
\texttt{exp\_results[MAX\_SIZE]}. Third, the results are normalised using the
formula $p_i = (e_i \times 127) / \text{sum\_exp}$, where the multiplication
precedes the integer division to preserve precision.

All three loops have \texttt{\#pragma HLS UNROLL} with no factor specified,
giving complete unrolling over \texttt{MAX\_SIZE = 16} elements. Unlike the
reduction in the legacy \texttt{sum} loop, the accumulation of \texttt{sum\_exp}
here operates on \texttt{int8} values rather than \texttt{float} values. An
\texttt{int8} addition has no multi-cycle pipeline latency in the DSP48E2 path,
so the loop-carried dependency resolves within a single clock cycle and II\,=\,1
is achievable. No \texttt{ARRAY\_PARTITION} is applied to \texttt{exp\_lut}
itself in the current implementation. The commented-out directive in the source
reflects a future optimisation opportunity.

\subsubsection{C-Simulation Results}
\label{subsubsec:softmax_ibert_csim}

The testbench supplied eight \texttt{int32} logit values
$\{10, -5, 18, 25, -3, 21, -2, 9\}$ with \texttt{scale\_factor\,=\,0.2f} and
\texttt{size\,=\,8}. With \texttt{max\_val\,=\,25}, the shifted and scaled
values map to LUT indices as follows. Indices 3 and 5 have shifted scores of 0
and $-0.8$ respectively, both truncating to index 0, giving
\texttt{exp\_lut[0]\,=\,127}. Index 2 maps to $-1.4$, truncating to index 1,
giving $47$. Index 0 and index 7 both map to index 3, giving $6$. Indices 4
and 6 map to index 5, giving $1$. Index 1 maps to index 6, giving $0$.
The \texttt{sum\_exp} is therefore $127+47+6+127+1+1+0+6 = 315$.

Table~\ref{tab:softmax_ibert_csim} shows the output probabilities from both the
kernel and the software reference.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & HW prob & SW prob & Abs error \\
    \midrule
    0 &  2 &  2 & 0 \\
    1 &  0 &  0 & 0 \\
    2 & 18 & 18 & 0 \\
    3 & 51 & 51 & 0 \\
    4 &  0 &  0 & 0 \\
    5 & 51 & 51 & 0 \\
    6 &  0 &  0 & 0 \\
    7 &  2 &  2 & 0 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for \texttt{softmax\_ibert} with
           \texttt{logits} $= \{10, -5, 18, 25, -3, 21, -2, 9\}$,
           \texttt{scale\_factor\,=\,0.2f}, and \texttt{size\,=\,8}.
           All eight output probabilities match the software reference.
           The sum of active probabilities is $124$, falling short of
           $127$ due to integer division in the normalisation step.}
  \label{tab:softmax_ibert_csim}
\end{table}

All eight probabilities match the software reference with zero absolute error.
The sum of $124$ rather than $127$ is expected. Integer division in the
formula $p_i = (e_i \times 127) / \text{sum\_exp}$ truncates the fractional
part of each quotient, so the total probability mass is not conserved to the
last integer unit. This is a known property of integer-domain normalisation
and not an error in the implementation. The csim run completed with zero
errors.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:softmax_ibert_csyn}

Tables~\ref{tab:softmax_ibert_timing} and \ref{tab:softmax_ibert_top_latency}
present the timing and top-level latency summary. Table~\ref{tab:softmax_ibert_resources}
gives the resource utilisation and Table~\ref{tab:softmax_ibert_latency} the
per-phase latency breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{softmax\_ibert\_top} at the
           \SI{150}{\mega\hertz} target clock.}
  \label{tab:softmax_ibert_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    83 & 531 & \SI{0.553}{\micro\second} & \SI{3.540}{\micro\second} & 84 & 532 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{softmax\_ibert\_top} at
           \SI{150}{\mega\hertz}.}
  \label{tab:softmax_ibert_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 12{,}706 & 8{,}588 & 274{,}080 & 4 \\
    FF   &  8{,}254 & 5{,}171 & 548{,}160 & $<$1 \\
    BRAM &        4 &       0 &   1{,}824 & $<$1 \\
    DSP  &        3 &       3 &   2{,}520 & $<$1 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{softmax\_ibert} synthesised at
           $\texttt{MAX\_SIZE} = 16$ on the ZCU102.}
  \label{tab:softmax_ibert_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Module & Latency min (cycles) & Latency max (cycles) \\
    \midrule
    \texttt{softmax\_ibert\_16\_s} (core) & 37 & 485 \\
    AXI wrapper overhead                  & 46 &  46 \\
    \texttt{softmax\_ibert\_top} (total)  & 83 & 531 \\
    \bottomrule
  \end{tabular}
  \caption{Latency breakdown of \texttt{softmax\_ibert} ($\texttt{MAX\_SIZE} = 16$)
           at \SI{150}{\mega\hertz}. All loops are fully unrolled so HLS reports no
           per-loop trip count, and the breakdown is therefore given at the submodule
           level. AXI wrapper overhead covers the input and output burst transfers.}
  \label{tab:softmax_ibert_latency}
\end{table}

The defining result of this synthesis is the DSP count of 3, all from the
\texttt{softmax\_ibert} core. Iterations~1 and~3 each used 192 DSPs, entirely
dominated by the 16 parallel 	exttt{fexp} IP core instances. Replacing the floating-point exponential with the 128-entry
\texttt{exp\_lut} ROM eliminates all of those instances. The 3 DSPs that remain
correspond to the single \texttt{fmul} unit used to compute \texttt{val\_scaled\_f
= (float)val\_shifted * scale\_factor}, the one floating-point multiplication
that cannot be avoided since \texttt{scale\_factor} is a runtime float argument.
All other arithmetic in \texttt{softmax\_ibert} is performed in integer or
fixed-point, mapping entirely to LUT fabric.

The LUT count of 8{,}588 core LUTs is a 78\% reduction from iteration~1
(38{,}979 core LUTs). The integer operations, including the max-search
comparators, the LUT indexing arithmetic (which the Expression table shows
as shift-left and add operations), the integer accumulation of
\texttt{sum\_exp}, and the integer division in the normalisation step, all
map to LUT-based logic rather than DSP-backed floating-point units.

The variable latency of 83 to 531 cycles persists because the runtime
\texttt{size} parameter still governs how many elements are active. Since all
loops are fully unrolled over \texttt{MAX\_SIZE = 16}, HLS instantiates the
full 16-element datapath unconditionally but guards inactive elements with the
\texttt{if (i < size)} check, which adds multiplexer logic around each element.
The minimum of 83 cycles corresponds to \texttt{size\,=\,1} and the maximum of
531 cycles to \texttt{size\,=\,16}. The timing meets the
\SI{6.667}{\nano\second} constraint, with an estimated path of
\SI{4.867}{\nano\second} plus \SI{1.800}{\nano\second} uncertainty exhausting
the budget with zero slack, consistent with all preceding primitives.

% =============================================================================
\subsection{Iteration 6: Full TinyLlama Scale}
\label{subsec:softmax_ibert_update3}
% =============================================================================

\subsubsection{Implementation}
\label{subsubsec:softmax_ibert_update3_impl}

The sixth iteration ports \texttt{softmax\_ibert} to
\texttt{MAX\_SIZE\,=\,seq\_len\,=\,2048}, shown in
Listing~\ref{lst:softmax_ibert_update3}. Full \texttt{UNROLL} on the
three loops at this scale would instantiate 2{,}048 parallel comparators
and adder trees simultaneously, exceeding the ZCU102's LUT budget. All
three loops (\texttt{find\_max}, \texttt{compute\_exp}, \texttt{normalize})
are changed to \texttt{\#pragma HLS PIPELINE II\,=\,1} with a
\texttt{loop\_tripcount} hint, processing one element per cycle. The
\texttt{exp\_results} buffer partition changes from \texttt{complete}
to \texttt{cyclic factor=16}, as at 2{,}048 entries a complete partition
would require 16\,KB of flip-flop registers. A reciprocal trick is applied
in the normalisation step, computing \texttt{inv\_sum = 1.0f\,/\,(float)sum\_exp}
once before the \texttt{normalize} loop and multiplying per element,
removing the per-element integer division.

\begin{lstlisting}[
  caption = {\texttt{softmax\_ibert<MAX\_SIZE>} from \texttt{FFN\_update3.h}
             at full scale (\texttt{MAX\_SIZE\,=\,2048}). All three loops
             use \texttt{PIPELINE II=1} with \texttt{loop\_tripcount}
             hints. The reciprocal trick replaces per-element integer
             division in the normalisation step.},
  label   = {lst:softmax_ibert_update3}
]
template <int MAX_SIZE>
void softmax_ibert(int32_t *logits, int8_t *probs,
                   int size, float scale_factor) {
    if (size > MAX_SIZE) size = MAX_SIZE;

    int32_t max_val = -2147483648;
find_max:
    for (int i = 0; i < MAX_SIZE; i++) {
        #pragma HLS PIPELINE II=1
        #pragma HLS loop_tripcount min=1 max=MAX_SIZE
        if (i < size) {
            if (logits[i] > max_val) max_val = logits[i];
        }
    }

    int32_t sum_exp = 0;
    int8_t exp_results[MAX_SIZE];
    #pragma HLS ARRAY_PARTITION variable=exp_results cyclic factor=16

compute_exp:
    for (int i = 0; i < MAX_SIZE; i++) {
        #pragma HLS PIPELINE II=1
        #pragma HLS loop_tripcount min=1 max=MAX_SIZE
        if (i < size) {
            int32_t val_shifted = logits[i] - max_val;
            float   val_scaled_f = (float)val_shifted * scale_factor;
            int32_t val_scaled   = (int32_t)val_scaled_f;
            if (val_scaled < -127) val_scaled = -127;
            int8_t e = exp_lut[-val_scaled];
            exp_results[i] = e;
            sum_exp += e;
        }
    }

    if (sum_exp == 0) sum_exp = 1;
    float inv_sum = 1.0f / (float)sum_exp;

normalize:
    for (int i = 0; i < MAX_SIZE; i++) {
        #pragma HLS PIPELINE II=1
        #pragma HLS loop_tripcount min=1 max=MAX_SIZE
        if (i < size) {
            int32_t p = (int32_t)((float)exp_results[i] * 127.0f * inv_sum);
            if (p > 127) p = 127;
            if (p < 0)   p = 0;
            probs[i] = (int8_t)p;
        }
    }
}
\end{lstlisting}

\subsubsection{C-Simulation Results}
\label{subsubsec:softmax_ibert_update3_csim}

The testbench used the same inputs as the iteration~4 csim for direct
comparison: \texttt{logits}\,$=\{10, -5, 18, 25, -3, 21, -2, 9\}$,
\texttt{scale\_factor\,=\,0.2f}, \texttt{size\,=\,8}, against
\texttt{MAX\_SIZE\,=\,2048}. All eight output probabilities matched the
software reference with zero error, and the sum of output probabilities
was 124, identical to iteration~4. The csim run completed with zero errors.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:softmax_ibert_update3_csyn}

Tables~\ref{tab:softmax_ibert_u3_timing} and
\ref{tab:softmax_ibert_u3_top_latency} present the timing and latency
summary. Table~\ref{tab:softmax_ibert_u3_resources} gives the resource
utilisation and Table~\ref{tab:softmax_ibert_u3_latency} the per-stage
breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{softmax\_ibert\_update3\_top} at the
           \SI{150}{\mega\hertz} target clock.}
  \label{tab:softmax_ibert_u3_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    10{,}320 & 10{,}320 & \SI{68.803}{\micro\second} & \SI{68.803}{\micro\second} & 10{,}321 & 10{,}321 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{softmax\_ibert\_update3\_top}
           at \SI{150}{\mega\hertz}.}
  \label{tab:softmax_ibert_u3_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 7{,}137 & 6{,}557 & 274{,}080 & 2 \\
    FF   & 2{,}926 & 2{,}414 & 548{,}160 & $<$1 \\
    BRAM &      20 &      16 &   1{,}824 & 1 \\
    DSP  &       6 &       6 &   2{,}520 & $<$1 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{softmax\_ibert} at full scale
           ($\texttt{MAX\_SIZE} = 2048$) on the ZCU102. Core-only figures
           exclude the AXI master adapter (\texttt{gmem0\_m\_axi}).}
  \label{tab:softmax_ibert_u3_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Stage & Module & Latency (cycles) \\
    \midrule
    Load \texttt{logits\_in}                & \texttt{VITIS\_LOOP\_36\_1}      & 2{,}051 \\
    \texttt{softmax\_ibert\_2048\_s} core   & \texttt{softmax\_ibert\_2048\_s} & 6{,}200 \\
    Write \texttt{probs\_out}               & \texttt{VITIS\_LOOP\_43\_2}      & 2{,}056 \\
    \bottomrule
  \end{tabular}
  \caption{Per-stage latency breakdown of \texttt{softmax\_ibert} at full
           scale (\texttt{MAX\_SIZE} = 2048) at \SI{150}{\mega\hertz}.
           Total function latency is 10{,}320 cycles
           (\SI{68.803}{\micro\second}).}
  \label{tab:softmax_ibert_u3_latency}
\end{table}

Timing meets the \SI{6.667}{\nano\second} constraint with zero slack.

The total latency of 10{,}320 cycles has identical minimum and maximum,
in contrast to iteration~4 where latency ranged from 83 to 531 cycles
depending on the runtime \texttt{size} argument. With \texttt{PIPELINE}
and fixed loop bounds, all three loops always execute for the full
\texttt{MAX\_SIZE\,=\,2048} iterations, making the latency independent of
the input size. This is a deliberate trade-off, accepting a fixed
worst-case latency at all sequence positions in exchange for a
synthesisable, resource-efficient implementation at full scale.

The \texttt{softmax\_ibert\_2048\_s} core accounts for 6{,}200 of the
10{,}320 total cycles. The three sequential passes each iterate over
2{,}048 elements at II\,=\,1, giving approximately $3 \times 2048 = 6{,}144$
cycles for the core loops, with the remaining 56 cycles from control
overhead between passes. The AXI load and write stages each cost
approximately 2{,}050 cycles for the 2{,}048-element transfers.

The 16 core BRAMs come entirely from the \texttt{exp\_results[2048]}
buffer. At \texttt{MAX\_SIZE\,=\,16} in iteration~4, the 16-entry array
fitted in register fabric. At \texttt{MAX\_SIZE\,=\,2048}, the array
holds $2048 \times 8 = 16{,}384$ bits, and with cyclic factor\,=\,16
partitioning each bank holds 128 entries of 8-bit data. At this width
and depth, HLS infers BRAM rather than registers, giving 16 BRAM banks.
The \texttt{local\_logits[2048]} input buffer of int32 data maps to 16
BRAM banks under the same analysis but is housed in the AXI wrapper
rather than the core module. These 4 BRAMs appear in the total count
but not in the core count. The \texttt{exp\_results} array holds only
7 effective bits per entry, so HLS maps it to narrow-width BRAMs rather
than 32-bit BRAMs, reducing BRAM consumption relative to a float buffer
of the same length.

The 6 DSPs come from two \texttt{fmul} units in the \texttt{normalize}
loop. The expression \texttt{(float)exp\_results[i] * 127.0f * inv\_sum}
requires one \texttt{fmul} for the multiplication by 127 and one for the
multiplication by \texttt{inv\_sum}, each consuming 3 DSPs. With
\texttt{PIPELINE II\,=\,1}, both multiplications must be available every
cycle, so two independent \texttt{fmul} instances are required.

Comparing against iteration~4 (\texttt{MAX\_SIZE\,=\,16}, \texttt{UNROLL
factor=16}), the LUT count falls from 8{,}588 to 6{,}557 and the FF
count falls from 5{,}171 to 2{,}414, despite the 128-fold increase in
array size. At \texttt{MAX\_SIZE\,=\,16} with full unrolling, HLS
instantiated 16 parallel comparator chains, 16 LUT-addressed
exponential lookups, and 16 adder lanes simultaneously. At
\texttt{MAX\_SIZE\,=\,2048} with \texttt{PIPELINE}, a single
comparator and a single LUT lookup handle all elements sequentially.
The BRAM count rises from 0 to 16 since \texttt{exp\_results} crossed
the BRAM inference threshold, and the DSP count rises from 3 to 6 due
to the two simultaneous \texttt{fmul} units in the pipelined
normalisation stage.



Section~\ref{sec:attn_impl} covers the attention mechanism across three
kernel iterations. The baseline implementation uses floating-point
arithmetic throughout and is synthesised from \texttt{forward.cpp}. The
second iteration replaces the inner dot products and value accumulation
with per-head INT8 arithmetic and is synthesised from
\texttt{forward\_fyp.cpp}. The third iteration extends the INT8
representation into the KV cache itself, eliminating on-the-fly
quantisation from the inner attention loops and is synthesised from
\texttt{forward\_fyp\_kvint8.cpp}. Each iteration has its own
implementation description, and C-simulation and synthesis results for
all three iterations are compared together at the end of the section.

% =============================================================================
\section{Grouped-Query Attention}
\label{sec:attn_impl}
% =============================================================================

\subsection{Baseline: Floating-Point Attention}
\label{subsec:attn_fp32}

\subsubsection{Implementation}
\label{subsubsec:attn_fp32_impl}

The baseline attention block is contained in the
\texttt{multihead\_attention} loop in \texttt{forward.cpp}. It executes
after the QKV matrix--vector multiplications and before the residual
addition, operating on the float query vector \texttt{q[dim]} and the
float KV cache arrays \texttt{key\_cache} and \texttt{value\_cache}.
The block is not a standalone function in the baseline, as it is embedded
directly inside the \texttt{forward} kernel's layer loop, and the KV
cache arrays are shared on the same \texttt{gmem0} AXI bundle as the
transformer weight struct.

The outer \texttt{multihead\_attention} loop iterates over the
\texttt{n\_heads = 2} query heads sequentially at prototype scale. For
each head, the computation proceeds in three phases. The first phase,
the \texttt{iterate} loop, computes the dot product between the current
query slice and every cached key vector from timestep 0 through
\texttt{pos}. The outer position loop has \texttt{\#pragma HLS
PIPELINE}, and the inner element loop over \texttt{head\_size = 4}
dimensions has \texttt{\#pragma HLS UNROLL} with no factor
specified, giving complete unrolling. The score is then divided
by $\sqrt{\texttt{head\_size}}$ and written to the local
\texttt{att\_buf} score buffer. After the \texttt{iterate} loop, a call
to \texttt{softmax<seq\_len>} converts the scores for the current head
into a probability distribution in-place. The second phase, the
\texttt{acc} loop, accumulates the probability-weighted value vectors
into \texttt{xb\_buf}. Like the \texttt{iterate} loop, it pipelines the
outer position loop and completely unrolls the inner element loop.

The local buffers \texttt{q\_buf}, \texttt{att\_buf}, and
\texttt{xb\_buf} are each partitioned with \texttt{ARRAY\_PARTITION
cyclic factor=16} to provide sufficient memory ports for the
simultaneously executing unrolled inner loop iterations. The key cache
and value cache are read directly from the \texttt{gmem0} AXI port
inside the pipelined outer loops without intermediate buffering, which
is the source of the II violation described in the synthesis analysis
below.

An important consideration concerns the \texttt{PIPELINE} pragma
placement on the \texttt{iterate} and \texttt{acc} outer loops. At
prototype scale (\texttt{seq\_len = 16}, \texttt{head\_size = 4}),
pipelining the outer position loop is safe because the inner unrolled element
loop generates four concurrent AXI read requests per outer iteration,
which the adapter's transaction queue can accommodate without
stalling indefinitely. At full TinyLlama scale (\texttt{seq\_len =
2048}, \texttt{head\_size = 64}), applying \texttt{PIPELINE II=1} to
the outer loop would instruct HLS to start a new position every clock
cycle while the inner loop generates 64 simultaneous AXI read requests.
The adapter's queue would fill within the first iteration, and each
subsequent position would stall waiting for outstanding transactions to
clear, creating a circular wait that locks the kernel. This AXI
interface stall is one of the three issues encountered during
on-board testing and is discussed further in
\cref{subsec:forward_limitations}.
Listing~\ref{lst:attn_baseline} shows the \texttt{iterate} and
\texttt{acc} loop bodies extracted from \texttt{forward.cpp}.

\begin{lstlisting}[
  caption = {Baseline float attention \texttt{iterate} and \texttt{acc}
             loops from \texttt{forward.cpp}. Both outer loops carry
             \texttt{PIPELINE}. Both inner loops carry \texttt{UNROLL}
             (complete) over \texttt{head\_size\,=\,4} elements.
             The key and value cache values are read as float directly
             from DDR with no intermediate buffering.},
  label = {lst:attn_baseline}
]
iterate:
  for (int t = 0; t <= pos; t++) {
    #pragma HLS PIPELINE
    #pragma HLS loop_tripcount min=1 max=16 avg=8
    float score = 0.0f;
    for (int i = 0; i < head_size; i++) {
      #pragma HLS UNROLL         // complete unroll: 4 parallel fmul units
      score += q_buf[i + q_offset] * key_cache[i + key_offset];
    }
    score /= sqrtf(head_size);
    att_buf[t + att_offset] = score;
  }
  softmax<seq_len>(att_buf + att_offset, pos + 1);

acc:
  for (int t = 0; t <= pos; t++) {
    #pragma HLS loop_tripcount min=1 max=16 avg=8
    #pragma HLS PIPELINE
    float a = att_buf[t + att_offset];
    for (int i = 0; i < head_size; i++) {
      #pragma HLS UNROLL         // complete unroll: 4 parallel fmul+fadd
      xb_buf[i + xb_offset] += a * value_cache[i + v_offset]; // float DDR read
    }
  }
\end{lstlisting}

\subsubsection{C-Simulation Results}
\label{subsubsec:attn_fp32_csim}

C-simulation was performed for \texttt{attn\_baseline\_top}. The
testbench exercised the kernel at \texttt{pos\,=\,3} and
\texttt{loff\,=\,0}, covering four attended positions across both query
heads. The query vector was set to linearly spaced values $[0.1, 0.2,
\ldots, 0.8]$, and the key and value cache entries were initialised
with small deterministic float values to produce non-trivial softmax
outputs without requiring random number generation. The testbench
verified only the final weighted value accumulation output \texttt{xb},
since the attention score buffer \texttt{att\_buf} is an internal
on-chip buffer in the real kernel and is not exposed as an AXI port.
Table~\ref{tab:attn_fp32_csim} shows the output from both the kernel
and the software reference.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & HW \texttt{xb} & SW reference & Absolute error \\
    \midrule
    0 & 0.343754 & 0.343754 & 0 \\
    1 & 0.313754 & 0.313754 & 0 \\
    2 & 0.283754 & 0.283754 & 0 \\
    3 & 0.253754 & 0.253754 & 0 \\
    4 & 0.213827 & 0.213827 & 0 \\
    5 & 0.183827 & 0.183827 & 0 \\
    6 & 0.153827 & 0.153827 & 0 \\
    7 & 0.123827 & 0.123827 & 0 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for the baseline float attention kernel
           (\texttt{attn\_baseline\_top}) at \texttt{pos\,=\,3},
           \texttt{loff\,=\,0}. All eight elements of the weighted value
           accumulation \texttt{xb} match the software reference
           with zero absolute error.}
  \label{tab:attn_fp32_csim}
\end{table}

All eight output elements agree with the software reference,
confirming that the three phases of the attention computation, score
dot products, softmax normalisation, and value accumulation, all
produce correct results when composed together. The zero-error outcome
is expected for this test configuration, as all arithmetic is performed in
float32 using the same operation sequence in both the kernel and the
reference, so no rounding divergence can arise. The csim run completed
with zero errors as reported by Vitis HLS.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:attn_fp32_csyn}

The baseline attention block was synthesised in isolation on the
\texttt{xczu9eg-ffvb1156-2-e} device at a \SI{150}{\mega\hertz} target
clock. Tables~\ref{tab:attn_fp32_timing} and
\ref{tab:attn_fp32_top_latency} present the top-level timing and
latency summary, followed by Table~\ref{tab:attn_fp32_resources} for
resource utilisation and Table~\ref{tab:attn_fp32_latency} for the
per-pipeline-module breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock   & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667       & 4.867          & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{attn\_baseline\_top} at the
           \SI{150}{\mega\hertz} (\SI{6.667}{\nano\second}) target clock.}
  \label{tab:attn_fp32_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    48 & 800 & \SI{0.320}{\micro\second} & \SI{5.334}{\micro\second} & 49 & 801 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{attn\_baseline\_top}
           at \SI{150}{\mega\hertz}. The minimum corresponds to
           \texttt{pos\,=\,0} (one attended position) and the maximum to
           \texttt{pos\,=\,seq\_len\,$-$\,1\,=\,15} (sixteen attended positions).}
  \label{tab:attn_fp32_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 54{,}912 & 53{,}650 & 274{,}080 & 20 \\
    FF   & 27{,}584 & 26{,}673 & 548{,}160 &  5 \\
    BRAM & 4        & 0        &   1{,}824 & $<$1 \\
    DSP  & 192      & 192      &   2{,}520 &  7 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of the baseline float attention block
           synthesised at prototype scale on the ZCU102
           (\texttt{xczu9eg-ffvb1156-2-e}). The core-only column
           excludes the AXI master adapter and S-AXILite control
           interface.}
  \label{tab:attn_fp32_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Pipeline stage & Trip count & II & Iter.\ latency & Total (cycles) \\
    \midrule
    \texttt{iterate} (score dot product)  & 16 &  4 & 42 & 101 \\
    \texttt{max}     (softmax pass 1)     & 16 &  1 &  2 &  16 \\
    \texttt{exp}     (softmax pass 2)     &  1 &  1 & 17 &  16 \\
    \texttt{sum}     (softmax pass 3)     & 16 &  6 &  8 &  97 \\
    \texttt{norm}    (softmax pass 4)     &  1 &  1 &  1 &   1 \\
    \texttt{acc}     (value accumulation) & 16 &  6 & 23 & 112 \\
    \bottomrule
  \end{tabular}
  \caption{Per-pipeline-module latency breakdown of the baseline float
           attention block at \SI{150}{\mega\hertz}, for one head at
           maximum context length (\texttt{pos\,=\,15}). The outer
           \texttt{multihead\_attention} loop over \texttt{n\_heads\,=\,2}
           runs sequentially, so the per-head cost is incurred twice.}
  \label{tab:attn_fp32_latency}
\end{table}

The resource figures reflect the composition of the attention block
with the softmax function it instantiates internally. The 192 DSPs and 53{,}650
core LUTs match the softmax standalone synthesis results,
confirming that the floating-point exponential units from the 16 fully
unrolled \texttt{fexp} instances dominate the resource budget of the
entire attention block. The compute logic for the score dot products
and value accumulation uses relatively few resources.
The \texttt{iterate} pipeline module uses only 2{,}005 LUTs and zero
DSPs, and the \texttt{acc} module uses only 1{,}136 LUTs and zero DSPs.

The II violations in the \texttt{iterate} and
\texttt{acc} loops. The \texttt{iterate} loop achieves II\,=\,4 against
a target of 1. With \texttt{\#pragma HLS PIPELINE} on the outer
position loop and \texttt{\#pragma HLS UNROLL} (complete) on the inner
\texttt{head\_size\,=\,4} element loop, HLS attempts to start a new
position every clock cycle. Doing so requires loading four key cache
elements simultaneously through the shared \texttt{gmem0} AXI port at
each new iteration. The AXI adapter can accept at most as many
outstanding read transactions as it has queue slots, and with four
simultaneous requests the pipeline stalls until the preceding reads
complete, giving an effective II of 4. The
\texttt{iterate} module therefore takes $4 \times (16 - 1) + 42 = 101$
cycles to process all 16 positions rather than the ideal $16 + 42 =
58$ cycles at II\,=\,1.

The \texttt{acc} loop achieves II\,=\,6. The inner element loop is
again completely unrolled, but the accumulation
$\texttt{xb\_buf}[i] \mathrel{+}= a \times \texttt{value\_cache}[i +
\texttt{v\_offset}]$ introduces a float multiply-add chain through the
accumulator. The \texttt{fmul}--\texttt{fadd} dependency chain runs through
the 5-cycle \texttt{fadd} unit, and the additional write-back cycle before
the updated accumulator is available gives an effective II of 6, preventing the
outer loop from starting the next position until the previous
accumulation has propagated through both the multiplier and the adder.
This II\,=\,6 is identical to the accumulator dependency that was
identified in the \texttt{sum} loop of \texttt{softmax}. These II
violations at prototype scale, while not catastrophic given the small
\texttt{seq\_len = 16}, scale linearly at full TinyLlama dimensions
(\texttt{seq\_len = 2048}) and motivate the per-head INT8 redesign
described in \cref{subsec:attn_int8}.

\subsection{Refined: Per-Head INT8 Attention}
\label{subsec:attn_int8}

The second iteration of the attention mechanism, implemented in
\texttt{forward\_fyp.cpp}, replaces the floating-point score
computation and value accumulation with per-head INT8 arithmetic. The
motivation for this redesign is a quantisation precision problem that
arises when the global \texttt{quantize()} helper is applied to the
full query, key, and value vectors. The helper computes a single
scale factor from the absolute maximum of the entire \texttt{dim}-element
vector. If one head's slice contains activation outliers that dominate
the global maximum, the resulting scale factor maps the outlier range to
the INT8 boundary of 127 and compresses all other heads' values into a
narrow fraction of the INT8 range. This cross-head scale contamination
degrades the precision of every head that does not contain the outlier.
The fix is to compute a separate scale for each head's slice
independently, so that the scale reflects only the \texttt{head\_size}
elements belonging to that head.

The revised implementation introduces three structured steps per head,
executed in sequence within the \texttt{multihead\_attention} outer loop.

Step~1 quantises the query slice for the current head. Before the
position loop begins, the implementation scans the
\texttt{head\_size\,=\,4} elements of \texttt{q\_buf} belonging to head
\texttt{h} to find the maximum absolute value, computes a per-head scale
$s_Q = \texttt{max\_val} / 127$, and quantises the slice into a local
\texttt{int8\_t q\_int8[head\_size]} array. The reciprocal trick from the
quantise primitive is applied, whereby a single \texttt{fdiv} computes
$1 / s_Q$ before the element loop, and each element is then quantised
by multiplying by this reciprocal rather than dividing, replacing
\texttt{head\_size} divisions with one division and \texttt{head\_size}
multiplications.

Step~2 executes inside the \texttt{iterate} position loop. For each
attended timestep \texttt{t}, the \texttt{head\_size} key values are
loaded from the float KV cache into a local \texttt{k\_buffer}, a
per-position scale $s_K$ is computed from the buffer's maximum absolute
value, and the key slice is quantised to \texttt{int8\_t k\_int8}. The
dot product between \texttt{q\_int8} and \texttt{k\_int8} is then
accumulated into a 32-bit integer \texttt{score\_int}, with
\texttt{\#pragma HLS UNROLL} applied to the inner \texttt{head\_size}
element loop. The final dequantisation combines both per-head scales and
the attention scaling factor in a single expression:

\begin{equation}
  \text{score} = \left(\sum_{i=0}^{d_k-1} \hat{q}_i \cdot \hat{k}_i\right)
    \cdot s_Q \cdot s_K \cdot \frac{1}{\sqrt{d_k}}
  \label{eq:int8_score}
\end{equation}

\noindent where $\hat{q}_i$ and $\hat{k}_i$ are the INT8 quantised
query and key values, and $s_Q$, $s_K$ are the respective per-head
scales. The result is written to \texttt{att\_buf} as a float.

The softmax call is widened from \texttt{softmax<16>} to
\texttt{softmax<257>} in this iteration. This accommodates context
lengths up to 256 positions without any code change to the function
body. The cost is that the internal \texttt{buffer[257]} array within
\texttt{softmax<257>} is too large for register inference at this scale
and is instead inferred as BRAM, as observed in the synthesis results.

Step~3 executes inside the \texttt{acc} position loop. The attention
probability for timestep \texttt{t} is first truncated to a signed
INT8 value by the mapping $\texttt{prob\_int8} = \lfloor
\texttt{prob\_float} \times 127 \rfloor$, which assigns the unit
probability $1.0$ to the value $127$ and uses a fixed implicit scale of
$1/127$. The value vector is then loaded from the float KV cache,
quantised on the fly with its own per-position scale $s_V$, and the
weighted contribution is accumulated as:

\begin{equation}
  \Delta \texttt{xb}(i) =
    \left(\texttt{prob\_int8} \times \hat{v}_i\right)
    \cdot \frac{1}{127} \cdot s_V
  \label{eq:int8_value_acc}
\end{equation}

\noindent where $\hat{v}_i$ is the INT8 quantised value element. The
accumulation target \texttt{xb\_accum[head\_size]} is declared with
\texttt{\#pragma HLS ARRAY\_PARTITION variable=xb\_accum complete},
which maps it to four independent registers rather than a BRAM array.
This is important because the accumulation loop updates all four
elements of \texttt{xb\_accum} in each pipeline iteration. A BRAM-based
array would introduce read-after-write latency on the same address,
whereas a register file supports simultaneous access to all four ports.
Listing~\ref{lst:attn_int8} shows the Step~1 per-head Q quantisation
and the Step~2 INT32 dot product, which together constitute the core
change from the baseline.

\begin{lstlisting}[
  caption = {Per-head Q quantisation (Step~1) and INT32 score dot product
             (Step~2) from \texttt{forward\_fyp.cpp}. Step~1 executes
             once per head before the position loop. Step~2 executes for
             each attended position inside \texttt{iterate}. The reciprocal
             trick replaces \texttt{head\_size} divisions with one
             \texttt{fdiv} and \texttt{head\_size} \texttt{fmul} calls.},
  label = {lst:attn_int8}
]
// STEP 1: per-head Q quantisation (before iterate loop)
float max_val = 0.0f;
for (int i = 0; i < head_size; i++) {
    #pragma HLS PIPELINE
    float v = std::fabs(q_buf[i + q_offset]);
    if (v > max_val) max_val = v;
}
float q_scale_val = (max_val == 0.0f) ? 1.0f : (max_val / 127.0f);
float q_inv_scale = 1.0f / q_scale_val;  // one division hoisted out
int8_t q_int8[head_size];
for (int i = 0; i < head_size; i++) {
    #pragma HLS UNROLL
    q_int8[i] = (int8_t)round(q_buf[i + q_offset] * q_inv_scale);
}

// STEP 2: on-the-fly K quantisation + INT32 dot product (inside iterate)
iterate:
for (int t = 0; t <= pos; t++) {
    #pragma HLS PIPELINE
    // ... loaded float k_buffer[head_size] from key_cache, computed k_scale ...
    int32_t score_int = 0;
    for (int i = 0; i < head_size; i++) {
        #pragma HLS UNROLL
        score_int += (int32_t)q_int8[i] * (int32_t)k_int8[i]; // integer MAC
    }
    float score = ((float)score_int) * q_scale_val * k_scale_val
                  * (1.0f / sqrtf(head_size));   // single dequantise step
    att_buf[t + att_offset] = score;
}
\end{lstlisting}

C-simulation was performed for both implementations using the same
test configuration of \texttt{pos\,=\,3} and \texttt{loff\,=\,0}, with
identical query, key cache, and value cache values. This allows the
outputs to be compared directly as a measure of the quantisation error
introduced by the INT8 redesign. Table~\ref{tab:attn_csim_compare}
shows the \texttt{xb} output from both iterations alongside the
absolute difference between them.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrr}
    \toprule
    Index & Baseline \texttt{xb} (float) & INT8 \texttt{xb} & Abs difference & Relative error (\%) \\
    \midrule
    0 & 0.343754 & 0.337795 & 0.005959 & 1.73 \\
    1 & 0.313754 & 0.307415 & 0.006339 & 2.02 \\
    2 & 0.283754 & 0.279273 & 0.004481 & 1.58 \\
    3 & 0.253754 & 0.248893 & 0.004861 & 1.92 \\
    4 & 0.213827 & 0.209449 & 0.004378 & 2.05 \\
    5 & 0.183827 & 0.179730 & 0.004097 & 2.23 \\
    6 & 0.153827 & 0.150702 & 0.003125 & 2.03 \\
    7 & 0.123827 & 0.120983 & 0.002844 & 2.30 \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of \texttt{xb} outputs from the baseline float
           attention and the refined per-head INT8 attention for identical
           inputs at \texttt{pos\,=\,3}. The absolute difference and
           relative error in each element reflect the quantisation error
           introduced by the INT8 arithmetic path. Both kernels passed
           their respective csim runs with zero errors against their own
           software references.}
  \label{tab:attn_csim_compare}
\end{table}

Both implementations passed their csim runs with zero errors against
their respective software references. The absolute differences between
the two kernels' outputs range from approximately $0.003$ to $0.006$,
representing relative errors of $1.58\%$ to $2.30\%$ with respect to
the float baseline. This level of quantisation error is consistent with
what is expected from symmetric INT8 quantisation at \texttt{head\_size\,=\,4},
since each individual scale is computed from only four values, so a single
outlier or uneven distribution can shift the scale enough to round
adjacent values by up to half an INT8 step. At full TinyLlama scale
(\texttt{head\_size\,=\,64}), the per-head scale is computed from a
much larger pool of values, and the expected quantisation error per
element is correspondingly smaller.

\subsubsection{HLS Synthesis Results}
\label{subsec:attn_csyn}

Table~\ref{tab:attn_compare_resources} summarises the resource
utilisation of both attention implementations, and
Table~\ref{tab:attn_int8_latency} provides the per-pipeline-module
latency breakdown for the refined INT8 design. The baseline latency
breakdown was presented in Table~\ref{tab:attn_fp32_latency}.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    & \multicolumn{2}{c}{Baseline (float)} & \multicolumn{2}{c}{Refined (INT8)} \\
    Resource & Total & Core only & Total & Core only \\
    \midrule
    LUT  & 54{,}912 & 53{,}650 & 46{,}448 & 45{,}186 \\
    FF   & 27{,}584 & 26{,}673 & 27{,}873 & 26{,}962 \\
    BRAM & 4        & 0        & 36       & 32       \\
    DSP  & 192      & 192      & 122      & 122      \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation comparison between the baseline float
           attention and the refined per-head INT8 attention, both synthesised
           at prototype scale on the \texttt{xczu9eg-ffvb1156-2-e} at
           \SI{150}{\mega\hertz}. Core-only figures exclude the AXI master
           adapter and S-AXILite control interface.}
  \label{tab:attn_compare_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Pipeline stage & Trip count & II & Iter.\ latency & Total (cycles) \\
    \midrule
    Q max scan                            &   4 &  1 &  2 &    4 \\
    Q quantise                            &   4 &  1 &  1 &    4 \\
    \texttt{iterate} (K quant + INT32)    & 257 &  4 & 73 & 1{,}096 \\
    \texttt{max}     (softmax pass 1)     & 257 &  1 &  2 &  257 \\
    \texttt{exp}     (softmax pass 2)     &  17 &  2 & 18 &   51 \\
    \texttt{sum}     (softmax pass 3)     & 257 &  6 &  9 & 1{,}544 \\
    \texttt{norm}    (softmax pass 4)     &  17 &  1 &  6 &   22 \\
    \texttt{xb\_accum} initialise         &   4 &  1 &  1 &    4 \\
    \texttt{acc}     (V quant + weighted) & 257 &  6 & 76 & 1{,}611 \\
    \bottomrule
  \end{tabular}
  \caption{Per-pipeline-module latency breakdown of the refined per-head
           INT8 attention block at \SI{150}{\mega\hertz}, for one head at
           maximum context (\texttt{pos\,=\,256}, \texttt{MAXSIZE\,=\,257}).
           The outer \texttt{multihead\_attention} loop over
           \texttt{n\_heads\,=\,2} runs sequentially, giving a worst-case
           total of approximately 9{,}332 cycles.}
  \label{tab:attn_int8_latency}
\end{table}

The largest change in the resource comparison is the DSP
reduction from 192 to 122, a saving of 70 DSP48E2 blocks. In the
baseline, the 192 DSPs are almost entirely accounted for by the 16
parallel \texttt{fexp} units inside \texttt{softmax<16>} (112 DSPs)
plus the \texttt{fsub} max-subtract units (32 DSPs) and \texttt{fmul}
normalisation units (48 DSPs). In the refined design, the
\texttt{softmax<257>} instance with \texttt{UNROLL factor=16} on its
257-element buffer generates 16 parallel \texttt{fexp} units for the
\texttt{exp} loop's 17 effective iterations, but HLS maps only 70 DSPs
to the \texttt{exp} pipeline sub-module at this larger template size,
compared to 112 at \texttt{MAXSIZE=16}. The \texttt{norm} sub-module
uses 36 DSPs, and the \texttt{iterate} module 2 DSPs for the
dequantisation multiplications. The INT32 dot product itself, being
an integer accumulation rather than a floating-point multiply-add
chain, is implemented in LUT fabric without using DSP blocks.

The BRAM increase from 0 to 32 core BRAMs reflects two new memory
consumers in the refined design. The first is the \texttt{softmax<257>}
internal \texttt{buffer[257]} array. At $257 \times 4 = 1{,}028$ bytes
this exceeds the register inference threshold and is partitioned by HLS
into 8 BRAM banks of 2 BRAM\_18K each, using 16 BRAM\_18K blocks
in total. The second is the \texttt{generic\_round\_double\_s} IP core
instantiated four times for the per-element Q quantisation rounding,
using 4 BRAM\_18K blocks per instance (16 total). This rounding IP
uses BRAM-backed lookup tables to implement IEEE~754 round-to-nearest
in double precision, analogous to the \texttt{mask\_table} and
\texttt{one\_half\_table} ROM inferences observed in the
\texttt{quantize} standalone synthesis.

The \texttt{iterate} loop maintains II\,=\,4 in the refined design,
unchanged from the baseline. The root cause is identical, with the on-the-fly
K loading inside the pipelined position loop generating four concurrent
AXI read requests through the shared \texttt{gmem0} port, and the
adapter's transaction queue limiting the sustainable request rate to one
group of four reads per 4 clock cycles. The on-the-fly K quantisation
steps that are unique to the refined design add to the
iteration latency (73 cycles versus 42 in the baseline), but do not
change the II since those operations execute after the four reads have
been issued. The \texttt{acc} loop similarly retains II\,=\,6 from the
float multiply-accumulate dependency introduced by the
\texttt{prob\_int8\,$\times$\,v\_int8\,$\times$\,scale} dequantisation
chain, which presents the same \texttt{fmul}--\texttt{fadd} latency
bottleneck as the float baseline's direct accumulation. The structural
benefit of the INT8 redesign is therefore not a reduction in II at
prototype scale, but rather the elimination of cross-head scale
contamination and the 37\% reduction in DSP consumption, which becomes
the enabling factor for fitting the full-scale attention design within
the ZCU102's resource budget.

\subsection{Iteration 3: INT8 KV Cache}
\label{subsec:attn_kvint8}

The third iteration, implemented in \texttt{forward\_fyp\_kvint8.cpp},
extends the INT8 representation into the KV cache itself. In the first
two iterations, the key and value vectors are stored in DDR as float32
arrays and quantised on the fly when they are read back during the
\texttt{iterate} and \texttt{acc} loops. The third iteration instead
quantises K and V immediately after the QKV projections, before writing
them into the cache, so that subsequent reads during attention retrieve
pre-quantised INT8 data directly. This eliminates the on-the-fly
quantisation sequences from the inner attention loops, at the cost of
making the KV cache write slightly more expensive.

The function signature gains two new AXI arguments. \texttt{key\_cache}
and \texttt{value\_cache} change type from \texttt{float*} to
\texttt{int8\_t*}, and two new \texttt{float*} arguments
\texttt{key\_scales} and \texttt{value\_scales} carry the group-wise
scale factors stored alongside the INT8 data. At prototype scale with
$\texttt{kv\_dim} = 8$ and $\texttt{GS} = 4$, each timestep's KV head
slice requires two scale factors, and the scale arrays are indexed with
a separate \texttt{loff\_scale} offset computed as $l \times
\texttt{seq\_len} \times (\texttt{kv\_dim}/\texttt{GS})$.

Inside the \texttt{iterate} loop, the on-the-fly K loading from iteration~2,
which required loading float K, scanning for the maximum, computing
$s_K$, and quantising element by element, is replaced by two direct
reads, namely a burst read of \texttt{head\_size} INT8 values from
\texttt{key\_cache} and a single float read of the pre-stored scale from
\texttt{key\_scales}. The INT32 dot product and dequantisation formula
are unchanged from Equation~\ref{eq:int8_score}. Inside the \texttt{acc}
loop, the equivalent on-the-fly V quantisation is similarly replaced by
direct INT8 and scale reads from \texttt{value\_cache} and
\texttt{value\_scales}. The value accumulation formula also simplifies.
Iteration~2 mapped the softmax probability to INT8 before accumulating,
while iteration~3 keeps it as a float and folds it directly with the
V scale into a \texttt{combined\_scale = prob\,$\times$\,v\_scale}
scalar. The \texttt{acc} loop additionally gains
\texttt{\#pragma HLS UNROLL factor=2} on the outer position loop, processing
two timesteps per pipeline iteration.

The \texttt{softmax} call reverts to \texttt{softmax<16>} from the
\texttt{softmax<257>} used in iteration~2, since the iteration~3 loop
trip count bounds return to \texttt{min=1 max=16}. The softmax function
in this header also includes the \texttt{UNROLL factor=16} on the
\texttt{sum} loop and the buffer partition discussed in
\cref{subsec:softmax_iter3}, so the attention block inherits both the
performance regression and the timing violation of that unrolled softmax.
Listing~\ref{lst:attn_kvint8} contrasts the \texttt{iterate} loop bodies
of iterations~2 and~3 side by side, making the elimination of on-the-fly
K quantisation concrete.

\begin{lstlisting}[
  caption = {Contrast between the \texttt{iterate} loop bodies of
             iteration~2 (\texttt{forward\_fyp.cpp}) and iteration~3
             (\texttt{forward\_fyp\_kvint8.cpp}). Iteration~2 loads float
             K and performs a full max-scan, scale-compute, and quantise
             sequence per position. Iteration~3 replaces the entire sequence
             with two direct reads from the pre-quantised cache arrays.},
  label = {lst:attn_kvint8}
]
// ITERATION 2: on-the-fly K quantisation inside iterate
for (int t = 0; t <= pos; t++) {
    #pragma HLS PIPELINE
    float k_buffer[head_size];
    for (int i = 0; i < head_size; i++)
        k_buffer[i] = key_cache[i + key_offset]; // float DDR load
    float k_max = 0.0f;
    for (int i = 0; i < head_size; i++) {
        float v = std::fabs(k_buffer[i]);         // max scan
        if (v > k_max) k_max = v;
    }
    float k_scale = k_max / 127.0f;               // scale compute
    float k_inv   = 1.0f / k_scale;
    int8_t k_int8[head_size];
    for (int i = 0; i < head_size; i++)
        k_int8[i] = (int8_t)round(k_buffer[i] * k_inv); // quantise
    // ... INT32 dot product using k_int8 ...
}

// ITERATION 3: direct INT8 load from pre-quantised cache
for (int t = 0; t <= pos; t++) {
    #pragma HLS PIPELINE
    float k_scale = key_scales[scale_offset];    // one float scalar read
    int8_t k_buffer[head_size];
    for (int i = 0; i < head_size; i++)
        k_buffer[i] = key_cache[data_offset + i]; // INT8 load
    // ... INT32 dot product using k_buffer directly ...
}
\end{lstlisting}

Table~\ref{tab:attn_csim_all} extends the earlier two-iteration csim
comparison to include iteration~3. All three csim runs passed with zero
errors against their respective software references.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & Baseline \texttt{xb} (float) & Iter.~2 \texttt{xb} (INT8) & Iter.~3 \texttt{xb} (INT8-KV) \\
    \midrule
    0 & 0.343754 & 0.337795 & 0.343771 \\
    1 & 0.313754 & 0.307415 & 0.312900 \\
    2 & 0.283754 & 0.279273 & 0.284313 \\
    3 & 0.253754 & 0.248893 & 0.253442 \\
    4 & 0.213827 & 0.209449 & 0.213733 \\
    5 & 0.183827 & 0.179730 & 0.183540 \\
    6 & 0.153827 & 0.150702 & 0.154046 \\
    7 & 0.123827 & 0.120983 & 0.123853 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation \texttt{xb} outputs for all three attention
           iterations at identical inputs (\texttt{pos\,=\,3},
           \texttt{loff\,=\,0}). Iteration~3 outputs are closer
           to the float baseline than iteration~2, reflecting
           the higher fidelity of pre-quantised group-scale storage
           compared to per-position on-the-fly quantisation.}
  \label{tab:attn_csim_all}
\end{table}

The iteration~3 outputs are noticeably closer to the float baseline
than the iteration~2 outputs. The maximum absolute difference from
the float baseline reduces from approximately $0.006$ (iteration~2) to
approximately $0.001$ (iteration~3). This improvement occurs because
iteration~3 stores K and V with group-wise scales computed over
$\texttt{GS} = 4$ elements of the full \texttt{kv\_dim\,=\,8}-element
vector, whereas iteration~2 computed one scale per head slice
(\texttt{head\_size\,=\,4} elements) on the fly. Although both use the
same number of elements per group, the pre-stored group scales in
iteration~3 are computed once at the write site using the full integer
range of 127, while iteration~2's on-the-fly scales were subject to
additional rounding at each read. The net effect is a more accurate
reconstruction of the float activations at the point where they
enter the attention score computation.

\subsubsection{HLS Synthesis Results}
\label{subsec:attn_kvint8_csyn}
all three iterations, and Table~\ref{tab:attn_kvint8_latency} presents
the per-pipeline-module latency for iteration~3.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    & Baseline & Iter.~2 & \multicolumn{2}{c}{Iter.~3} \\
    Resource & (core) & (core) & Total & Core \\
    \midrule
    LUT  & 53{,}650 & 45{,}186 & 72{,}096 & 70{,}514 \\
    FF   & 26{,}673 & 26{,}962 & 35{,}036 & 33{,}947 \\
    BRAM & 0        & 32       & 22       & 18 \\
    DSP  & 192      & 122      & 194      & 194 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation across all three attention iterations
           on the \texttt{xczu9eg-ffvb1156-2-e} at \SI{150}{\mega\hertz}.
           Core figures for iterations~1 and~2 exclude the AXI master
           adapter and S-AXILite control interface. The same exclusions
           apply to the iteration~3 core column.}
  \label{tab:attn_all_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Pipeline stage & Trip count & II & Iter.\ latency & Total (cycles) \\
    \midrule
    Q max scan                                  &  4 &  1 &  2 &   4 \\
    Q quantise                                  &  4 &  1 &  1 &   4 \\
    \texttt{iterate} (INT8 K load + INT32)      & 16 &  5 & 26 & 100 \\
    \texttt{max}     (softmax pass 1)           & 16 &  1 &  2 &  16 \\
    \texttt{exp}     (softmax pass 2)           &  1 &  1 & 17 &  16 \\
    \texttt{sum}     (softmax pass 3, unrolled) &  1 & 96 & 97 & 185 \\
    \texttt{norm}    (softmax pass 4)           &  1 &  1 &  1 &   1 \\
    \texttt{xb\_accum} initialise               &  4 &  1 &  1 &   4 \\
    \texttt{acc}     (INT8 V load + weighted)   &  8 & 12 & 36 & 124 \\
    \bottomrule
  \end{tabular}
  \caption{Per-pipeline-module latency breakdown of the iteration~3
           INT8-KV attention block at \SI{150}{\mega\hertz}, for one
           head at \texttt{pos\,=\,15}. The \texttt{sum} stage reflects
           the unrolled softmax from \cref{subsec:softmax_iter3}.}
  \label{tab:attn_kvint8_latency}
\end{table}

In iteration~3, the \texttt{iterate} iteration latency drops from 42 cycles (baseline) and 73
cycles (iteration~2) to 26 cycles. This directly reflects the
elimination of the on-the-fly K quantisation sequence. Where
iteration~2's \texttt{iterate} body spent the majority of its 73
cycles on the max-scan, scale computation, and element-wise
quantisation of the loaded float K vector, iteration~3 instead
issues two AXI reads, a burst of \texttt{head\_size} INT8 bytes
from \texttt{key\_cache} and a single float from
\texttt{key\_scales}, and then proceeds directly to the integer dot
product. The achieved II is 5 rather than the 4 observed in the
baseline and iteration~2, because the two separate AXI read streams
(\texttt{key\_cache} and \texttt{key\_scales}) now issue through the
same \texttt{gmem0} bundle and require one additional cycle of
arbitration compared to the single float stream of the baseline.

The \texttt{acc} loop shows trip count~8 with II\,=\,12 and total
124 cycles. The \texttt{UNROLL factor=2} on the outer position loop
halves the trip count from 16 to 8, but also doubles the work per
pipeline iteration. The II\,=\,12 reflects the two simultaneous V
vector loads per iteration combined with the float
\texttt{fmul}--\texttt{fadd} accumulation chain, which has a combined
pipeline depth of approximately 10 cycles. Total \texttt{acc} latency
at 124 cycles is slightly more than the baseline's 112 cycles, as the
on-the-fly V scale load adds one extra memory access per iteration.

LUT usage increases
from iteration~2 (45,186 core) to iteration~3 (70,514 core). The
primary driver is not the attention block itself but the
\texttt{sum} pipeline module inherited from the iteration~3 softmax
function, which grew from 233 LUTs in the baseline to 2,416 LUTs
after the \texttt{UNROLL factor=16} change. The BRAM count reduced
from 32 core (iteration~2) to 18 core (iteration~3), because
\texttt{softmax<16>} is reinstated and its 64-byte buffer stays in
registers rather than BRAM, while the \texttt{generic\_round\_double\_s}
rounding IP instances remain from the per-head Q quantisation, each
still using 4 BRAM\_18K blocks. Both iteration~3 designs
(softmax and attention) carry the \SI{-0.217}{\nano\second} timing
violation introduced by the \texttt{sum} unroll, with an estimated
critical path of \SI{5.084}{\nano\second} against the
\SI{4.867}{\nano\second} budget.

% =============================================================================
\subsection{Iteration 4: I-BERT Integer Softmax Attention}
\label{subsec:attn_ibert}
% =============================================================================

Iteration~4 resolves the timing violation from iteration~3 by replacing the
floating-point \texttt{softmax<16>} with the integer \texttt{softmax\_ibert<16>}
described in \cref{subsec:softmax_ibert}. This change also requires two
structural modifications to the \texttt{iterate} and \texttt{acc} loops to
keep the attention score pipeline consistent with integer types throughout.

\subsubsection{Implementation}
\label{subsubsec:attn_ibert_impl}

The \texttt{att} buffer type changes from \texttt{float[]} to
\texttt{int32\_t[]}. In \texttt{iterate}, the full dequantisation step of
iteration~3, which multiplied \texttt{score\_int} by all three scale factors
before storing, is split into two parts. Only the K-scale partial result
\texttt{(int32\_t)((float)score\_int * k\_scale)} is stored in \texttt{att},
deferring the Q-scale and attention-scale factors to the softmax call. After
the position loop, the deferred factors are combined as \texttt{remaining\_scale
= q\_scale\_val * (1.0f / sqrtf(head\_size))} and passed to
\texttt{softmax\_ibert<16>}, which returns \texttt{int8\_t att\_probs[16]}
with probabilities in $[0, 127]$.

In the \texttt{acc} loop, \texttt{prob\_int8 = att\_probs[t]} replaces the
float probability from iteration~3. The combined scale becomes
\texttt{(1.0f/127.0f) * v\_scale}, where the factor $1/127$ dequantises the
integer probability back to the $[0, 1]$ range before the weighted value
accumulation. The outer position loop retains \texttt{\#pragma HLS PIPELINE}
and \texttt{\#pragma HLS UNROLL factor=2}, and the inner element loop retains
\texttt{\#pragma HLS UNROLL factor=16}, unchanged from iteration~3.

\subsubsection{C-Simulation Results}
\label{subsubsec:attn_ibert_csim}

Table~\ref{tab:attn_csim_all} from iteration~3 is extended in
Table~\ref{tab:attn_csim_all_iter4} to include the iteration~4 outputs for
the same input at \texttt{pos\,=\,3}.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrr}
    \toprule
    Index & Baseline \texttt{xb} & Iter.~2 \texttt{xb} & Iter.~3 \texttt{xb} & Iter.~4 \texttt{xb} \\
    \midrule
    0 & 0.343754 & 0.337795 & 0.343771 & 0.341732 \\
    1 & 0.313754 & 0.307415 & 0.312900 & 0.311557 \\
    2 & 0.283754 & 0.279273 & 0.284313 & 0.283688 \\
    3 & 0.253754 & 0.248893 & 0.253442 & 0.253512 \\
    4 & 0.213827 & 0.209449 & 0.213733 & 0.224567 \\
    5 & 0.183827 & 0.179730 & 0.183540 & 0.195083 \\
    6 & 0.153827 & 0.150702 & 0.154046 & 0.166292 \\
    7 & 0.123827 & 0.120983 & 0.123853 & 0.136808 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation \texttt{xb} outputs for all four attention iterations
           at \texttt{pos\,=\,3}, \texttt{loff\,=\,0}. Iteration~4 introduces
           additional approximation error relative to iteration~3 from the
           coarse LUT exponential, but all four csim runs passed with zero
           errors against their respective software references.}
  \label{tab:attn_csim_all_iter4}
\end{table}

The csim run completed with zero errors. Comparing the iteration~4 outputs
to the three preceding iterations, the values lie between iteration~2 and
the float baseline rather than close to iteration~3. For example, index~0
gives $0.341732$ in iteration~4 versus $0.343754$ (baseline), $0.337795$
(iteration~2), and $0.343771$ (iteration~3). The LUT exponential maps score
differences to coarse integer indices, concentrating the probability mass on
the leading positions differently from the true softmax, which is why the
output distribution differs visibly from iteration~3 even though the
underlying attention scores originate from the same inputs. The absolute
difference between iteration~4 and the float baseline is larger for the
lower-ranked elements (indices 4--7 diverge more noticeably), because those
positions have smaller exponent arguments that the LUT resolution can
represent less accurately. At full TinyLlama scale (\texttt{head\_size\,=\,64}),
the per-head scores span a wider integer range, giving the
\texttt{scale\_factor} more dynamic range to distribute across LUT indices.

\subsubsection{HLS Synthesis Results}
\label{subsec:attn_ibert_csyn}

Tables~\ref{tab:attn_all_resources_iter4} and \ref{tab:attn_kvint8_iter4_latency}
extend the resource and latency comparisons to include iteration~4.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    & \multicolumn{2}{c}{Iteration 3} & \multicolumn{2}{c}{Iteration 4} \\
    \cmidrule(lr){2-3} \cmidrule(lr){4-5}
    Resource & Total & Core & Total & Core \\
    \midrule
    LUT  & 72{,}096 & 70{,}514 & 33{,}410 & 31{,}828 \\
    FF   & 35{,}036 & 33{,}947 & 16{,}017 & 14{,}928 \\
    BRAM & 22       & 18       & 22       & 18 \\
    DSP  & 194      & 194      & 16       & 16 \\
    \bottomrule
  \end{tabular}
  \caption{Resource comparison between iteration~3 and iteration~4 attention
           implementations on the ZCU102. Replacing \texttt{softmax<16>}
           with \texttt{softmax\_ibert<16>} reduces DSP usage from 194 to 16
           and LUT usage from 70{,}514 to 31{,}828 core LUTs. BRAM usage
           is unchanged at 18 core blocks.}
  \label{tab:attn_all_resources_iter4}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Module & Latency min (cycles) & Latency max (cycles) \\
    \midrule
    Q quantisation (VITIS\_LOOP\_73\_2)  & 10  &  10 \\
    \texttt{iterate}                     & 28  & 103 \\
    \texttt{softmax\_ibert}              & 37  & 485 \\
    \texttt{acc}                         & 43  & 127 \\
    \texttt{xb} write (VITIS\_LOOP\_126\_6) &  6  &   6 \\
    \texttt{multihead\_attention} (total, 2 heads) & 178 & 1{,}536 \\
    \texttt{attn\_softmaxint8\_top} (total) & 208 & 1{,}566 \\
    \bottomrule
  \end{tabular}
  \caption{Per-module latency breakdown for the iteration~4 attention
           kernel at \SI{150}{\mega\hertz}. The minimum corresponds to
           \texttt{pos\,=\,0} (one attended position) and the maximum
           to \texttt{pos\,=\,15} (\texttt{MAX\_SIZE\,=\,16}).}
  \label{tab:attn_kvint8_iter4_latency}
\end{table}

The headline result of iteration~4 is the DSP reduction from 194 to 16. In
iteration~3, 192 of the 194 DSPs came from the 16 parallel 	exttt{fexp} IP core instances inside the unrolled
\texttt{softmax<16>} function. Replacing those with \texttt{softmax\_ibert}
eliminates all of them. The 16 DSPs that remain are accounted for by the
per-head floating-point arithmetic that cannot be avoided in the current design:
four \texttt{fmul\_32ns\_32ns\_32\_4\_max\_dsp\_1} instances (3 DSPs each) for
\texttt{remaining\_scale}, \texttt{combined\_scale}, and the
\texttt{val\_scaled\_f} multiplication inside \texttt{softmax\_ibert}, plus
2 DSPs each in the \texttt{iterate} and \texttt{acc} pipeline modules for their
respective multiply operations.

The LUT reduction from 70{,}514 to 31{,}828 core LUTs is similarly large,
a 55\% reduction. This reflects the removal of the floating-point exponential
hardware, which in iterations~1 and~3 dominated the LUT count through the
combinational logic required for each \texttt{fexp} pipeline stage.

Timing meets the \SI{6.667}{\nano\second} constraint, with an estimated path
of \SI{4.867}{\nano\second} plus \SI{1.800}{\nano\second} uncertainty
exhausting the budget with zero slack. This resolves the
\SI{-0.217}{\nano\second} violation present in iteration~3, which was
caused by the wide adder reduction tree from the \texttt{UNROLL factor=16}
on the floating-point \texttt{sum} loop. The integer arithmetic inside
\texttt{softmax\_ibert} avoids this entirely since integer adders have no
multi-cycle pipeline latency.

The \texttt{iterate} module latency of 28--103 cycles reflects the position-
dependent trip count, and the \texttt{softmax\_ibert} latency of 37--485 cycles
propagates the same variability from the runtime \texttt{size} parameter. The
worst-case total of 1{,}566 cycles (10.441\,µs per two-head attention block at
\texttt{pos\,=\,15}) is larger than iteration~3's worst case due to the longer
tail of \texttt{softmax\_ibert} at maximum size. The BRAM count of 18 core
blocks is unchanged from iteration~3, confirming that the on-chip memory
layout for the \texttt{att\_buf}, \texttt{xb\_accum}, and working arrays is
unaffected by the softmax replacement.

% =============================================================================
\subsection{Iteration 6: Full TinyLlama Scale}
\label{subsec:attn_update3}
% =============================================================================

\subsubsection{Implementation}
\label{subsubsec:attn_update3_impl}

The sixth iteration scales the attention block to full TinyLlama dimensions
(\texttt{dim\,=\,2048}, \texttt{n\_heads\,=\,32}, \texttt{n\_kv\_heads\,=\,4},
\texttt{seq\_len\,=\,2048}, \texttt{head\_size\,=\,64}), shown in
Listing~\ref{lst:attn_update3}. Two structural changes are applied beyond
the scale increase. First, \texttt{softmax\_ibert} is widened from
\texttt{MAX\_SIZE\,=\,16} to \texttt{MAX\_SIZE\,=\,seq\_len\,=\,2048},
using the \texttt{PIPELINE II\,=\,1} implementation from
\cref{subsec:softmax_ibert_update3} in place of the \texttt{UNROLL} version.
Second, the \texttt{PIPELINE} directive is removed from the outer position
loops (\texttt{score\_loop} and \texttt{value\_loop}) and applied only to the
inner \texttt{k\_load} and \texttt{v\_load} loops. At \texttt{head\_size\,=\,64},
pipelining the outer loop at II\,=\,1 would force HLS to fully unroll the
inner load loop, issuing 64 simultaneous AXI reads per outer iteration.
When \texttt{key\_cache} and \texttt{key\_scales} share the same AXI
bundle, issuing 64 simultaneous read requests from the fully-unrolled inner
loop saturates the adapter's request queue and stalls the pipeline. Moving
\texttt{PIPELINE} to the inner loop produces one sequential AXI read per
cycle, keeping the queue depth below two.
The key and value caches are placed on separate AXI bundles
(\texttt{gmem1} and \texttt{gmem2}) to prevent their transactions
from competing for the same port.

\begin{lstlisting}[
  caption = {Per-head score and value loops from \texttt{attn\_update3\_top.cpp}
             at full TinyLlama scale. \texttt{PIPELINE} is applied to the
             inner \texttt{k\_load} and \texttt{v\_load} loops only.
             The outer \texttt{score\_loop} and \texttt{value\_loop} are
             unpipelined to avoid issuing 64 simultaneous AXI reads.},
  label   = {lst:attn_update3}
]
multihead_attention:
for (int h = 0; h < n_heads; h++) {
    const int q_offset = h * head_size;
    const int kv_head  = h / kv_mul;

    // Per-head Q quantisation (find max, scale, convert to INT8)
    float max_abs = 0.0f;
    q_find_max:
    for (int i = 0; i < head_size; i++) { // PIPELINE II=1
        #pragma HLS PIPELINE II=1
        float val = fabsf(q_buf[i + q_offset]);
        if (val > max_abs) max_abs = val;
    }
    float q_scale_val = (max_abs == 0.0f) ? 1.0f : max_abs / 127.0f;
    float q_inv_scale = 1.0f / q_scale_val;
    int8_t q_int8[head_size];
    #pragma HLS ARRAY_PARTITION variable=q_int8 complete
    q_quantize:
    for (int i = 0; i < head_size; i++) { // UNROLL complete
        #pragma HLS UNROLL
        float qf = q_buf[i + q_offset] * q_inv_scale;
        int32_t r = (int32_t)(qf + (qf >= 0.0f ? 0.5f : -0.5f));
        if (r >  127) r =  127;
        if (r < -127) r = -127;
        q_int8[i] = (int8_t)r;
    }

    score_loop:                          // outer: no PIPELINE
    for (int t = 0; t <= pos; t++) {
        int8_t k_buffer[head_size];
        #pragma HLS ARRAY_PARTITION variable=k_buffer complete
        k_load:
        for (int i = 0; i < head_size; i++) { // PIPELINE II=1
            #pragma HLS PIPELINE II=1
            int kv_off = t*kv_dim + kv_head*head_size;
            k_buffer[i] = key_cache[kv_off + i];
        }
        int32_t score_int = 0;
        dot_qk:
        for (int i = 0; i < head_size; i++) { // UNROLL complete
            #pragma HLS UNROLL
            score_int += (int32_t)q_int8[i] * (int32_t)k_buffer[i];
        }
        att[t] = (int32_t)(((float)score_int) * key_scales[...]);
    }

    float remaining_scale = q_scale_val / sqrtf((float)head_size);
    softmax_ibert<seq_len>(att, att_probs, pos+1, remaining_scale);

    value_loop:                          // outer: no PIPELINE
    for (int t = 0; t <= pos; t++) {
        int8_t v_buffer[head_size];
        #pragma HLS ARRAY_PARTITION variable=v_buffer complete
        v_load:
        for (int i = 0; i < head_size; i++) { // PIPELINE II=1
            #pragma HLS PIPELINE II=1
            int v_off = t*kv_dim + kv_head*head_size;
            v_buffer[i] = value_cache[v_off + i];
        }
        float prob = ((float)att_probs[t]) / 127.0f;
        for (int i = 0; i < head_size; i++) { // UNROLL complete
            #pragma HLS UNROLL
            xb_accum[i] += prob * (float)v_buffer[i] * value_scales[...];
        }
    }
    // Write head result into xb_buf
    for (int i = 0; i < head_size; i++) xb_buf[q_offset+i] = xb_accum[i];
}
\end{lstlisting}

\subsubsection{C-Simulation Results}
\label{subsubsec:attn_update3_csim}

The testbench used the same input pattern as all prior attention iterations:
\texttt{pos\,=\,3}, with INT8 KV caches pre-populated for positions 0--3 using
the same \texttt{quantise\_to\_cache} helper as the iteration~3 testbench. The
first eight \texttt{xb} outputs are shown in
Table~\ref{tab:attn_csim_all_update3} alongside the preceding iteration's
values for comparison. All 2048 output elements agreed with the software
reference with zero absolute error. The csim run completed with zero errors.

\begin{table}[H]
  \centering
  \begin{tabular}{rrr}
    \toprule
    Index & Iter.~4 \texttt{xb} & Iter.~5 \texttt{xb} \\
    \midrule
    0 & 0.341732 & 0.207582 \\
    1 & 0.311557 & 0.181561 \\
    2 & 0.283688 & 0.154159 \\
    3 & 0.253512 & 0.116507 \\
    4 & 0.224567 & 0.089105 \\
    5 & 0.195083 & 0.063084 \\
    6 & 0.166292 & 0.037062 \\
    7 & 0.136808 & -0.001970 \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation \texttt{xb} outputs at \texttt{pos\,=\,3} for
           iteration~4 (prototype scale, $\texttt{n\_heads\,=\,2}$) and
           iteration~5 (full scale, $\texttt{n\_heads\,=\,32}$). The
           output values differ because the full-scale kernel processes
           32 heads at \texttt{head\_size\,=\,64} rather than 2 heads
           at \texttt{head\_size\,=\,4}, producing a different output
           range from the same KV cache input pattern.}
  \label{tab:attn_csim_all_update3}
\end{table}

\subsubsection{HLS Synthesis Results}
\label{subsubsec:attn_update3_csyn}

Tables~\ref{tab:attn_u3_timing} and \ref{tab:attn_u3_top_latency}
present the timing and latency summary. Table~\ref{tab:attn_u3_resources}
gives the resource utilisation and Table~\ref{tab:attn_u3_latency} the
per-module latency breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{attn\_update3\_top} at the
           \SI{150}{\mega\hertz} target clock.}
  \label{tab:attn_u3_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    206{,}200 & 13{,}510{,}040 &
    \SI{1.375}{\milli\second} & \SI{90.071}{\milli\second} &
    206{,}201 & 13{,}510{,}041 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{attn\_update3\_top} at
           \SI{150}{\mega\hertz}. The minimum corresponds to
           \texttt{pos\,=\,0} (one attended position per head) and the
           maximum to \texttt{pos\,=\,2047} (full context length).}
  \label{tab:attn_u3_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 95{,}556 & 92{,}942 & 274{,}080 & 34 \\
    FF   & 50{,}468 & 48{,}431 & 548{,}160 &  9 \\
    BRAM &       92 &       80 &   1{,}824 &  5 \\
    DSP  &      358 &      358 &   2{,}520 & 14 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{attn\_update3\_top} at full
           TinyLlama scale on the ZCU102. Core-only figures exclude the
           three AXI master adapters (\texttt{gmem1}--\texttt{gmem3},
           4\,BRAM + 512\,FF + 580\,LUT each) and the S-AXILite control
           block (501\,FF + 874\,LUT).}
  \label{tab:attn_u3_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Module & Latency min (cycles) & Latency max (cycles) \\
    \midrule
    Load \texttt{q}                          & 2{,}051   & 2{,}051 \\
    \texttt{q\_find\_max} (per head)         &      67   &      67 \\
    \texttt{k\_load} inner pipeline          &      74   &      74 \\
    \texttt{softmax\_ibert\_2048\_s}         &   6{,}200 &  6{,}200 \\
    \texttt{v\_load} inner pipeline          &      74   &      74 \\
    \texttt{score\_loop} (per head)          &     102   & 208{,}896 \\
    \texttt{value\_loop} (per head)          &     101   & 206{,}848 \\
    \texttt{multihead\_attention} (32 heads) & 202{,}080 & 13{,}505{,}920 \\
    \texttt{attn\_update3\_top} (total)      & 206{,}200 & 13{,}510{,}040 \\
    \bottomrule
  \end{tabular}
  \caption{Per-module latency breakdown for \texttt{attn\_update3\_top}
           at \SI{150}{\mega\hertz}. Minimum latency corresponds to
           \texttt{pos\,=\,0} and maximum to \texttt{pos\,=\,2047}.}
  \label{tab:attn_u3_latency}
\end{table}

Timing meets the \SI{6.667}{\nano\second} constraint with zero slack,
with an estimated path of \SI{4.867}{\nano\second}. The move from
\texttt{UNROLL} to \texttt{PIPELINE} in the inner load loops, and the
removal of the outer-loop \texttt{PIPELINE}, does not widen the critical
path since the score and value computations are dominated by memory
transfer latency rather than deep combinational logic.

The total latency spans from 206{,}200 cycles at \texttt{pos\,=\,0} to
13{,}510{,}040 cycles (\SI{90.071}{\milli\second}) at \texttt{pos\,=\,2047}.
At maximum context length the dominant cost is the 32-head multihead
attention loop at 13{,}505{,}920 cycles. Each head runs a score loop over
up to 2{,}048 positions. Per iteration, the key-load pipeline transfers
64 INT8 values at II\,=\,1 (64 cycles) and the dot product loop computes
the fully-unrolled INT8 result in a single cycle. With 2{,}048 positions
the score loop costs $2048 \times (64 + \text{overhead}) \approx
208{,}896$ cycles per head. The value loop has the same structure and
costs a comparable 206{,}848 cycles per head. The
\texttt{softmax\_ibert} call adds a fixed 6{,}200 cycles per head,
giving $32 \times (208{,}896 + 206{,}848 + 6{,}200) \approx
13{,}502{,}208$ cycles for the 32-head loop, with the remaining cycles
from the load, write, and inter-head control stages.

At \texttt{pos\,=\,0} each score loop runs for only 1 position,
reducing per-head cost to approximately $1 \times 102 + 6{,}200 =
6{,}302$ cycles, and the 32-head total to
$32 \times 6{,}302 \approx 201{,}664$ cycles,
matching the reported 202{,}080-cycle minimum for the attention loop.

The 80 core BRAMs are held by three on-chip buffers and the
\texttt{softmax\_ibert} submodule. The \texttt{q\_buf[2048]} and
\texttt{xb\_buf[2048]} float arrays each require 16 BRAM banks under
cyclic factor\,=\,16 partitioning, contributing 32 BRAMs. The
\texttt{att[2048]} int32 array adds a further 16 BRAM banks, for 48 BRAMs
across these three buffers. The \texttt{att\_probs[2048]} int8 array is
narrow enough (7 effective bits per entry) that HLS infers register fabric
rather than BRAM, giving 224 FF at zero BRAM cost. The
\texttt{softmax\_ibert\_2048\_s} module contributes 0 BRAM since it
operates on the pointer to \texttt{att} already in on-chip memory. The
remaining 32 BRAMs come from the per-head working arrays
(\texttt{q\_int8}, \texttt{k\_buffer}, \texttt{v\_buffer},
\texttt{xb\_accum}) which are each fully partitioned for the 64-element
head size and mapped to BRAM banks by HLS.

The 358 DSPs break down as follows. Across all 32 heads, the per-head
scale arithmetic instantiates 64 \texttt{fmul} units for the four
multiplications per head (\texttt{k\_scale}, \texttt{combined\_scale},
\texttt{q\_scale\_val}, and \texttt{remaining\_scale}), consuming
$64 \times 3 = 192$ DSPs. The \texttt{softmax\_ibert\_2048\_s} module
uses 6 DSPs for the two \texttt{fmul} units in its normalisation step.
The Expression block contributes 32 DSPs from the \texttt{fadd} units
in the 32 parallel \texttt{xb\_accum} accumulation chains, one per head.
The \texttt{dot\_qk} and \texttt{v\_accumulate} INT8 multiply operations
(\texttt{mul\_8s\_8s\_16\_1\_1} instances) use only LUT fabric and
contribute no DSPs, consistent with the smaller attention iterations where
INT8 multipliers mapped to LUTs throughout. The DSP count of 358 is higher
than iteration~4's 16 DSPs because the full-scale kernel instantiates 32
heads simultaneously rather than executing two heads sequentially, and the
per-head floating-point scale multiplications are each replicated 32 times.



% =============================================================================
\section{SwiGLU Feed-Forward Network}
\label{sec:ffn_impl}
% =============================================================================

\subsection{Implementation}
\label{subsec:ffn_impl_detail}

The SwiGLU FFN block is first enabled in the fifth kernel iteration
(\texttt{FFN\_baseline.cpp}). In all preceding iterations the FFN was
disabled and the kernel terminated after the attention residual addition.
The implementation follows the seven-step sequence of an FFN RMSNorm, quantisation of the normalised
activation, two parallel matmul projections using weights \texttt{w1} and
\texttt{w3}, the SwiGLU element-wise gating, quantisation of the gated
activation, and a final down-projection using weight \texttt{w2}.

After the FFN RMSNorm, the activation
\texttt{xb} is quantised into \texttt{xq} using \texttt{quantize<dim>}.
The gate projection \texttt{matmul<dim, hidden\_dim>} and the up projection
\texttt{matmul<dim, hidden\_dim>} are called sequentially using the same
quantised input \texttt{xq}, writing their outputs into \texttt{hb} and
\texttt{hb2} respectively. Because both calls share identical template
parameters, HLS synthesises a single \texttt{matmul\_8\_32\_s} instance
that is reused for both calls in sequence rather than duplicated.

The \texttt{swi\_glu} loop applies
the SiLU activation to the gate projection output and multiplies the
result element-wise by the up projection output. For each element
\texttt{hb[i]}, the SiLU function computes \texttt{hb[i] *= 1.0f /
(1.0f + expf(-hb[i]))}, then the gate product \texttt{hb[i] *= hb2[i]}
is applied and the result is written to a local buffer \texttt{hb\_out}
before being copied back to \texttt{hb}. The loop has \texttt{\#pragma HLS
UNROLL factor=4} and \texttt{\#pragma HLS PIPELINE}, meaning four elements
are processed in parallel. Because the SiLU function requires
\texttt{expf()}, each of the four parallel lanes instantiates a separate
\texttt{fexp} IP core, giving four floating-point exponential units in the
synthesised datapath.

After the gate, \texttt{hb} is
quantised into \texttt{hq} using \texttt{quantize<hidden\_dim>} and
the down projection \texttt{matmul<hidden\_dim, dim>} produces the FFN
output \texttt{xb}. This matmul has different template parameters from
the gate and up projections ($N = 32$, $D = 8$ versus $N = 8$, $D = 32$),
so HLS synthesises it as a separate \texttt{matmul\_32\_8\_s} instance.

\subsection{C-Simulation Results}
\label{subsec:ffn_csim}

The testbench supplied the same incremental input vector
$\mathbf{x} = [0.1, 0.2, \ldots, 0.8]$ used in all previous testbenches,
with FFN RMSNorm weights set to $1.0$ and small fixed weight matrices for
\texttt{w1}, \texttt{w2}, and \texttt{w3}. The hardware kernel output was
compared against a software reference running the same seven-step sequence
in full floating-point. Table~\ref{tab:ffn_csim} shows the results.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & HW \texttt{xb} & SW reference & Abs error \\
    \midrule
    0 &  0.001142 &  0.001125 & $1.70 \times 10^{-5}$ \\
    1 & -0.000340 & -0.000340 & $4.09 \times 10^{-7}$ \\
    2 & -0.002035 & -0.002002 & $3.33 \times 10^{-5}$ \\
    3 &  0.004489 &  0.004458 & $3.10 \times 10^{-5}$ \\
    4 & -0.000618 & -0.000633 & $1.44 \times 10^{-5}$ \\
    5 & -0.001165 & -0.001163 & $2.34 \times 10^{-6}$ \\
    6 &  0.001142 &  0.001125 & $1.70 \times 10^{-5}$ \\
    7 & -0.000340 & -0.000340 & $4.09 \times 10^{-7}$ \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for \texttt{ffn\_baseline} with input
           $\mathbf{x} = [0.1, 0.2, \ldots, 0.8]$ and unit RMSNorm
           weights. All eight elements agree with the software reference
           within the INT8 quantisation tolerance.}
  \label{tab:ffn_csim}
\end{table}

All eight elements agree with the software reference with zero failures
against a tolerance of $0.05$. The absolute errors are all below
$3.5 \times 10^{-5}$, considerably smaller than the tolerance, and
arise from the two INT8 quantisation steps in the pipeline. Rounding
error in \texttt{xq} propagates through both the gate and up projections,
and rounding error in \texttt{hq} propagates through the down projection.
The symmetry in the output, where elements at indices 0 and 6 and elements
at indices 1 and 7 share identical values, reflects the periodic structure
of the test weight matrices. The csim run completed with zero errors.

\subsection{HLS Synthesis Results}
\label{subsec:ffn_csyn}

Tables~\ref{tab:ffn_timing} and \ref{tab:ffn_top_latency} present the
timing and top-level latency summary. Table~\ref{tab:ffn_resources} gives
the resource utilisation and Table~\ref{tab:ffn_latency} the per-module
latency breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 4.867 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{ffn\_baseline\_top} at the
           \SI{150}{\mega\hertz} target clock.}
  \label{tab:ffn_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    1{,}487 & 1{,}487 & \SI{9.914}{\micro\second} & \SI{9.914}{\micro\second} & 1{,}488 & 1{,}488 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{ffn\_baseline\_top} at
           \SI{150}{\mega\hertz}.}
  \label{tab:ffn_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 36{,}686 & 35{,}526 & 274{,}080 & 13 \\
    FF   & 28{,}557 & 27{,}533 & 548{,}160 &  5 \\
    BRAM &       24 &       16 &   1{,}824 &  1 \\
    DSP  &       85 &       85 &   2{,}520 &  3 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{ffn\_baseline} synthesised at
           prototype dimensions ($\texttt{dim} = 8$, $\texttt{hidden\_dim} = 32$)
           on the ZCU102. Core-only figures exclude the two AXI master
           adapters (\texttt{gmem0\_m\_axi} and \texttt{gmem1\_m\_axi}).}
  \label{tab:ffn_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Stage & Module & Latency (cycles) \\
    \midrule
    FFN RMSNorm           & \texttt{rmsnorm\_8\_s}           & 132 \\
    Quantise \texttt{xb}  & \texttt{Pipeline\_main\_loop}    &  42 \\
    Load buffers          & \texttt{Pipeline\_2}, \texttt{Pipeline\_3} & 14 \\
    Gate projection (w1)  & \texttt{matmul\_8\_32\_s}        & 355 \\
    Up projection (w3)    & \texttt{matmul\_8\_32\_s} (reuse) & 355 \\
    SwiGLU + copy         & \texttt{Pipeline\_swi\_glu}, \texttt{Pipeline\_5} & 78 \\
    Quantise \texttt{hb}  & \texttt{Pipeline\_main\_loop1}, \texttt{Pipeline\_7}, \texttt{Pipeline\_8} & 92 \\
    Down projection (w2)  & \texttt{matmul\_32\_8\_s}        & 386 \\
    \bottomrule
  \end{tabular}
  \caption{Per-stage latency breakdown of \texttt{ffn\_baseline\_top}
           at \SI{150}{\mega\hertz} (prototype dimensions).
           Total function latency is 1{,}487 cycles (\SI{9.914}{\micro\second}).
           The gate and up projections share a single \texttt{matmul\_8\_32\_s}
           instance called sequentially.}
  \label{tab:ffn_latency}
\end{table}

Unlike the attention and softmax kernels, the FFN latency is fixed at
1{,}487 cycles with no variability, since all loop bounds are compile-time
constants. The three matrix multiplications together account for $355 +
355 + 386 = 1{,}096$ of those cycles, making them the dominant cost.
The gate and up projections both use \texttt{matmul<dim, hidden\_dim>}
($N = 8$, $D = 32$) and share one synthesised instance. Each call
achieves II\,=\,10 on the outer row loop, requiring 8 weight reads plus 2 scale
reads per row, and with $D = 32$ outer iterations this gives
$10 \times 32 + 35 = 355$ cycles per call. The down projection uses
\texttt{matmul<hidden\_dim, dim>} ($N = 32$, $D = 8$), which has a
higher per-row II of 40 ($32 + 8$ reads) but only 8 outer iterations,
giving $40 \times 8 + 66 = 386$ cycles.

The 85 DSPs are split across four functional areas. The \texttt{swi\_glu}
pipeline uses 34 DSPs for four parallel SiLU computations at
\texttt{UNROLL factor=4}: each lane requires one \texttt{fexp} IP core
(7 DSPs) for the \texttt{expf(-val)} evaluation, giving 28 DSPs, with
the remaining 6 DSPs from the sigmoid and gate multiplications.
\texttt{matmul\_32\_8\_s} uses 16 DSPs and \texttt{matmul\_8\_32\_s} uses
9 DSPs for their respective multiply-accumulate chains. Eight
\texttt{fmul\_32ns\_32ns\_32\_4\_max\_dsp\_1} instances (3 DSPs each, 24
total) handle the scale multiplications inside the two \texttt{quantize}
calls. The remaining 2 DSPs are from the \texttt{fadd} in the FFN
RMSNorm reduction.

The 16 core BRAMs come from four \texttt{generic\_round\_double\_s}
rounding IP instances (4 BRAMs each), the same rounding logic seen in the
attention kernels, inferred by HLS to implement the \texttt{round()} calls
inside \texttt{quantize}. Each of the two \texttt{quantize} calls generates
two such instances. The timing meets the \SI{6.667}{\nano\second}
constraint, with the estimated path of \SI{4.867}{\nano\second} plus
\SI{1.800}{\nano\second} uncertainty exhausting the budget with zero
slack.

% =============================================================================
\subsection{Full TinyLlama Scale}
\label{subsec:ffn_update3}
% =============================================================================

\subsubsection{Implementation}
\label{subsubsec:ffn_update3_impl}

The sixth iteration scales the FFN block to full TinyLlama dimensions
(\texttt{dim\,=\,2048}, \texttt{hidden\_dim\,=\,5632}), shown in
Listing~\ref{lst:ffn_update3}. All three \texttt{matmul} calls are
replaced with the tiled implementation from \cref{subsec:matmul_update3}.
The \texttt{ARRAY\_PARTITION cyclic factor} increases from 4 to 16
throughout to match \texttt{PARALLEL\_MACS\,=\,16}. The SwiGLU
\texttt{UNROLL factor} increases from 4 to 16 for the same reason, so
that all 16 BRAM banks of the \texttt{hb} and \texttt{hb2} buffers are
utilised in each pipeline cycle. The \texttt{round()} calls in
\texttt{quantize} are replaced by manual rounding arithmetic, eliminating
the \texttt{generic\_round\_double\_s} rounding IP that consumed 4 BRAMs
per \texttt{quantize} call in the prototype. Four AXI bundles assign the
three weight matrices and the activation input/output to independent HP
ports, preventing queue saturation that arises when multiple large
weight arrays share a single AXI bundle.

\begin{lstlisting}[
  caption = {FFN block from \texttt{FFN\_update3.cpp} at full TinyLlama
             scale (\texttt{dim\,=\,2048}, \texttt{hidden\_dim\,=\,5632}).
             The tiled \texttt{matmul} replaces the prototype's row-pipelined
             version. The \texttt{swi\_glu} loop uses \texttt{UNROLL factor=16}
             to match \texttt{PARALLEL\_MACS} and the 16-bank buffer partition.},
  label   = {lst:ffn_update3}
]
// FFN RMSNorm
rmsnorm<dim>(xb, x, w->rms_ffn_weight + l * dim);

// Gate and Up projections
quantize(&xq, xb);
matmul<dim, hidden_dim>(hb,  xq.q, xq.s,
                        (w->w1+l)->q, (w->w1+l)->s);
matmul<dim, hidden_dim>(hb2, xq.q, xq.s,
                        (w->w3+l)->q, (w->w3+l)->s);

// SwiGLU: SiLU(gate) * up
float hb_out[hidden_dim];
#pragma HLS ARRAY_PARTITION variable=hb_out cyclic factor=16

swi_glu:
for (int i = 0; i < hidden_dim; i++) {
    #pragma HLS PIPELINE II=1
    #pragma HLS UNROLL factor=16
    float val = hb[i];
    val *= (1.0f / (1.0f + expf(-val)));  // SiLU = x * sigmoid(x)
    val *= hb2[i];
    hb_out[i] = val;
}
memcpy(hb, hb_out, hidden_dim * sizeof(float));

// Down projection
quantize(&hq, hb);
matmul<hidden_dim, dim>(xb, hq.q, hq.s,
                        (w->w2+l)->q, (w->w2+l)->s);
\end{lstlisting}

The SwiGLU loop structure is unchanged from the prototype in that it
computes \texttt{SiLU(hb[i]) * hb2[i]} element-wise. The change from
\texttt{UNROLL factor=4} to \texttt{factor=16} instantiates 16 parallel
SiLU computation lanes, each containing one \texttt{expf()} unit. At
\texttt{hidden\_dim\,=\,5632}, the loop executes $5632/16 = 352$ pipeline
iterations, each processing 16 elements in the time one would otherwise
take. The \texttt{expf()} unit has a longer pipeline latency than simple
\texttt{fmul}, so the II\,=\,1 constraint requires the pipeline to accept
a new group every cycle even while earlier groups' exponential computations
are still in flight. HLS achieves this by staggering the 16 lanes.

\subsubsection{C-Simulation Results}
\label{subsubsec:ffn_update3_csim}

The testbench supplied a 2048-element input vector with unit RMSNorm weights
and small fixed-pattern float weight matrices that were quantised offline to
INT8 before being passed to the kernel. A software reference implemented the
same seven-step FFN sequence in floating-point arithmetic. The first eight
output elements are shown in Table~\ref{tab:ffn_u3_csim}.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrr}
    \toprule
    Index & HW \texttt{xb} & SW reference & Abs error \\
    \midrule
    0 & $-0.001738$ & $-0.002241$ & $5.03 \times 10^{-4}$ \\
    1 & $-0.001452$ & $-0.002026$ & $5.74 \times 10^{-4}$ \\
    2 & $ 0.003680$ & $ 0.003261$ & $4.19 \times 10^{-4}$ \\
    3 & $-0.001308$ & $-0.002060$ & $7.52 \times 10^{-4}$ \\
    4 & $ 0.000309$ & $-0.000249$ & $5.58 \times 10^{-4}$ \\
    5 & $ 0.003015$ & $ 0.002563$ & $4.52 \times 10^{-4}$ \\
    6 & $-0.001738$ & $-0.002241$ & $5.03 \times 10^{-4}$ \\
    7 & $-0.001452$ & $-0.002026$ & $5.74 \times 10^{-4}$ \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation output for \texttt{ffn\_update3} (first 8 of 2048
           elements). All absolute errors are well within the
           \num{0.05} tolerance. The small output magnitudes reflect
           three sequential quantised matrix multiplications with weight
           scale factors of \num{0.03}--\num{0.05}.}
  \label{tab:ffn_u3_csim}
\end{table}

All 2048 elements passed the \num{0.05} tolerance check. The absolute errors
lie in the range $4 \times 10^{-4}$ to $7.5 \times 10^{-4}$, two orders of
magnitude below the tolerance threshold. These errors arise from INT8
quantisation rounding across the three matrix multiplications and are of
similar magnitude to those seen in \texttt{ffn\_baseline}. The csim run
completed with zero errors.

\subsubsection{HLS Synthesis Results}
\label{subsubsec:ffn_update3_csyn}

Tables~\ref{tab:ffn_u3_timing} and \ref{tab:ffn_u3_top_latency}
present the timing and latency summary. Table~\ref{tab:ffn_u3_resources}
gives the resource utilisation and Table~\ref{tab:ffn_u3_latency} the
per-stage breakdown.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 7.456 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for \texttt{ffn\_update3\_top} at the
           \SI{150}{\mega\hertz} target clock. The estimated path of
           \SI{7.456}{\nano\second} plus \SI{1.800}{\nano\second}
           uncertainty exceeds the \SI{6.667}{\nano\second} budget by
           \SI{2.589}{\nano\second}.}
  \label{tab:ffn_u3_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (absolute)} &
    \multicolumn{2}{c}{Interval (cycles)} & Pipeline \\
    min & max & min & max & min & max & type \\
    \midrule
    38{,}095{,}467 & 38{,}095{,}467 &
    \SI{0.284}{\second} & \SI{0.284}{\second} &
    38{,}095{,}468 & 38{,}095{,}468 & no \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for \texttt{ffn\_update3\_top}
           at \SI{150}{\mega\hertz}.}
  \label{tab:ffn_u3_top_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core only) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 67{,}432 & 63{,}918 & 274{,}080 & 23 \\
    FF   & 35{,}356 & 32{,}635 & 548{,}160 &  6 \\
    BRAM &      352 &      336 &   1{,}824 & 18 \\
    DSP  &      307 &      307 &   2{,}520 & 12 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of \texttt{ffn\_update3\_top} on the
           ZCU102 (\texttt{xczu9eg-ffvb1156-2-e}) at full TinyLlama
           scale. Core-only figures exclude the four AXI master adapters
           (\texttt{gmem0}--\texttt{gmem3}, 4\,BRAM + 512\,FF + 580\,LUT
           each) and the S-AXILite control block (673\,FF + 1{,}194\,LUT).}
  \label{tab:ffn_u3_resources}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Stage & Module & Latency (cycles) \\
    \midrule
    Load \texttt{x\_in}                       & \texttt{Pipeline\_load\_x}     &          2{,}051 \\
    \texttt{rmsnorm\_2048\_s}                  & \texttt{rmsnorm\_2048\_s}       &         14{,}529 \\
    \texttt{quantize\_2048\_s} ($\mathbf{xq}$) & \texttt{quantize\_2048\_s}     &          7{,}780 \\
    Gate projection $\mathbf{W}_1$ (reused for $\mathbf{W}_3$) & \texttt{matmul\_2048\_5632\_s} & 12{,}744{,}452 \\
    \texttt{swi\_glu}                          & \texttt{Pipeline\_swi\_glu}    &             390 \\
    \texttt{memcpy} + write                    & \texttt{Pipeline\_3}           &          5{,}634 \\
    \texttt{quantize\_5632\_s} ($\mathbf{hq}$) & \texttt{quantize\_5632\_s}     &         21{,}388 \\
    Down projection $\mathbf{W}_2$             & \texttt{matmul\_5632\_2048\_s} & 12{,}552{,}708 \\
    Write \texttt{xb\_out}                     & \texttt{Pipeline\_write\_out}  &          2{,}051 \\
    \bottomrule
  \end{tabular}
  \caption{Per-stage latency breakdown of \texttt{ffn\_update3\_top}
           at \SI{150}{\mega\hertz}. Total latency is 38{,}095{,}467 cycles
           (\SI{0.284}{\second}). HLS infers one \texttt{matmul\_2048\_5632\_s}
           instance reused for both the gate ($\mathbf{W}_1$) and up
           ($\mathbf{W}_3$) projections, so the listed cycles represent
           two sequential calls through the same module.}
  \label{tab:ffn_u3_latency}
\end{table}

The synthesis reports a timing violation of \SI{2.589}{\nano\second}
at \SI{150}{\mega\hertz}. The estimated path is \SI{7.456}{\nano\second},
exceeding the \SI{6.667}{\nano\second} target. This arises from the
\texttt{rmsnorm\_2048\_s} submodule, whose 16-way parallel reduction tree
cannot close at this clock frequency at $S = 2048$, as analysed in
\cref{subsubsec:rmsnorm_update3_csyn}. The remaining submodules, including
the tiled matmuls and the SwiGLU pipeline, all meet the timing constraint
individually and are not on the critical path.

The total latency of 38{,}095{,}467 cycles (\SI{0.284}{\second}) is
determined almost entirely by the three tiled matmul calls.
HLS infers a single \texttt{matmul\_2048\_5632\_s} module and reuses it
sequentially for both the gate ($\mathbf{W}_1$) and the up ($\mathbf{W}_3$)
projections, since both calls have identical template parameters
\texttt{<dim,\,hidden\_dim>}. The two sequential calls therefore
contribute $2 \times 12{,}744{,}452 = 25{,}488{,}904$ cycles. The down
projection uses a separate \texttt{matmul\_5632\_2048\_s} instance
(different template parameters \texttt{<hidden\_dim,\,dim>}) and
contributes 12{,}552{,}708 cycles. Together the three matmul calls account
for $38{,}041{,}612$ of the $38{,}095{,}467$ total cycles, leaving only
53{,}855 cycles for all other stages combined, covering the input
load (2{,}051), \texttt{rmsnorm} (14{,}529), two \texttt{quantize}
calls (7{,}780 and 21{,}388), the \texttt{swi\_glu} loop (390), and
the output write (2{,}051).
The \texttt{swi\_glu} loop, despite processing 5{,}632 elements with 16
parallel SiLU lanes each containing an \texttt{expf()} unit, runs in
only 390 cycles at II\,=\,1 because the 16-lane unroll reduces the trip
count to $5632/16 = 352$ iterations.

The 336 core BRAMs come from two sources. The top-level wrapper holds
six static local arrays totalling 96 BRAM banks: \texttt{hb},
\texttt{hb2}, and \texttt{hb\_out} each hold 5{,}632 floats and with
cyclic factor\,=\,16 produce 16 banks of 352 entries each. \texttt{hq.q}
holds 5{,}632 int8 values in 16 BRAM banks. \texttt{xb} and \texttt{xq.q}
hold 2{,}048 floats and 2{,}048 int8 values respectively, each requiring
16 BRAM banks at that size. The submodule instances contribute the
remaining 240 BRAMs: \texttt{rmsnorm\_2048\_s} uses 48, both
\texttt{matmul\_2048\_5632\_s} and \texttt{matmul\_5632\_2048\_s} each
use 80, and both \texttt{quantize} variants use 16 each.

Comparing against the prototype FFN baseline, the BRAM count grows from
16 to 336 core BRAMs. At prototype scale the six static arrays each hold
at most 32 floats (for \texttt{hb} and \texttt{hb2} at
\texttt{hidden\_dim\,=\,32}), which fit in register files. At full scale
the same arrays hold thousands of entries and cross the BRAM threshold
on each of their 16 banks. The elimination of the
\texttt{generic\_round\_double\_s} rounding IP confirms that the manual
rounding change successfully removes the 4-BRAM-per-call overhead that
contributed 16 BRAMs in the prototype.

The 307 DSPs break into three groups. The \texttt{swi\_glu} pipeline
accounts for 142 DSPs from its 16 parallel SiLU computation lanes, where each
lane contains one \texttt{fexp} IP unit and the sigmoid and gate
multiplications, and with 16 lanes active simultaneously the DSP
requirement is much higher than the prototype's 4-lane 34-DSP cost.
The two tiled matmul modules use 35 and 32 DSPs respectively for the
floating-point accumulation step \texttt{val += (float)ival * ws * xs}.
The remaining 130 DSPs come from the 32 \texttt{fmul} instances across
the two \texttt{quantize} calls and the scale multiplications in the
\texttt{rmsnorm} core.

% =============================================================================
\section{Top-Level Forward Function}
\label{sec:forward_impl}
% =============================================================================

The top-level \texttt{forward} function in \texttt{forward.cpp} is the
entry point of the FPGA kernel. It is the function registered with
Vitis HLS via \texttt{set\_top forward} and is the only
\texttt{extern~"C"} function in the design. All the primitives
described in Sections~\ref{sec:rmsnorm_impl}
through~\ref{sec:attn_impl} are inlined into this single synthesised
module, as Vitis HLS resolves the template instantiations at compile time
and merges the function bodies into one hardware design with shared
AXI interfaces and shared on-chip buffers. The top-level function is
therefore the composition point where all prior design decisions
interact, and the choices made here about the AXI interface
configuration, buffer organisation, and loop structure determine the
overall hardware behaviour of the prototype kernel.

\subsection{AXI Interface and Buffer Layout}
\label{subsec:forward_axi}

The \texttt{forward} function declares six arguments. Two are scalar
integers passed through the S-AXILite control register interface:
\texttt{token} (the current input token index) and \texttt{pos} (the
current sequence position). The remaining four are memory-mapped
pointers to DDR-resident arrays, each assigned to one of two AXI
master bundles via \texttt{\#pragma HLS INTERFACE m\_axi} directives.

The prototype uses two AXI master bundles. The first bundle,
\texttt{gmem0}, has three ports simultaneously, namely the
\texttt{transformer} struct pointer (which contains all model weights
and the embedding table), \texttt{key\_cache}, and
\texttt{value\_cache}. The second bundle, \texttt{gmem1}, carries only
\texttt{out}, the output activation vector. This two-bundle arrangement
is depicted in Table~\ref{tab:axi_bundles}.

\begin{table}[H]
  \centering
  \begin{tabular}{lll}
    \toprule
    Bundle  & Port(s)                                          & Contents \\
    \midrule
    gmem0   & \texttt{transformer}, \texttt{key\_cache},
               \texttt{value\_cache}                          & Weights, KV cache \\
    gmem1   & \texttt{out}                                    & Output activation \\
    \bottomrule
  \end{tabular}
  \caption{AXI master bundle assignment in the prototype
           \texttt{forward} kernel (\texttt{forward.cpp}).
           All weight and KV cache traffic shares a single AXI master
           port on \texttt{gmem0}.}
  \label{tab:axi_bundles}
\end{table}

Sharing \texttt{key\_cache} and \texttt{value\_cache} on the same
bundle as the transformer weights is acceptable at prototype scale
because the small dimensions keep the total AXI transaction volume low.
At prototype configuration ($\texttt{dim} = 8$,
$\texttt{seq\_len} = 16$, $\texttt{n\_layers} = 2$), the number of
outstanding AXI read transactions generated within any single pipeline
stage remains within the adapter's 16-slot queue limit.

At full TinyLlama scale, however, this shared-bundle arrangement
becomes untenable. The \texttt{matmul} weight reads and the KV cache
reads in the \texttt{iterate} loop both issue through \texttt{gmem0}
concurrently. With \texttt{dim} = 2048 and a completely unrolled inner
\texttt{head\_size} = 64 element loop inside a pipelined outer position
loop, the \texttt{iterate} loop generates 64 simultaneous read requests
targeting the key cache while the weight loading logic is issuing its
own burst reads. The adapter's transaction queue fills and the pipeline stalls.
The fix is to separate the KV caches onto dedicated bundles:
\texttt{gmem0} for weights, \texttt{gmem1} for the key cache,
\texttt{gmem2} for the value cache, and \texttt{gmem3} for the output.
This four-bundle configuration is used in the full-scale kernel but is
not needed at prototype scale.

All intermediate activation buffers are declared as \texttt{static}
local arrays inside \texttt{forward}, which causes HLS to allocate
them as persistent on-chip storage rather than stack-allocated memory.
The buffers \texttt{x}, \texttt{xb}, \texttt{xb2}, \texttt{q},
\texttt{k}, \texttt{v}, \texttt{att}, \texttt{xq}, and \texttt{hq} are
all declared this way, and each receives a \texttt{\#pragma HLS
ARRAY\_PARTITION cyclic factor=16} directive so that the 16-wide
unrolled loop bodies in the attention block have sufficient memory ports
to sustain their read and write bandwidth requirements. The
\texttt{cyclic factor=16} partitioning splits each array into 16
independently addressable banks, allowing 16 simultaneous accesses per
clock cycle.

\subsection{Layer Loop and KV Cache Management}
\label{subsec:forward_layers}

The body of \texttt{forward} follows a fixed sequence for each decode
step. Before the layer loop begins, the token embedding is loaded from
the float \texttt{token\_embedding\_table} field of the transformer
weights struct via a single \texttt{memcpy} call, copying
\texttt{dim\,$\times$\,sizeof(float)} bytes into the activation buffer
\texttt{x}. This lookup is $\mathcal{O}(1)$ in the sense that it reads
one row of a table regardless of sequence length, but it is
memory-bandwidth-bound because the embedding matrix is stored in DDR
and must be transferred over \texttt{gmem0}.

The \texttt{main\_forward\_loop} then iterates over all \texttt{L =
n\_layers} transformer layers sequentially. The structure within each
layer iteration proceeds as follows. First, \texttt{rmsnorm<dim>} is
applied to \texttt{x}, writing the normalised result into \texttt{xb}.
The activation \texttt{xb} is then quantised into the
\texttt{QuantizedTensor} struct \texttt{xq} via \texttt{quantize(\&xq,
xb, GS)}. Three \texttt{matmul} calls consume \texttt{xq} to compute
the Q, K, and V projections, using \texttt{matmul<dim,dim>} for Q,
\texttt{matmul<dim,kv\_dim>} for K, and \texttt{matmul<dim,kv\_dim>}
for V, where \texttt{kv\_dim = (dim\,$\times$\,n\_kv\_heads) /
n\_heads}. At prototype scale where \texttt{n\_kv\_heads = n\_heads =
2}, \texttt{kv\_dim = dim = 8}, so all three projections operate on
identically sized vectors.

After the QKV projections, the current key and value vectors are
written into the KV cache at the position corresponding to the current
layer and timestep. The layer offset \texttt{loff = l\,$\times$\,seq\_len\,$\times$\,kv\_dim}
indexes the start of the current layer's portion of the cache, and the
position within that layer is \texttt{pos\,$\times$\,kv\_dim}. Both
the key and value vectors are written as float32 via \texttt{memcpy},
reflecting the prototype's use of a floating-point KV cache.

The multihead attention block then executes as described in
\cref{sec:attn_impl}, reading from the accumulated KV cache entries at
positions 0 through \texttt{pos} and producing the weighted value
accumulation in \texttt{xb}. The output projection matrix
$\mathbf{W}_O$ is not applied in the prototype. The commented-out
\texttt{matmul<dim,dim>} call for \texttt{wo} means that \texttt{xb}
is used directly in the residual addition rather than \texttt{xb2}.
The residual loop then adds \texttt{xb} element-wise into \texttt{x}
with \texttt{\#pragma HLS UNROLL factor=64}, completing one transformer
layer.

After all \texttt{n\_layers} iterations, the final activation \texttt{x}
is written to the output buffer via \texttt{memcpy(out, x,
dim\,$\times$\,sizeof(float))}. The FFN sub-block, the final RMSNorm,
and the language model head projection are all commented out and do
not execute in the prototype kernel.

\subsection{Known Limitations and Design Decisions}
\label{subsec:forward_limitations}

Three deliberate simplifications were made in the prototype kernel to
focus the implementation effort on the attention sub-block and to
remain within the resource and complexity budget of the initial
development phase. These are documented here as engineering scope
decisions rather than defects, and each is accompanied by the
corresponding fix that would be required in a production implementation.

The first simplification is the omission of the output projection
$\mathbf{W}_O$. In the full transformer, after the weighted value
accumulation \texttt{xb} is computed, it is passed through a final
linear projection \texttt{matmul<dim,dim>} using the weight matrix
\texttt{wo} to mix information across heads before the residual
addition. The prototype comments out this call and uses \texttt{xb}
directly in the residual, meaning the output dimension of the softmax
attention already matches \texttt{dim} without the projection. This
reduces the compute cost of each layer iteration by one \texttt{dim
$\times$ dim} matrix--vector multiply, which at full scale
(\texttt{dim = 2048}) is the second most expensive operation in the
attention block after the Q projection. Restoring the output projection
requires uncommenting the \texttt{quantize} and \texttt{matmul<dim,dim>}
calls and updating the residual to add \texttt{xb2} rather than
\texttt{xb}.

The second simplification is the complete omission of the SwiGLU
feed-forward network. The FFN sub-block, which follows the attention
residual in each transformer layer, would contribute three further
matrix--vector multiplications and the SwiGLU activation function.
At full TinyLlama scale the FFN weight matrices are larger than the
attention matrices, making the FFN the dominant contributor to both
the per-layer compute cost and the memory bandwidth requirement. The
prototype excludes the FFN entirely, meaning the closing brace of
\texttt{main\_forward\_loop} appears immediately after the first
residual addition. This conscious omission was taken to keep the
prototype kernel focused on validating the attention hardware, which
is the architecturally novel component. The FFN implementation in
\texttt{FFN\_update2.cpp} and \texttt{FFN\_update3.cpp} provides the
full SwiGLU implementation with INT8 quantisation.

The third simplification is the use of a floating-point KV cache. The
key and value vectors are stored in DDR as \texttt{float32} arrays
and retrieved as \texttt{float32} in the attention block. In the full
kernel, these vectors are stored as INT8 with accompanying group scales
to reduce the KV cache memory footprint by a factor of four and to
halve the DDR bandwidth required when reading past positions during
the \texttt{iterate} loop. The float cache was retained in the
prototype because it avoids the additional quantise-before-write and
dequantise-on-read steps, keeping the attention data path simpler while
functional correctness was being established. Bug~2 discovered during
on-board testing of the full-scale kernel is directly related to this
transition. The full kernel reads token embeddings from
\texttt{q\_tokens} (the quantised embedding table) rather than
\texttt{token\_embedding\_table} (the float table), but the host
checkpoint loader only populated the quantised table. The prototype
avoids this issue by always reading from the float table.
