% Chapter5/chapter5.tex -- Results and Discussion

\chapter{Results and Discussion}
\label{ch:results}

\section{Experimental Setup}
\label{sec:exp_setup}

All HLS synthesis and C-simulation experiments were performed using Vitis HLS
2021.2, targeting the \texttt{xczu9eg-ffvb1156-2-e} device on the Xilinx
ZCU102 evaluation board at a design clock of \SI{150}{\mega\hertz}
(\SI{6.667}{\nano\second} period). Development proceeded in two configurations.
A scaled-down prototype configuration ($d_{\text{model}} = 8$,
$d_{\text{ff}} = 32$, $T = 16$) was used to establish each kernel
primitive's structure and verify its logic through HLS C-simulation before
scaling to full model dimensions. The full TinyLlama-1.1B configuration
($d_{\text{model}} = 2048$, $d_{\text{ff}} = 5632$, $H = 32$,
$G = 4$, $T = 2048$, $\text{GS} = 64$), summarised in
Table~\ref{tab:config_comparison}, was used for the synthesis and
functional verification results reported in this chapter.

The kernel design is implemented in the sixth kernel iteration
(\texttt{FFN\_update3}), which incorporates the tiled matrix--vector
multiplication, the I-BERT integer softmax, the AXI interface
restructuring for the attention loops, and the corrected
\texttt{QuantizedTensor} scale
array layout described in \cref{ch:implementation}. All synthesis results
in this chapter refer to this iteration unless stated otherwise.

Functional verification was performed using HLS C-simulation, which compiles
the kernel source together with a software testbench. Two levels of
verification were performed. The first was standalone verification of each
primitive in isolation, and the second was end-to-end verification of the
complete \texttt{forward()} function through a multi-token stress test.

Hardware deployment on the ZCU102 was not completed within the scope of this
project. The Vivado implementation flow, which translates the synthesised RTL
into a placed-and-routed bitstream, was not run due to time constraints. The
throughput and latency figures reported in \cref{sec:performance} are therefore
derived from HLS synthesis estimates at the target clock frequency, and should
be treated as indicative estimates of the achievable hardware performance
pending post-implementation validation.

\section{HLS Synthesis Results}
\label{sec:synthesis_results}

\subsection{Full Forward Kernel}
\label{subsec:forward_synthesis}

Table~\ref{tab:forward_timing} presents the timing summary and
Table~\ref{tab:forward_resources} the resource utilisation for the complete
\texttt{forward()} kernel after HLS synthesis. The kernel implements one full
TinyLlama-1.1B decode step across all 22 transformer layers, including the
embedding lookup, attention sub-block, and SwiGLU FFN sub-block per layer.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Clock & Target (ns) & Estimated (ns) & Uncertainty (ns) \\
    \midrule
    ap\_clk & 6.667 & 7.456 & 1.800 \\
    \bottomrule
  \end{tabular}
  \caption{Timing summary for the \texttt{forward()} kernel at the
           \SI{150}{\mega\hertz} target clock. The estimated path of
           \SI{7.456}{\nano\second} plus \SI{1.800}{\nano\second}
           uncertainty exceeds the \SI{6.667}{\nano\second} budget by
           \SI{2.589}{\nano\second}.}
  \label{tab:forward_timing}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Resource & Used (total) & Used (core) & Available & Utilisation (\%) \\
    \midrule
    LUT  & 200{,}486 & 197{,}694 & 274{,}080 & 73 \\
    FF   & 100{,}006 &  97{,}589 & 548{,}160 & 18 \\
    BRAM &       652 &       610 &   1{,}824 & 35 \\
    DSP  &       709 &       709 &   2{,}520 & 28 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of the \texttt{forward()} kernel after HLS
           synthesis on the ZCU102.
           Core-only figures exclude the four AXI master adapters.
           The \texttt{gmem0} adapter consumes 30\,BRAM, 881\,FF, and 1{,}052\,LUT.
           The \texttt{gmem1}--\texttt{gmem3} adapters each consume 4\,BRAM,
           512\,FF, and 580\,LUT.}
  \label{tab:forward_resources}
\end{table}

The timing violation of \SI{2.589}{\nano\second} at \SI{150}{\mega\hertz} is
caused by the \texttt{rmsnorm\_2048\_s} submodule, whose 16-way reduction tree
creates a combinational path that cannot close at this frequency at $S = 2048$.
The same violation was observed for the standalone \texttt{rmsnorm} and
\texttt{ffn} primitives in \cref{subsec:rmsnorm_update3,subsubsec:ffn_update3_csyn},
confirming that the forward kernel inherits this path from its submodules.
A lower target clock frequency, or replacing the reduction tree with a
pipelined partial-sum accumulator, would resolve this violation.

The LUT utilisation of 73\% is the most constraining resource figure in the
report. The Vivado place-and-route tool typically requires 10--20\% headroom
above the raw HLS estimates to close routing, so this design sits at the edge
of what the device can accommodate. The high LUT count arises from three
distinct sources. The first is the 64-element fully-unrolled \texttt{q\_quantize}
and \texttt{dot\_qk} loops in the multihead attention block, where 64 parallel
float-to-int8 converters and 64 parallel INT8 multipliers each map to LUT
fabric. The second is the 12{,}340 LUTs in the Expression block from the
combinational comparison and shift logic across all 22 layers. The third is
the 21{,}928 LUT multiplexer network that arbitrates the large number of
memory bank accesses throughout the design.

The 610 core BRAMs break into two groups. The Memory section of the synthesis
report accounts for 210 BRAMs from the on-chip activation buffers, comprising
the six static float arrays (\texttt{x}, \texttt{xb}, \texttt{xb2}, \texttt{hb},
\texttt{hb2}, \texttt{q}) and the quantised working arrays (\texttt{xq},
\texttt{hq}, \texttt{att}, \texttt{att\_probs}), each requiring 16 BRAM banks
under cyclic factor\,=\,16 partitioning at full model dimensions. The Instance
section contributes 400 BRAMs from the seven synthesised submodule instances.
The \texttt{softmax\_ibert\_2048\_s} module uses no BRAM, confirming that the
I-BERT integer path avoids the large floating-point exponential lookup tables
that dominated resource usage in the prototype attention kernels.

Table~\ref{tab:submodule_resources} summarises the resource contribution of
the key synthesised submodules.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Module & BRAM & DSP & FF & LUT \\
    \midrule
    \texttt{rmsnorm\_2048\_s}      &  48 &  98 &  8{,}989 &  9{,}494 \\
    \texttt{quantize\_2048\_s}     &  16 &   5 &  1{,}480 &  3{,}376 \\
    \texttt{quantize\_256\_s}      &   0 &   5 &  1{,}703 &  3{,}382 \\
    \texttt{quantize\_5632\_s}     &  16 &   0 &  1{,}113 &  3{,}074 \\
    \texttt{matmul\_2048\_2048\_s} &  80 &  35 &  4{,}168 & 12{,}370 \\
    \texttt{matmul\_2048\_256\_s}  &  80 &  35 &  4{,}153 & 12{,}364 \\
    \texttt{matmul\_2048\_5632\_s} &  80 &  35 &  4{,}165 & 12{,}373 \\
    \texttt{matmul\_5632\_2048\_s} &  80 &  32 &  4{,}102 & 12{,}519 \\
    \texttt{softmax\_ibert\_2048\_s} &  0 &   0 &  1{,}199 &  3{,}722 \\
    \bottomrule
  \end{tabular}
  \caption{Resource utilisation of the key synthesised submodule instances
           within the \texttt{forward()} kernel.}
  \label{tab:submodule_resources}
\end{table}

\subsection{Latency Breakdown}
\label{subsec:latency_breakdown}

Table~\ref{tab:forward_latency} presents the top-level latency summary and
Table~\ref{tab:layer_latency} the per-operation latency for one transformer
layer.

\begin{table}[H]
  \centering
  \begin{tabular}{rrrrl}
    \toprule
    \multicolumn{2}{c}{Latency (cycles)} &
    \multicolumn{2}{c}{Latency (s)} & \\
    min & max & min & max & Note \\
    \midrule
    1{,}072{,}750{,}728 & 1{,}365{,}291{,}592 & 7.152 & 9.102 & pos\,=\,0 to pos\,=\,2047 \\
    \bottomrule
  \end{tabular}
  \caption{Top-level latency summary for the \texttt{forward()} kernel at the
           \SI{150}{\mega\hertz} target clock. The minimum corresponds to
           \texttt{pos\,=\,0} (one attended position per head) and the maximum
           to \texttt{pos\,=\,2047} (full context length).}
  \label{tab:forward_latency}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Operation & Module & Latency (cycles) \\
    \midrule
    Attention RMSNorm              & \texttt{rmsnorm\_2048\_s}      &      12{,}486 \\
    Quantise $\mathbf{xb}$         & \texttt{quantize\_2048\_s}     &       7{,}780 \\
    Q projection $\mathbf{W}_Q$    & \texttt{matmul\_2048\_2048\_s} &   4{,}635{,}652 \\
    K projection $\mathbf{W}_K$    & \texttt{matmul\_2048\_256\_s}  &     581{,}252 \\
    V projection $\mathbf{W}_V$    & \texttt{matmul\_2048\_256\_s}  &     581{,}252 \\
    Quantise KV cache entries      & \texttt{quantize\_256\_s}      &          977 \\
    Multihead attention (pos = 0)  & score + softmax + value        & $\approx$204{,}900 \\
    Multihead attention (pos = 2047) & score + softmax + value      & $\approx$13{,}502{,}200 \\
    Residual addition (attention)  & \texttt{Pipeline\_residual1}   &          137 \\
    FFN RMSNorm                    & \texttt{rmsnorm\_2048\_s}      &      12{,}486 \\
    Quantise $\mathbf{xb}$ (FFN)   & \texttt{quantize\_2048\_s}     &       7{,}780 \\
    Gate projection $\mathbf{W}_1$ & \texttt{matmul\_2048\_5632\_s} &  12{,}744{,}452 \\
    Up projection $\mathbf{W}_3$   & \texttt{matmul\_2048\_5632\_s} &  12{,}744{,}452 \\
    Quantise $\mathbf{hb}$         & \texttt{quantize\_5632\_s}     &      21{,}388 \\
    Down projection $\mathbf{W}_2$ & \texttt{matmul\_5632\_2048\_s} &  12{,}552{,}708 \\
    Residual addition (FFN)        & \texttt{Pipeline\_residual2}   &          137 \\
    \bottomrule
  \end{tabular}
  \caption{Per-operation latency for one transformer layer at
           \SI{150}{\mega\hertz}. The Q projection uses a dedicated
           \texttt{matmul\_2048\_2048\_s} instance since its output
           dimension equals \texttt{dim}. The K and V projections share
           a \texttt{matmul\_2048\_256\_s} instance, as their output
           dimension equals \texttt{kv\_dim\,=\,256} under GQA. The gate
           and up projections share one \texttt{matmul\_2048\_5632\_s}
           instance, called sequentially.}
  \label{tab:layer_latency}
\end{table}

The per-layer latency is overwhelmingly determined by the six matrix--vector
multiplications. At \texttt{pos\,=\,0}, the three FFN projections account for
approximately $12{,}744{,}452 + 12{,}744{,}452 + 12{,}552{,}708 = 38{,}041{,}612$
cycles, or 83\% of the per-layer budget. The Q projection contributes a
further 4{,}635{,}652 cycles (10\%), and the K and V projections together add
1{,}162{,}504 cycles (3\%). All remaining operations (two RMSNorm calls, three quantise calls, the
KV cache write, the multihead attention at \texttt{pos\,=\,0}, and both
residual additions) account for the remaining 4\%.
This confirms that the tiled matrix--vector multiplication, rather than the
attention mechanism, is the bottleneck for short context lengths. At longer
context lengths the attention block grows proportionally, reaching
approximately 13{,}502{,}200 cycles per layer at \texttt{pos\,=\,2047}, which
is comparable to a single FFN projection call.

\section{Functional Verification}
\label{sec:functional_verification}

\subsection{Primitive-Level C-Simulation}
\label{subsec:primitive_csim}

Each kernel primitive was verified in isolation through a dedicated standalone
HLS project. The testbench compiled the kernel function with GCC alongside a
software reference, ran both on the same test inputs, and compared outputs
element-by-element. Table~\ref{tab:primitive_csim} summarises the csim
outcome for each primitive at prototype and full TinyLlama scale.

\begin{table}[H]
  \centering
  \begin{tabular}{llll}
    \toprule
    Primitive & Scale & Test inputs & Result \\
    \midrule
    \texttt{rmsnorm}           & Proto ($S = 8$)         & Fixed, unit weights     & PASS (0 error) \\
    \texttt{rmsnorm}           & Full ($S = 2048$)        & Incremental vector      & PASS (0 RMS error) \\
    \texttt{quantize}          & Proto ($S = 8$)         & Mixed values            & PASS (0 error) \\
    \texttt{quantize}          & Full ($S = 2048$)        & Modular, 32 groups      & PASS (0 error) \\
    \texttt{dequantize}        & Proto ($S = 8$)         & Known INT8, fixed scale & PASS (0 error) \\
    \texttt{dequantize}        & Full ($S = 2048$)        & Uniform scale 0.03      & PASS (0 error) \\
    \texttt{matmul}            & Proto ($N = D = 8$)     & Quantised pattern       & PASS (tol.\ $10^{-5}$) \\
    \texttt{matmul} (tiled)    & Full ($N = 2048$, $D = 5632$) & 1{,}408 tiles   & PASS (tol.\ $10^{-3}$) \\
    \texttt{softmax}           & Proto (size 16)          & Scores, float ref       & PASS (0 error) \\
    \texttt{softmax\_ibert}    & Proto (size 16)          & LUT path                & PASS (0 error) \\
    \texttt{softmax\_ibert}    & Full (size 2048)          & Same as prototype       & PASS (0 error) \\
    \texttt{attn} (float)      & Proto                    & pos\,=\,3, pre-built cache & PASS (tol.\ $10^{-4}$) \\
    \texttt{attn} (INT8-KV)    & Proto                    & pos\,=\,3, INT8 cache   & PASS (tol.\ $10^{-4}$) \\
    \texttt{attn} (I-BERT)     & Full ($n\_heads = 32$)   & pos\,=\,3, INT8 cache   & PASS (0 error) \\
    \texttt{ffn}               & Proto ($d = 8$)         & Quantised, float ref    & PASS (tol.\ 0.05) \\
    \texttt{ffn}               & Full ($d = 2048$)        & Quantised, float ref    & PASS (tol.\ 0.05) \\
    \bottomrule
  \end{tabular}
  \caption{C-simulation results for all kernel primitives at both prototype
           and full TinyLlama scale. Tolerance values indicate the per-element
           absolute error threshold. Zero-error results indicate bit-identical
           outputs between the kernel and the software reference.}
  \label{tab:primitive_csim}
\end{table}

All sixteen standalone csim runs completed with zero errors as reported by
Vitis HLS. The zero-error results for the scalar primitives reflect the
deterministic nature of these operations, where the kernel and the sequential
software reference perform identical arithmetic in the same order. The matrix
multiplication and attention primitives use a non-zero tolerance to accommodate
the rounding difference from INT8 group quantisation. The \texttt{ffn}
tolerance of 0.05 accommodates the accumulated rounding from three sequential
quantised matrix multiplications.

\subsection{End-to-End Stress Test}
\label{subsec:stress_csim}

The complete \texttt{forward()} function was verified through a multi-token
stress test. The testbench
ran 5 autoregressive decode steps through all 22 transformer layers, using
randomly-initialised weights and a fresh random input embedding at each
position. A float-precision software reference implemented the same attention
and FFN sequence sequentially, maintaining separate per-layer KV history
vectors to match the kernel's per-layer cache layout. The relative error
between the kernel output and the float reference was computed after each
position.

Table~\ref{tab:stress_results} shows the relative error at each position.

\begin{table}[H]
  \centering
  \begin{tabular}{rrr}
    \toprule
    Token position & KV entries attended & Relative error (\%) \\
    \midrule
    0 & 1 &  9.13 \\
    1 & 2 &  7.72 \\
    2 & 3 &  8.02 \\
    3 & 4 & 16.01 \\
    4 & 5 & 18.14 \\
    \bottomrule
  \end{tabular}
  \caption{End-to-end relative error between the \texttt{forward()} kernel
           and the float software reference across five token positions.
           Relative error is computed as total absolute difference divided
           by total reference magnitude, expressed as a percentage.}
  \label{tab:stress_results}
\end{table}

The kernel completed all five positions without error or abnormal
termination, confirming that the AXI interface configuration, the four-bundle
separation of weight and KV cache traffic, and the inner-loop pipelining
changes in the attention block all operate correctly in the composed design. Positions 0--2 show relative errors in the range 7.7--9.1\%, which is
consistent with the expected accumulation of INT8 quantisation rounding across
22 transformer layers. The step increase to 16.0\% and 18.1\% at positions 3
and 4 reflects the growing influence of the I-BERT softmax approximation at
longer context lengths. With randomly-initialised weights producing hidden state
magnitudes in the thousands, the attention score differences between positions
can easily exceed the I-BERT LUT's effective dynamic range of approximately six
units. When several attended positions fall outside this range, the LUT assigns
them zero weight, concentrating the probability mass on the top one or two
positions and producing the observed step increase in error.

These error figures are measured against a float reference using
randomly-initialised weights rather than the trained TinyLlama weights. Trained
weights produce numerically stable activations through the gradient descent
constraint, and the actual approximation error under trained weights is expected
to differ. Measuring perplexity on a held-out text corpus with trained INT8
weights is identified as a necessary next step for a complete accuracy
characterisation.

\section{Inference Performance}
\label{sec:performance}

Hardware deployment on the ZCU102 was not completed within the scope of this
project, so measured inference throughput figures are not available. Theoretical
throughput is derived from the HLS synthesis latency estimates at the
\SI{150}{\mega\hertz} target clock. Table~\ref{tab:theoretical_throughput}
presents the theoretical tokens per second across the context length range.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrr}
    \toprule
    Context length & Latency (cycles) & Latency (s) & Throughput (tok/s) \\
    \midrule
    1 (pos\,=\,0)        & 1{,}072{,}750{,}728 & 7.152 & 0.140 \\
    2048 (pos\,=\,2047)  & 1{,}365{,}291{,}592 & 9.102 & 0.110 \\
    \bottomrule
  \end{tabular}
  \caption{Theoretical inference throughput derived from HLS synthesis
           latency estimates at \SI{150}{\mega\hertz}. These figures assume
           the timing violation is resolved and the design operates at the
           target frequency. No hardware deployment measurements are available.}
  \label{tab:theoretical_throughput}
\end{table}


The root cause is the sequential tiled matrix--vector multiplication
architecture. Each of the six weight matrices per layer is read tile-by-tile
from DDR, and within each tile the load and compute phases execute sequentially
rather than concurrently. At \SI{150}{\mega\hertz} with \texttt{TILE\_D\,=\,4},
the DDR burst latency for each weight tile dominates the compute time, as the
arithmetic intensity of a matrix--vector multiplication is only two operations
per byte at INT8 precision. The tiled approach was necessary to prevent the AXI
queue saturation that occurred in earlier iterations, but it sacrifices the
pipeline overlap between loading and computing that is required to approach
theoretical DDR bandwidth efficiency.

\section{Discussion}
\label{sec:discussion}

\subsection{Bottleneck Analysis}
\label{subsec:bottleneck}

The synthesis results confirm that matrix--vector multiplication is the
throughput bottleneck for TinyLlama decode-phase inference on this FPGA
architecture, accounting for over 99\% of the per-layer compute budget at
short context lengths. This is consistent with the profiling analysis of the
GPU baseline in \cref{sec:motivations}, which identified the decode phase as
memory-bandwidth-bound rather than compute-bound. On the GPU the bottleneck
manifests as low memory bandwidth utilisation due to kernel launch overhead.
On this FPGA implementation it manifests as sequential DDR bursts within the
tiling loop, where the hardware stalls during each load phase before the
compute phase can begin.

The LUT utilisation of 73\% indicates that the design is approaching the
capacity of the ZCU102's programmable logic fabric. The 64-element
fully-unrolled \texttt{dot\_qk} and \texttt{q\_quantize} loops in the
multihead attention block are the largest individual contributors. Reducing
the unroll factor from 64 to 16 in these loops would substantially lower the
LUT count, accepting a proportionally longer per-head score computation time
in exchange for headroom for Vivado routing.

\subsection{Accuracy Characterisation}
\label{subsec:accuracy}

The end-to-end relative errors of 7.7--18.1\% from the stress test reflect
two distinct sources of approximation. The first is INT8 group quantisation
with \texttt{GS\,=\,64}, which introduces a rounding error of up to
$\frac{w_{\max}}{254}$ per element per quantised operation. With six weight
matrices per layer across 22 layers these errors accumulate through the
residual stream. The second source is the I-BERT LUT exponential, which
assigns zero weight to any attention score more than approximately six units
below the maximum. At positions 3 and 4, where random weights produce large
hidden state magnitudes, this threshold is frequently exceeded and the
attention distribution collapses, producing the observed step increase in
error. The accuracy figures should be interpreted in light of the
randomly-initialised weights used in the stress test. Trained TinyLlama
weights would produce numerically controlled activations and are expected to
exhibit substantially lower relative error.

\subsection{Design Limitations}
\label{subsec:limitations}

Several deliberate simplifications in the current implementation limit its
completeness as a deployed inference engine. The output projection
$\mathbf{W}_O$ after the multihead attention block is disabled. Its computation
is present in the source but the result is discarded and the attention output
is added directly to the residual stream without the learned cross-head mixing
transformation. Rotary positional embeddings are also omitted. Without RoPE
the model has no positional awareness, and the attention scores treat all
sequence positions identically.

Double-buffering in the tiled matmul is the highest-impact optimisation
available without changing the algorithm. Overlapping the DDR load of tile
$k+1$ with the compute of tile $k$ would approximately halve the per-layer
matmul latency, reducing the theoretical decode latency from approximately
8--10 seconds per token to approximately 4--5 seconds. This would bring the
design within range of the ARM Cortex-A53 software baseline and make a direct
throughput comparison meaningful.
