% Front/abstract.tex
% -----------------------------------------------------------------------------
% The abstract is an unnumbered chapter (*) so it does not get a chapter number,
% but we still add it to the table of contents manually with \addcontentsline.
% NTU specifies a maximum of one page for the abstract.
%
% A good abstract for an engineering FYP covers four things in order:
%   1. The problem and why it matters (1--2 sentences)
%   2. What you did / your approach (2--3 sentences)
%   3. Your key results with numbers (2--3 sentences)
%   4. The significance / conclusion (1 sentence)

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Small-scale large language models (LLMs) such as TinyLlama-1.1B have 
emerged as a compelling class of compact transformer models that deliver 
meaningful natural language understanding capabilities within a fraction 
of the memory footprint of frontier models such as GPT-4. Yet even at 
the billion-parameter scale, these models impose substantial computational 
demands that limit their deployment in embedded and edge environments.

%, mostly from Nvidia and AMD, as they offer superior software support 
%for Machine Learning (ML) and Artificial Intelligence (AI) researchers and engineers with their respective 
%Compute Unified Device Architecture (CUDA) and Radeon Open Compute (ROCM) offerings. 
Currently, these language models are mostly deployed on Graphics Procesing Units (GPU).
However, GPU inference is energy-intensive and memory-bandwidth-bound, particularly 
during the autoregressive decode phase since the time and memory complexity
of self-attention are quadratic in sequence length.
% May I should introduce the process of inference and its definition and its application
% rather than talking about cuda to give more depth.

Field-programmable gate arrays (FPGAs) offer a 
compelling alternative to GPU-based inference by providing reconfigurable
hardware with low power consumption and predictable latency. But mapping a billion-parameter 
transformer to resource-constrained FPGA fabric presents significant challenges 
in quantization, memory tiling, and hardware-software partitioning.

This project starts off with a comprehensive profiling of a RTX4060 Laptop GPU 
running a pytorch inference of the Llama 2-based TinyLlama 1.1B model
from HuggingFace with default GPU kernel settings to observe for bottlenecks and 
to measure propotion of wall-clock time required by the self-attention and 
feed-forward layer during the forward pass. Based on the results, an 
FPGA-based accelerator is proposed to accelerate the architecture of the Tinyllama model. 
For this project, emphasis is placed on the hardware implementation and optimization of 
the self attention layer and the feed-forward layer.

This project implements and optimises the inference pipeline of TinyLlama-1.1B
on the Xilinx ZCU102 FPGA platform using Vitis High-Level Synthesis (HLS).
The core transformer components -- including RMSNorm, matrix--vector
multiplication, multi-head attention, and the SwiGLU feed-forward network --
were implemented as quantised INT8 HLS kernels. 
 

% --- Replace this block with your actual results once available ---------------
% Example of the kind of quantitative results you should report here:
%   "The final implementation achieves a throughput of X tokens/second at
%    Y MHz clock frequency, utilising 43.9\% of available LUTs and 34.2\%
%    of BRAM on the ZCU102. This represents a Z$\times$ improvement in
%    energy efficiency compared to a software baseline running on an
%    ARM Cortex-A53 processor."
% -----------------------------------------------------------------------------

(The results demonstrate that FPGA acceleration with INT8 quantisation
achieves competitive inference performance while operating within the
ZCU102's resource budget, validating the viability of transformer-based
language model deployment on reconfigurable hardware.)
