% Chapter1/chapter1.tex -- Introduction
% -----------------------------------------------------------------------------
% The introduction serves three purposes:
%   1. Motivate the problem (why does it matter?)
%   2. State your objectives (what specifically did YOU do?)
%   3. Orientate the reader (what does the rest of the report contain?)
%
% Keep Section 1.1 broad and accessible -- your examiner may not be an ML expert.
% Keep Section 1.2 specific and measurable -- these become your evaluation criteria.

\chapter{Introduction}
\label{ch:introduction}

\section{Motivations}
\label{sec:motivations}

Large language models (LLMs) have demonstrated remarkable capabilities
across a wide range of natural language understanding and generation tasks,
but their deployment in resource-constrained environments remains a
significant challenge \cite{PLACEHOLDER}. State-of-the-art large language models (LLMs) 
such as GPT-4 and LLaMA require tens or hundreds of gigabytes of memory and specialised
GPU clusters to achieve practical inference throughput, making them
inaccessible for embedded, edge, and real-time applications.

The llama 2-based TinyLlama-1.1B model \cite{PLACEHOLDER} represents a class of 
small-scale LLMs that seek to balance capability with deployability. With only 1.1 billion
parameters,
TinyLlama retains competitive performance on common benchmarks while fitting
within a much smaller memory footprint. Nevertheless, even at this scale,
running inference on a conventional CPU is far too slow for real-time use,
and GPUs consume too much power for battery-operated or always-on deployments.

Field-programmable gate arrays (FPGAs) offer a compelling alternative.
Unlike fixed-function ASICs, FPGAs are reconfigurable: the hardware can be
tailored precisely to the computational patterns of a specific model, enabling
high-throughput, low-latency inference at a fraction of the power cost of a GPU
\cite{PLACEHOLDER}. Two structural inefficiencies of GPU-based inference motivate
this choice. First, autoregressive decode is memory-bandwidth-bound: each decode
step must reload the full model weights from DRAM to perform matrix--vector
multiplications against a single token, yet profiling of TinyLlama-1.1B on an
RTX~4060 Laptop GPU reveals low memory bandwidth utilisation as the 
remainder is lost to kernel launch overhead and memory access latency between
hundreds of fine-grained CUDA kernel invocations per step. Second, the GPU
runtime cannot fuse operations across transformer sub-layers: RMSNorm,
matrix--vector multiplication, and activation functions are dispatched as
separate kernels, each incurring CPU-to-GPU scheduling latency and cache
eviction between them. An FPGA accelerator eliminates both inefficiencies
simultaneously: weight data flows directly through a statically scheduled
pipeline without repeated DRAM round-trips, and the entire transformer
sub-layer (normalisation, projection, and activation) can be implemented
as a single fused datapath with no inter-kernel overhead. Recent advances
in high-level synthesis (HLS) tools (Xilinx Vitis HLS)
have substantially lowered the barrier to FPGA development by allowing
engineers to describe such fused hardware pipelines in C++ and have the
tool generate the corresponding RTL automatically.

\section{Objectives and Scope}
\label{sec:objectives}
The general objective of this project is to implement an AI accelerator by
implementing and optimizing the inference pipeline of
TinyLlama-1.1B on the Xilinx ZCU102 evaluation board using Vitis HLS. Due to the 
complexity of the ZCU102 board, we begin our initial kernel design by targeting the 
KRIA KR260 FPGA board. Since the KRIA KR260 FPGA have insufficient resource to 
implement the TinyLlama model, we decided to start with a scaled-down version of 
the model's dimensions.

To understand the bottleneck of AI model inference which will be targeted for acceleration, 
literature review is performed and profiliers are used to obtain performance metrics from 
Nvidia's RTX4060 Laptop GPU. Then, a simple matrix multiplication kernel is 
first implemented with High-Level Synthesis (HLS). By the end of the project, 
a fused kernel consisting of the self-attention and feed-forward layer of the TinyLlama 
archietecture is implemented and synthesized on an FPGA. 

% The primary objectives are as follows.

% \begin{enumerate}
%   \item Implement the core transformer components of TinyLlama -- including
%         RMSNorm, matrix--vector multiplication, multi-head attention (MHA),
%         and the SwiGLU feed-forward network -- as optimised HLS kernels.

%   \item Apply INT8 post-training quantisation (PTQ) to reduce memory bandwidth
%         requirements and exploit the ZCU102's DSP blocks efficiently, while
%         maintaining acceptable output quality.

%   \item Apply HLS optimisation directives (PIPELINE, UNROLL, ARRAY\_PARTITION)
%         to minimise the initiation interval and achieve the target throughput
%         within the ZCU102's resource budget.

%   \item Evaluate the final implementation against a CPU software baseline in
%         terms of inference throughput (tokens per second), latency per token,
%         and energy efficiency (tokens per joule).
% \end{enumerate}

% The scope of this project is limited to autoregressive token generation
% (i.e., the decode phase of inference). Prefill-phase optimisation and
% training are out of scope.

\section{Organisation of Report}
\label{sec:organisation}

The remainder of this report is organised as follows.
\Cref{ch:litreview} reviews the relevant literature on FPGA-based
transformer acceleration, INT8 quantisation techniques, and related HLS
optimisation strategies. \Cref{ch:architecture} presents the overall system
architecture and the design decisions made at the hardware level.
\Cref{ch:implementation} describes the detailed implementation of each HLS
kernel and the software driver layer. \Cref{ch:results} presents the
experimental results including resource utilisation, performance benchmarks,
and comparison against baseline systems. \Cref{ch:conclusion} draws
conclusions and identifies directions for future work.
